---
title: "Dimensionality Reduction: Principal Component Analysis of the Kepler KOI Dataset"
subtitle: "PCA on astronomical data"
author: "Alejandro Treny Ortega"
date: "`r Sys.Date()`"
prefer-html: true
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    echo: true
    code-summary: "Show code"
    theme: cosmo
    number-sections: true
  pdf:
    toc: true
    toc-depth: 3
    code-fold: false
    echo: false
    number-sections: true
---

# Introduction: An Astronomical question

Each star is unique—some are enormous and flaming hot, others are small and cool. Around these stars, we've discovered hundreds of planets, each with its own orbit and size. But: **How far do correlations go in astronomical observations?**

The Kepler Space Telescope dataset contains **153 variables** describing exoplanet candidates. 

We have used **Principal Component Analysis (PCA)** as a tool to approach this data. PCA understood as an statistical technique that takes high-dimensional data and transforms it into a simpler set of "new dimensions"—dimensions that capture the most important patterns hidden in the original data. **Is this feasible in such data?**

In this analysis, we'll explore this technique through the Kepler exoplanet dataset, discovering how stellar properties and orbital mechanics interact, why some variables are redundant, and ultimately, what the **true underlying structure** of the exoplanet population looks like.

---

# Section 1: Setting the Stage—Data Preparation and Exploration

## Understanding Our Dataset

The Kepler Object of Interest (KOI) dataset is full of information about exoplanet candidates identified by NASA's Kepler Space Telescope. Rather than work with all 153 variables, we've carefully selected **14 meaningful variables** that tell the story of both the host stars and their planetary companions.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(ggrepel)
library(dplyr)
library(tidyr)
library(ggcorrplot)
library(GGally)
library(ggforce)
library(ggridges)
library(purrr)
library(broom)
library(plotly)

# Set seed for reproducibility
set.seed(42)
```

## The Variables

Let's understand what we're measuring. Our 14 selected variables fall into four natural categories:

### Identifiers and Classification
- **kepid**: Unique target identifier
- **koi_disposition**: Final classification (CONFIRMED, CANDIDATE, or FALSE POSITIVE)

### Vetting Flags (Red Flags for False Detections)
- **koi_fpflag_nt**: Not Transit-Like Flag (binary: 0 or 1)
- **koi_fpflag_ss**: Stellar-Stochastic Flag (binary: 0 or 1)

### Orbital Parameters (What We Observe)
- **koi_period**: Orbital period in days
- **koi_duration**: Transit duration in hours
- **koi_depth**: Transit depth in ppm (parts per million)

### Planetary Properties (Relative to Earth)
- **koi_prad**: Planetary radius in Earth radii
- **koi_insol**: Insolation flux in Earth flux units

### Host Star Properties (Relative to Our Sun)
- **koi_steff**: Stellar effective temperature (Kelvin)
- **koi_srad**: Stellar radius (Solar radii)
- **koi_smass**: Stellar mass (Solar masses)
- **koi_slogg**: Stellar surface gravity (log₁₀(g) in cm/s²)
- **koi_kepmag**: Kepler-band magnitude (brightness)

```{r load-data, message=FALSE, warning = FALSE}
# Load the Kepler KOI dataset
# Using a local copy to avoid network delays
# df_koi <- read.csv("df_koi.csv")
url_koi = "https://exoplanetarchive.ipac.caltech.edu/TAP/sync?query=select+*+from+q1_q17_dr25_koi&format=csv"

df_koi = read_csv(url_koi, show_col_types = FALSE)

# Select and rename columns for clarity
df_selected <- df_koi %>%
  select(
    kepid, koi_disposition, koi_fpflag_nt, koi_fpflag_ss,
    koi_period, koi_duration, koi_depth, koi_prad, koi_insol,
    koi_steff, koi_srad, koi_smass, koi_slogg, koi_kepmag
  ) %>%
  rename(
    id = kepid,
    disposition = koi_disposition,
    flag_notransit = koi_fpflag_nt,
    flag_stellareclipse = koi_fpflag_ss,
    period_days = koi_period,
    duration_hours = koi_duration,
    depth_ppm = koi_depth,
    radius_earth = koi_prad,
    insolation = koi_insol,
    teff_K = koi_steff,
    radius_sun = koi_srad,
    mass_sun = koi_smass,
    logg = koi_slogg,
    magnitude = koi_kepmag
  )

cat("Dataset dimensions:", nrow(df_selected), "observations,", ncol(df_selected), "variables\n")
```

## Creating Domain-Knowledge Variables

Before diving into PCA, we create some binary and categorical variables based on **astrophysical knowledge**. These won't be used in the PCA itself—they'll help us validate whether our new dimensions actually capture meaningful physical phenomena.These created binary/categorical variables are:
  - Hot Star (using the median of tekk_k) - > binary variable.
  - Large Planet (using radius = 4 from physics convention) - > binary variable.
  - Insolation Class (using insolation terciles) -> categorical variable (low, medium, high).
  - Stellat Brightness (using magnitude terciles) -> categoricla variable (bright, medium, dim).

```{r create-derived-vars, warning=FALSE}
df_selected <- df_selected %>%
  mutate(
    # Binary: Temperature
    # A star hotter than the median is classified as "hot"
    hot_star = if_else(teff_K > median(teff_K, na.rm = TRUE), 1, 0),
    
    # Binary: Planet Size (Physical Threshold)
    # 4 Earth-radii is a critical astrophysical boundary.
    # Planets > 4 R⊕ are almost certainly gas giants with thick atmospheres.
    # Planets < 4 R⊕ are likely rocky or water worlds.
    large_planet = if_else(radius_earth > 4, 1, 0),
    
    # Categorical: Energy Reception (Terciles)
    # We divide insolation into three equal-count groups.
    # This allows us to test if planets in different energy environments
    # have fundamentally different properties.
    insolation_class = case_when(
      insolation < quantile(insolation, 0.33, na.rm = TRUE) ~ "low",
      insolation < quantile(insolation, 0.66, na.rm = TRUE) ~ "medium",
      TRUE ~ "high"
    ),
    
    # Categorical: Stellar Brightness (Terciles)
    # Note: magnitude is inverse. Smaller values = brighter stars.
    # This captures observational bias: bright stars are easier to observe.
    magnitude_class = case_when(
      magnitude < quantile(magnitude, 0.33, na.rm = TRUE) ~ "bright",
      magnitude < quantile(magnitude, 0.66, na.rm = TRUE) ~ "medium",
      TRUE ~ "dim"
    )
  )

cat("Domain knowledge variables created successfully.\n")
```

---

# Section 2: Why Standardization Matters

## The Variance Problem

Before we perform PCA, we must make a critical decision: **Should we use the Covariance Matrix (S) or the Correlation Matrix (R)?**

To understand why this matters, let's examine the raw variance of each variable. This is where things get dramatic.

```{r select-pca-vars, warning=FALSE}
# Select only continuous variables for PCA
pca_vars <- c(
  "period_days", "duration_hours", "depth_ppm", "radius_earth",
  "insolation", "teff_K", "radius_sun", "mass_sun", "logg", "magnitude"
)

# Remove rows with missing values
df_numeric_pca <- df_selected %>%
  select(all_of(pca_vars)) %>%
  na.omit()

cat("Clean dataset for PCA:", nrow(df_numeric_pca), "observations\n")
```

```{r covariance-matrix}
# Calculate the covariance matrix of raw, unstandardized data
S_matrix_raw <- cov(df_numeric_pca)

# Calculate raw variances
raw_variances <- df_numeric_pca %>%
  summarise(across(everything(), ~ var(.x, na.rm = TRUE)))

print("Raw Variances:")
print(round(raw_variances[1:10], 3))
```

```{r plot-raw-variance, fig.cap="Massive variance differences demonstrate the need for standardization. Note the logarithmic scale!"}
# Convert to long format for visualization
variances_long <- raw_variances %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Variance"
  )

# Plot with logarithmic scale
variance_plot <- ggplot(variances_long, aes(x = reorder(Variable, -Variance), y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  scale_y_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  labs(
    title = "Variance Crisis: Raw Data Scale Disparity",
    subtitle = "Why PCA on the Covariance Matrix Fails",
    x = "Variable",
    y = "Variance (Log Scale)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )

print(variance_plot)
```

### The Problem Explained

The differences are **astronomical**. Some variables (like `depth_ppm` and `teff_K`) have variances in the millions, while others (like `logg` and `magnitude`) are tiny. 

**Here's the catastrophe:** If we use the Covariance Matrix, the PCA algorithm would be completely dominated by the largest-variance variables. It would ignore the smaller ones entirely. The first principal component would essentially be an axis aligned with `depth_ppm` and `teff_K`—not because they're the most important physically, but because they **shout the loudest** in raw units.

We need **standardization**.

---

# Section 3: Quantifying the Intercorrelation

Before we standardize, let's ask: **Is PCA even worth doing?** 

If our variables are completely independent, PCA gives us nothing. But if they're highly correlated, PCA will find underlying patterns. Let's measure this.

## Six Measures of Intercorrelation

We calculate six different metrics (q1-q6), those studied in class, that all measure correlation from different angles. Each returns a score between 0 and 1, where 1 = perfect intercorrelation and 0 = no correlation.

```{r correlation-matrix}
# Calculate correlation matrix R
R_matrix <- cor(df_numeric_pca)

# Calculate eigenvalues
eigenvalues <- sort(eigen(R_matrix)$values, decreasing = TRUE)

# Calculate inverse matrix diagonal (VIF components)
r_jj <- diag(solve(R_matrix))

# Number of variables
p <- ncol(R_matrix)

cat("Number of variables (p):", p, "\n")
cat("First three eigenvalues:", round(eigenvalues[1:3], 3), "\n")
```

```{r intercorr-measures}
# Calculate all six intercorrelation measures
q1 <- (1 - (min(eigenvalues) / max(eigenvalues)))^(p + 2)
q2 <- 1 - (p / sum(1 / eigenvalues))
q3 <- 1 - sqrt(abs(det(R_matrix)))
q4 <- (max(eigenvalues) / p)^(3/2)
q5 <- (1 - min(eigenvalues) / p)^5
q6 <- sum((1 - 1 / r_jj) / p)

# Create summary table
intercorr_measures <- data.frame(
  Measure = c("q1", "q2", "q3", "q4", "q5", "q6"),
  Value = c(q1, q2, q3, q4, q5, q6),
  Description = c(
    "Eigenvalue Spread",
    "Harmonic Mean of Eigenvalues",
    "Determinant-Based (|R| near 0)",
    "Max Eigenvalue Dominance",
    "Min Eigenvalue Magnitude",
    "Avg. Variance Inflation Factor"
  ),
  Interpretation = c(
    "Medium spread (not all-or-nothing)",
    "Medium intercorrelation",
    "HIGH - Strong multicollinearity",
    "LOW - Multiple components needed",
    "VERY HIGH - Severe redundancy",
    "Moderate VIF inflation"
  )
)

knitr::kable(intercorr_measures, caption = "Six Intercorrelation Measures: What They Tell Us")
```

### Interpretation

**PCA is strongly justified.**

- **q3 (0.763)** and **q5 (0.934)** are the smoking guns. The determinant of R is near zero, and the minimum eigenvalue is tiny. This means the variables contain **massive redundancy**.

- **q4 (0.143)** tells us something equally important: The largest eigenvalue is NOT dominant. This isn't a simple "size effect"—the correlation structure is **complex**, requiring multiple components to understand.

- The moderate values for **q1, q2, q6** confirm that we'll need **3-4 components** to capture the story, not just one.

---

# Section 4: Visual Correlation Landscape

```{r correlation-heatmap, fig.cap="Correlation structure reveals clusters of related variables", warning=FALSE}
# Create the correlation heatmap
col_neg <- "#7b3294"
col_mid <- "white"
col_pos <- "#005a9c"

corr_heatmap <- ggcorrplot(
  R_matrix,
  hc.order = FALSE,
  type = "full",
  lab = TRUE,
  lab_size = 2.5,
  method = "square",
  legend.title = "Correlation"
) +
  scale_fill_gradient2(
    low = col_neg,
    mid = col_mid,
    high = col_pos,
    midpoint = 0,
    limit = c(-1, 1),
    name = "Correlation"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8)
  ) +
  labs(title = "Correlation Matrix of Kepler Variables")

print(corr_heatmap)
```

**Key observations:**
- **Stellar variables cluster together**: `mass_sun`, `radius_sun`, and `teff_K` are highly correlated. This makes physical sense—larger stars tend to be more massive and hotter.
- **Orbital variables cluster separately**: `period_days` and `duration_hours` correlate strongly with each other (Kepler's laws) but negatively with `insolation` (more distant planets receive less energy).
- **Surface gravity is anticorrelated** with both `mass_sun` and `radius_sun`, which is physically expected.

---

# Section 5: Log Transformation

Raw scatter plots are disappointing. The data is heavily skewed—most points cluster near zero, with extreme outliers stretching the axes. We need a transformation to "zoom in" on the dense regions.

```{r log-transformation}
# Apply log10 transformation to correct for positive skewness
df_log <- df_numeric_pca %>%
  mutate(
    period_days = log10(period_days + 1),
    depth_ppm = log10(depth_ppm + 1),
    insolation = log10(insolation + 1),
    radius_earth = log10(radius_earth + 1),
    duration_hours = log10(duration_hours + 1),
    teff_K = log10(teff_K + 1),
    radius_sun = log10(radius_sun + 1),
    mass_sun = log10(mass_sun + 1)
  )

cat("Log transformation applied. Note: logg and magnitude were already in log scale.\n")
```

```{r pairs-plot, fig.width=12, fig.height=12, fig.cap="Log-transformed scatter plot matrix reveals linear relationships hidden in raw data"}
# Create pairs plot (this may take a moment)
pairs_plot_log <- ggpairs(
  df_log,
  diag = list(continuous = wrap("barDiag", fill = "steelblue", bins = 30)),
  lower = list(continuous = wrap("points", size = 0.3, alpha = 0.3)),
  upper = list(continuous = wrap("points", size = 0.3, alpha = 0.3)),
  progress = FALSE
) +
  theme_minimal() +
  theme(axis.text = element_text(size = 5))

print(pairs_plot_log)
```

**The transformation worked.** After trying different transformation options, now we can see clear relationships:
- **radius_sun vs. mass_sun**: A perfect linear relationship
- **depth_ppm vs. radius_earth**: Larger planets create deeper transit signals
- **radius_sun vs. logg**: Inverse relationship (larger stars have lower surface gravity)

This visual proof confirms that PCA will find meaningful patterns.

---

# Section 6: Standardization

Now, we apply Z-score standardization:

$$z_j = \frac{x_j - \bar{x}_j}{\sigma_j}$$

This transforms each variable to mean 0 and variance 1, ensuring all variables contribute equally to the PCA.

```{r standardization}
# Apply Z-score standardization
df_scaled <- as.data.frame(scale(df_log))

# Verify standardization worked
scaled_variances <- df_scaled %>%
  summarise(across(everything(), ~ var(.x, na.rm = TRUE)))

cat("Verification - All variances after standardization:\n")
print(round(scaled_variances, 3))
```

All variances are now 1.0. The playing field is level.

---

# Section 7: Principal Component Analysis

```{r perform-pca}
# Perform PCA on standardized data
pca_result <- prcomp(
  df_scaled,
  center = FALSE,  # Already centered
  scale. = FALSE   # Already scaled
)

# Display summary
pca_summary <- summary(pca_result)
print(pca_summary)
```

## Reading the Results

- **Standard Deviation**: The square root of each eigenvalue (√λ)
- **Proportion of Variance**: How much variance this single component explains
- **Cumulative Proportion**: The running total—this is our north star

Look at the **Cumulative Proportion** row:
- PC1 explains 36.9% of variance
- PC1 + PC2 explain 58.8% of variance
- PC1 + PC2 + PC3 explain 75.3% of variance
- PC1 + PC2 + PC3 + PC4 explain 86.7% of variance

We're reaching +80% target with just 4 components!

---

# Section 8: Interpreting the Components—Reading the PCA Equations

Each principal component is a **linear combination** of the original variables. The coefficients (loadings) tell us which variables contribute most.

```{r extract-loadings}
# Get loadings for first 4 components
loadings_matrix <- pca_result$rotation
print(round(loadings_matrix[, 1:4], 3))
```

### PC1: "Stellar Scale"

**Positive loadings:**
- `radius_sun` (0.467), `mass_sun` (0.462), `teff_K` (0.269)

**Negative loadings:**
- `logg` (-0.463), `magnitude` (-0.312)

**Interpretation:** PC1 separates large, hot, bright stars (positive) from small, cool, dim stars (negative). **It's a "stellar mass and size" axis.**

### PC2: "Orbital Dynamics"

**Positive loadings:**
- `period_days` (0.564), `duration_hours` (0.516)

**Negative loadings:**
- `insolation` (-0.459)

**Interpretation:** PC2 captures the planet's distance from its star. Distant planets have long periods and receive little energy (high PC2). Close planets orbit quickly and receive intense energy (low PC2). **It's a "planetary distance" axis.**

### PC3: "Planetary Size & Signal Depth"

**Negative loadings:**
- `depth_ppm` (-0.628), `radius_earth` (-0.560)

**Interpretation:** Large planets create deep transit signals. **It's a "planet size" axis.**

### PC4: "Stellar Evolution"

**Positive loadings:**
- `teff_K` (0.773), `logg` (0.399)

**Negative loadings:**
- `radius_sun` (-0.392)

**Interpretation:** This contrasts small, hot, dense stars against large, cooler stars. It captures **stellar evolutionary stage**—main sequence vs. evolved stars.

```{r plot-loadings, fig.cap="Which variables contribute most to each component?"}
# Convert loadings to long format
loadings_df <- as.data.frame(loadings_matrix[, 1:4]) %>%
  rownames_to_column(var = "Variable") %>%
  pivot_longer(cols = -Variable, names_to = "Component", values_to = "Loading")

loadings_plot <- ggplot(loadings_df, aes(x = reorder(Variable, Loading), y = Loading, fill = Component)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~ Component, scales = "free_x") +
  labs(
    title = "PCA Loadings: The Component Equations",
    x = "Variable",
    y = "Loading Coefficient"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(loadings_plot)
```

---

# Section 9: How Many Components Do We Need?

We apply **four different criteria** to find the optimal number of components:

```{r dimensionality-reduction-criteria}
# Calculate eigenvalues and related metrics
eigenvalues <- pca_result$sdev^2
p <- length(eigenvalues)

pca_summary_data <- summary(pca_result)$importance

# Method 1: Percentage of Variance
cumulative_variance <- pca_summary_data["Cumulative Proportion", ]
variance <- pca_summary_data["Proportion of Variance", ]
cutoff_80 <- min(which(cumulative_variance >= 0.80))

# Method 2: Kaiser's Criterion (eigenvalue > 1)
kaiser_cutoff <- sum(eigenvalues > 1.0)

# Method 3: Jolliffe's Criterion (eigenvalue > 0.7)
jolliffe_cutoff <- sum(eigenvalues > 0.7)

cat("Method 1 (80% Variance):", cutoff_80, "components\n")
cat("Method 2 (Kaiser, λ > 1):", kaiser_cutoff, "components\n")
cat("Method 3 (Jolliffe, λ > 0.7):", jolliffe_cutoff, "components\n")
```

```{r scree-plot, fig.cap="Scree plot shows the 'elbow' where eigenvalues stabilize"}
# Create scree plot
variance_df <- data.frame(
  Component = 1:p,
  Eigenvalue = eigenvalues,
  Cumulative = cumulative_variance,
  Variance = variance
)

scree_plot <- ggplot(variance_df, aes(x = Component, y = Eigenvalue)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "red", label = "Kaiser's Cutoff") +
  geom_hline(yintercept = 0.7, linetype = "dotted", color = "green", label = "Jolliffe's Cutoff") +
  labs(
    title = "Scree Plot: Finding the Elbow",
    x = "Principal Component",
    y = "Eigenvalue (λ)"
  ) +
  scale_x_continuous(breaks = 1:p) +
  theme_minimal()

print(scree_plot)

# Cumulative variance plot
cumulative_plot <- ggplot(variance_df, aes(x = Component, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +  # Barras azules
  geom_line(aes(y = Cumulative), color = "red", size = 1) +      # Línea roja acumulada
  geom_point(aes(y = Cumulative), color = "red", size = 2) +     # Puntos rojos
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "black") +
  annotate("text", x = 2, y = 0.80, label = "80% Target", vjust = -0.5) +
  labs(
    title = "Cumulative Variance Explained",
    subtitle = "Bars: Individual Variance | Red Line: Cumulative",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  scale_x_continuous(breaks = 1:p) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal()

print(cumulative_plot)
```

**Consensus:** All methods point to **4 components** as optimal. This captures +85% of the variance while remaining interpretable.

---

# Section 10: Visualizing the PCA Space—Do Groups Separate?

Now for the moment of truth: Does PCA actually separate meaningful groups in our data?

```{r prepare-plot-data}
# Extract PCA scores
pca_scores_df <- as.data.frame(pca_result$x)

# Get labels (must match the exact same rows)
df_labels_clean <- df_selected %>%
  select(
    disposition, large_planet, hot_star, insolation_class, magnitude_class,
    all_of(pca_vars)
  ) %>%
  na.omit()

# Combine
plot_df <- cbind(pca_scores_df, df_labels_clean)
```

```{r pc1-v-pc2-pc3-scatter, fig.cap="PC1 vs PC2 vs PC3 colored by KOI disposition", message=FALSE, warning=FALSE}
library(plotly)
library(htmlwidgets)
library(webshot2)

# 1. Create the Plot
pca_3d_cloud <- plot_ly(
  data = plot_df,
  x = ~PC1, y = ~PC2, z = ~PC3,
  color = ~disposition, colors = "Dark2",
  type = "scatter3d", mode = "markers",
  marker = list(size = 1.5, opacity = 0.3, line = list(width = 0))
) %>%
  layout(
    title = "PCA 3D Group Density",
    scene = list(
      xaxis = list(title = "PC1 - Stellar Scale"),
      yaxis = list(title = "PC2 - Orbital Distance"),
      zaxis = list(title = "PC3 - Planet Size")
    ),
    legend = list(title = list(text = "KOI Disposition"))
  )

# 2. Save paths (Save locally to project folder, NOT tempdir)
# This prevents permission issues and makes debugging easier
html_file <- "temp_pca_3d.html"
png_file  <- "pca_3d_static.png"

saveWidget(pca_3d_cloud, html_file, selfcontained = TRUE)

# 3. Screenshot with longer delay
# We increase delay to 5 seconds to allow WebGL to initialize
webshot2::webshot(
  html_file,
  file = png_file,
  vwidth = 1000,
  vheight = 800,
  delay = 5  
)
```

The plot shows overlap, but let's apply a **statistical test** to ask: Are the groups really different?

---

# Section 11: Statistical Proof—The Kruskal-Wallis Test

```{r kruskal-wallis-tests}
# Test if groups differ significantly in PC space
pc1_test <- kruskal.test(PC1 ~ disposition, data = plot_df)
pc2_test <- kruskal.test(PC2 ~ disposition, data = plot_df)
pc3_test <- kruskal.test(PC3 ~ disposition, data = plot_df)

cat("Kruskal-Wallis Test - PC1 by Disposition:\n")
print(pc1_test)
cat("\nKruskal-Wallis Test - PC2 by Disposition:\n")
print(pc2_test)
cat("\nKruskal-Wallis Test - PC3 by Disposition:\n")
print(pc3_test)
```

**The Result:** P-values < 2.2e-16 and 4.933e-07(essentially zero). **This statistically PROVES that the groups have different medians in PC space.** The overlap we see is "noise," but the signal is real and highly significant.

---

# Section 12: The Loadings Plot—Visualizing Component Meanings

```{r loadings-biplot, fig.cap="Each arrow shows how original variables contribute to the new PCA axes"}
loadings_df <- as.data.frame(pca_result$rotation) %>%
  rownames_to_column(var = "Variable")

loadings_arrow_plot <- ggplot(loadings_df, aes(x = PC1, y = PC2, label = Variable)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.0), color = "gray", alpha = 0.3) +
  geom_segment(
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "blue"
  ) +
  geom_text_repel(
    segment.color = "gray",
    segment.size = 0.5
  ) +
  coord_fixed(ratio = 1) +
  labs(
    title = "PCA Loadings Plot",
    subtitle = "How original variables project onto PC1 and PC2",
    x = "PC1 - Stellar Scale",
    y = "PC2 - Orbital Dynamics"
  ) +
  theme_minimal()

print(loadings_arrow_plot)
```

**Key Insights:**
- **Top-right**: Stellar properties (`mass_sun`, `radius_sun`) point right (large stars)
- **Top**: Orbital properties (`period_days`, `duration_hours`) point up (wide orbits)
- **Bottom-left**: `insolation` points down-left (close, hot planets around small stars)
- **The 90-degree angle** between stellar and orbital vectors proves they're **independent**—the PCA has successfully decoupled them

---

# Section 13: Validating Components Against Domain Knowledge

Do our PCA axes capture real astrophysical phenomena? Let's test using ridge plots.

```{r ridge-plot-hot-star, fig.cap="PC1 strongly separates hot from cool stars; PC2 shows minimal separation"}
# Prepare data for ridge plots
plot_df$hot_star_label <- factor(plot_df$hot_star,
                                  levels = c(0, 1),
                                  labels = c("Cool Star", "Hot Star"))

plot_df_long <- plot_df %>%
  select(PC1, PC2, PC3, PC4, hot_star_label) %>%
  pivot_longer(
    cols = c(PC1, PC2, PC3, PC4),
    names_to = "Component",
    values_to = "Score"
  )

ridge_plot_hot <- ggplot(plot_df_long, aes(x = Score, y = hot_star_label, fill = hot_star_label)) +
  geom_density_ridges(alpha = 0.7) +
  facet_wrap(~ Component, scales = "free_x") +
  scale_fill_brewer(palette = "Set1") +
  coord_cartesian(xlim = c(-10, 10)) +
  labs(
    title = "Ridge Plot: Hot vs. Cool Stars Across PCA Components",
    x = "PC Score",
    y = "Stellar Temperature"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(ridge_plot_hot)
```

**Observation:** PC1 clearly separates hot from cool stars (peaks are offset). PC2-PC4 show much less separation. This confirms our interpretation: **PC1 is the stellar temperature/size axis.**

```{r ridge-plot-insolation, fig.cap="PCA components capture planetary energy environment across multiple dimensions"}
plot_df_long_insol <- plot_df %>%
  select(PC1, PC2, PC3, PC4, insolation_class) %>%
  mutate(insolation_class = factor(insolation_class, levels = c("low", "medium", "high"))) %>%
  pivot_longer(
    cols = c(PC1, PC2, PC3, PC4),
    names_to = "Component",
    values_to = "Score"
  )

ridge_plot_insol <- ggplot(plot_df_long_insol, aes(x = Score, y = insolation_class, fill = insolation_class)) +
  geom_density_ridges(alpha = 0.7) +
  facet_wrap(~ Component, scales = "free_x") +
  scale_fill_viridis_d(option = "C") +
  coord_cartesian(xlim = c(-10, 10)) +
  labs(
    title = "Ridge Plot: Insolation Class Across PCA Components",
    x = "PC Score",
    y = "Planetary Energy Level"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(ridge_plot_insol)
```

**Striking Result:** All four components show clear separation by insolation. **Planetary energy is not a simple one-dimensional problem—it's encoded across the full PC space.** This makes sense: to have high insolation, you need either a hot star (PC1) or a close orbit (PC2), or both.

---

# Section 14: Comprehensive Statistical Validation

Let's run the Kruskal-Wallis test for all our domain-knowledge variables against all 4 components.

```{r comprehensive-tests}
# Create all combinations
variables <- c("large_planet", "hot_star", "insolation_class", "magnitude_class")
components <- c("PC1", "PC2", "PC3", "PC4")

all_combinations <- crossing(
  Variable_Group = variables,
  Component = components
)

# Run tests
all_tests_summary <- all_combinations %>%
  rowwise() %>%
  mutate(
    formula_str = paste(Component, "~", Variable_Group),
    test_result = list(tidy(kruskal.test(as.formula(formula_str), data = plot_df)))
  ) %>%
  ungroup() %>%
  unnest(test_result) %>%
  select(Variable_Group, Component, p.value)

# Create summary table
final_p_value_table <- all_tests_summary %>%
  mutate(p.value = format.pval(p.value, digits = 2, eps = 1e-100)) %>%
  pivot_wider(
    names_from = Component,
    values_from = p.value
  ) %>%
  select(Variable_Group, PC1, PC2, PC3, PC4)

knitr::kable(final_p_value_table, caption = "P-values from Kruskal-Wallis Tests (Lower = More Significant)")
```

### Interpreting the Results

**Key Findings:**

1. **`hot_star`**: Significant on PC1 (p < 1e-100), PC2 (p ≈ 3.3e-05), PC3 (p ≈ 2.0e-23), and PC4 (p < 1e-100). Stellar temperature correlates across all principal components, affecting both stellar properties and orbital characteristics.

2. **`insolation_class`**: Universally significant (p < 1e-100). Energy reception is the most fundamental axis of variation in the exoplanet population.

3. **`large_planet`**: Significant on PC1, PC2, and PC3 (p < 1e-100), and PC4 (p ≈ 4.0e-14). Planet size correlates strongly with multiple aspects of the system.

4. **`magnitude_class`**: Significant on PC1 (p < 1e-100), PC3 (p < 1e-100), and PC4 (p ≈ 4.0e-30), but NOT on PC2 (p = 0.11). This is elegant confirmation that:
 - PC1 captures stellar brightness and size effects
 - PC2 is independent of apparent magnitude—it represents
   pure orbital dynamics free from observational biases
 - PC3 relates to transit depth and detection efficiency
 - PC4 separates stellar evolutionary stages

---

# Conclusion

Through this analysis, we've transformed a high-dimensional, complex dataset into a **meaningful lower-dimensional space** where astrophysical concepts are clearly expressed:

1. **PC1 = Stellar Scale**: A spectrum from massive, hot stars to small, cool dwarfs
2. **PC2 = Orbital Distance**: A spectrum from close-orbiting planets to distant, slow-moving worlds
3. **PC3 = Planet Size**: A measure of transit depth and planetary radius
4. **PC4 = Stellar Evolution**: A secondary axis capturing evolutionary stage

Before PCA, we had 10 correlated variables describing exoplanet systems. Now we have 4 **uncorrelated** components with clear physical meanings. We've:

- **Reduced dimensionality** from 10 to 4 (60% reduction)
- **Explained +85% of variance** with just 4 components
- **Decoupled independent phenomena** (stellar properties from orbital mechanics)
- **Quantitatively validated** our components against domain knowledge
- **Proven group separability**

---

# Technical Appendix

## Session Information

```{r session-info}
sessionInfo()
```

## Complete Code Summary

This QMD file contains fully executable R code. All code chunks are toggled with `code-fold: true`, meaning you can view or hide the code as desired.

Key libraries used:
- `tidyverse`: Data manipulation and visualization
- `ggplot2`: Advanced plotting
- `ggcorrplot`: Correlation heatmaps
- `GGally`: Pairs plots
- `plotly`: Interactive 3D visualization
- `ggridges`: Ridge plots for distribution visualization
- `broom`: Tidy statistical test results

---

**Created:** `r Sys.Date()`

**Analysis Purpose:** Dimensionality reduction and interpretation of exoplanet discovery data using Principal Component Analysis with full methodological validation.

