---
title: "Dimensionality Reduction: Principal Component Analysis of the Kepler KOI Dataset"
subtitle: "PCA on astronomical data"
author: "Alejandro Treny Ortega"
date: "`r Sys.Date()`"
prefer-html: true
format:
  html:
    format-links: false
    toc: true
    toc-depth: 3
    code-fold: true
    echo: true
    code-summary: "Show code"
    theme: cosmo
    number-sections: true
  pdf:
    pdf-engine: xelatex
    toc: true
    toc-depth: 3
    code-fold: false
    echo: false
    number-sections: true
---

# Introduction: An Astronomical question

Each star is unique—some are enormous and flaming hot, others are small and cool. Around these stars, we've discovered hundreds of planets, each with its own orbit and size. But: **How far do correlations go in astronomical observations?**

The Kepler Space Telescope dataset contains **153 variables** describing exoplanet candidates.

We have used **Principal Component Analysis (PCA)** as a tool to approach this data. PCA is a statistical technique that takes high-dimensional data and transforms it into a simpler set of "new dimensions"—dimensions that capture the most important patterns hidden in the original data. **Is this feasible in such data?**

In this analysis, we'll explore this technique through the Kepler exoplanet dataset, discovering how stellar properties and orbital mechanics interact, why some variables are redundant, and ultimately, what the **linear structure in this selected sample** of the exoplanet population looks like.

---

# Section 1: Setting the Stage—Data Preparation and Exploration

## Understanding Our Dataset

The Kepler Object of Interest (KOI) dataset is full of information about exoplanet candidates identified by NASA's Kepler Space Telescope. Rather than work with all 153 variables, we've carefully selected **14 meaningful variables** that tell the story of both the host stars and their planetary companions.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(ggrepel)
library(dplyr)
library(tidyr)
library(ggcorrplot)
library(GGally)
library(ggforce)
library(ggridges)
library(purrr)
library(broom)
library(plotly)

# Set seed for reproducibility
set.seed(42)
```

## The Variables

Let's understand what we're measuring. Our 14 selected variables fall into five categories:

### Identifiers and Classification
- **kepid**: Unique target identifier
- **koi_disposition**: Final classification (CONFIRMED, CANDIDATE, or FALSE POSITIVE)

### Vetting Flags (Red Flags for False Detections)
- **koi_fpflag_nt**: Not Transit-Like Flag (binary: 0 or 1)
- **koi_fpflag_ss**: Stellar Eclipse Flag (binary: 0 or 1)

### Orbital Parameters (What We Observe)
- **koi_period**: Orbital period in days
- **koi_duration**: Transit duration in hours
- **koi_depth**: Transit depth in ppm (parts per million)

### Planetary Properties (Relative to Earth)
- **koi_prad**: Planetary radius in Earth radii
- **koi_insol**: Insolation flux in Earth flux units

### Host Star Properties (Relative to Our Sun)
- **koi_steff**: Stellar effective temperature (Kelvin)
- **koi_srad**: Stellar radius (Solar radii)
- **koi_smass**: Stellar mass (Solar masses)
- **koi_slogg**: Stellar surface gravity (log₁₀(g) in cm/s²)
- **koi_kepmag**: Kepler-band magnitude (brightness)

```{r load-data, message=FALSE, warning = FALSE}
# Load the versioned Kepler snapshot. Refreshing the NASA source is an explicit
# maintenance step so rendered results stay reproducible.
df_koi = read_csv("df_koi.csv", show_col_types = FALSE)

# Select and rename columns for clarity
df_selected <- df_koi %>%
  select(
    kepid, koi_disposition, koi_fpflag_nt, koi_fpflag_ss,
    koi_period, koi_duration, koi_depth, koi_prad, koi_insol,
    koi_steff, koi_srad, koi_smass, koi_slogg, koi_kepmag
  ) %>%
  rename(
    id = kepid,
    disposition = koi_disposition,
    flag_notransit = koi_fpflag_nt,
    flag_stellareclipse = koi_fpflag_ss,
    period_days = koi_period,
    duration_hours = koi_duration,
    depth_ppm = koi_depth,
    radius_earth = koi_prad,
    insolation = koi_insol,
    teff_K = koi_steff,
    radius_sun = koi_srad,
    mass_sun = koi_smass,
    logg = koi_slogg,
    magnitude = koi_kepmag
  )

cat("Dataset dimensions:", nrow(df_selected), "observations,", ncol(df_selected), "variables\n")
```

## Creating Descriptive Groups

Before diving into PCA, we create some binary and categorical groups based on input values. These won't be used in the PCA itself—they describe associations with the inputs; they are not independent validation labels. The derived groups are:

- Hot Star (using the median of teff_K): binary variable.
- Large Planet (using a descriptive radius threshold of 4 Earth radii): binary variable.
- Insolation Class (using insolation terciles): categorical variable (low, medium, high).
- Stellar Brightness (using magnitude terciles): categorical variable (bright, medium, dim).

```{r create-derived-vars, warning=FALSE}
df_selected <- df_selected %>%
  mutate(
    # Binary: Temperature
    # A star hotter than the median is classified as "hot"
    hot_star = if_else(teff_K > median(teff_K, na.rm = TRUE), 1, 0),

    # Binary: Planet Size (Physical Threshold)
    # A descriptive threshold of 4 Earth radii; radius alone does not identify composition.
    large_planet = if_else(radius_earth > 4, 1, 0),

    # Categorical: Energy Reception (Terciles)
    # We divide insolation into three equal-count groups.
    # This allows us to test if planets in different energy environments
    # have fundamentally different properties.
    insolation_class = case_when(
      is.na(insolation) ~ NA_character_,
      insolation < quantile(insolation, 0.33, na.rm = TRUE) ~ "low",
      insolation < quantile(insolation, 0.66, na.rm = TRUE) ~ "medium",
      TRUE ~ "high"
    ),

    # Categorical: Stellar Brightness (Terciles)
    # Note: magnitude is inverse. Smaller values = brighter stars.
    # These bins describe brightness; they do not correct observational bias.
    magnitude_class = case_when(
      is.na(magnitude) ~ NA_character_,
      magnitude < quantile(magnitude, 0.33, na.rm = TRUE) ~ "bright",
      magnitude < quantile(magnitude, 0.66, na.rm = TRUE) ~ "medium",
      TRUE ~ "dim"
    )
  )

cat("Domain knowledge variables created successfully.\n")
```

---

# Section 2: Why Standardization Matters

## The Variance Problem

Before we perform PCA, we must make a critical decision: **Should we use the Covariance Matrix (S) or the Correlation Matrix (R)?**

To understand why this matters, let's examine the raw variance of each variable. This is where things get dramatic.

```{r select-pca-vars, warning=FALSE}
# Select only continuous variables for PCA
pca_vars <- c(
  "period_days", "duration_hours", "depth_ppm", "radius_earth",
  "insolation", "teff_K", "radius_sun", "mass_sun", "logg", "magnitude"
)

# Remove rows with missing values
pca_rows <- complete.cases(df_selected[, pca_vars])
df_numeric_pca <- df_selected[pca_rows, pca_vars]
stopifnot(all(vapply(df_numeric_pca, function(x) all(is.finite(x)) && sd(x) > 0, logical(1))))

cat("Clean dataset for PCA:", nrow(df_numeric_pca), "observations\n")
```

```{r covariance-matrix}
# Calculate the covariance matrix of raw, unstandardized data
S_matrix_raw <- cov(df_numeric_pca)

# Calculate raw variances
raw_variances <- df_numeric_pca %>%
  summarise(across(everything(), ~ var(.x, na.rm = TRUE)))

print("Raw Variances:")
print(round(raw_variances[1:10], 3))
```

```{r plot-raw-variance, fig.cap="Raw-unit variances on a logarithmic scale; standardization is the weighting choice used in this analysis."}
# Convert to long format for visualization
variances_long <- raw_variances %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Variance"
  )

# Plot with logarithmic scale
variance_plot <- ggplot(variances_long, aes(x = reorder(Variable, -Variance), y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  scale_y_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  labs(
    title = "Variance Crisis: Raw Data Scale Disparity",
    subtitle = "Raw units determine covariance-PCA weighting",
    x = "Variable",
    y = "Variance (Log Scale)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )

print(variance_plot)
```

### The Problem Explained

The differences are **astronomical**. Some variables (like `depth_ppm` and `teff_K`) have variances in the millions, while others (like `logg` and `magnitude`) are tiny.

Covariance PCA weights variables in their raw units. Here we choose correlation PCA so each input has unit variance. This is an analysis choice, not a universal requirement.

---

# Section 3: Quantifying the Intercorrelation

Before we standardize, let's ask: **Is PCA even worth doing?**

For standardized uncorrelated inputs, variance is not concentrated in a smaller number of PCs. Correlation diagnostics can indicate scope for linear compression; they do not test statistical independence.

## Six Measures of Intercorrelation

We calculate six classroom diagnostics of the raw correlation matrix. They have different baselines: at an identity matrix, q4 equals $p^{-3/2}$ and q5 equals $(1-1/p)^5$, so a high q5 alone does not prove redundancy.

```{r correlation-matrix}
# Calculate correlation matrix R
R_matrix <- cor(df_numeric_pca)

# Calculate eigenvalues
eigenvalues <- sort(eigen(R_matrix)$values, decreasing = TRUE)

# Calculate inverse matrix diagonal (VIF components)
r_jj <- diag(solve(R_matrix))

# Number of variables
p <- ncol(R_matrix)

cat("Number of variables (p):", p, "\n")
cat("First three eigenvalues:", round(eigenvalues[1:3], 3), "\n")
```

```{r intercorr-measures}
# Calculate all six intercorrelation measures
q1 <- (1 - (min(eigenvalues) / max(eigenvalues)))^(p + 2)
q2 <- 1 - (p / sum(1 / eigenvalues))
q3 <- 1 - sqrt(abs(det(R_matrix)))
q4 <- (max(eigenvalues) / p)^(3/2)
q5 <- (1 - min(eigenvalues) / p)^5
q6 <- sum((1 - 1 / r_jj) / p)

# Create summary table
intercorr_measures <- data.frame(
  Measure = c("q1", "q2", "q3", "q4", "q5", "q6"),
  Value = c(q1, q2, q3, q4, q5, q6),
  Description = c(
    "Eigenvalue Spread",
    "Harmonic Mean of Eigenvalues",
    "Determinant-Based (|R| near 0)",
    "Max Eigenvalue Dominance",
    "Min Eigenvalue Magnitude",
    "Avg. Variance Inflation Factor"
  )
)

knitr::kable(intercorr_measures, caption = "Six Intercorrelation Measures: What They Tell Us")
```

### Interpretation

The table describes raw-variable correlations. Retained variance and loadings from the transformed analysis below determine how useful the chosen reduction is.

---

# Section 4: Visual Correlation Landscape

```{r correlation-heatmap, fig.cap="Correlation structure reveals clusters of related variables", warning=FALSE}
# Create the correlation heatmap
col_neg <- "#7b3294"
col_mid <- "white"
col_pos <- "#005a9c"

corr_heatmap <- ggcorrplot(
  R_matrix,
  hc.order = FALSE,
  type = "full",
  lab = TRUE,
  lab_size = 2.5,
  method = "square",
  legend.title = "Correlation",
  colors = c(col_neg, col_mid, col_pos),
  legend.limit = c(-1, 1)
) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8)
  ) +
  labs(title = "Correlation Matrix of Kepler Variables")

print(corr_heatmap)
```

Read the signs and magnitudes from this snapshot's heatmap. These correlations alone do not establish physical laws or causal relationships.

---

# Section 5: Log Transformation

Raw scatter plots are disappointing. The data is heavily skewed—most points cluster near zero, with extreme outliers stretching the axes. We need a transformation to "zoom in" on the dense regions.

```{r log-transformation}
# Apply log10 transformation to correct for positive skewness
df_log <- df_numeric_pca %>%
  mutate(
    period_days = log10(period_days + 1),
    depth_ppm = log10(depth_ppm + 1),
    insolation = log10(insolation + 1),
    radius_earth = log10(radius_earth + 1),
    duration_hours = log10(duration_hours + 1),
    teff_K = log10(teff_K + 1),
    radius_sun = log10(radius_sun + 1),
    mass_sun = log10(mass_sun + 1)
  )

cat("Log transformation applied. Note: logg and magnitude were already in log scale.\n")
```

```{r pairs-plot, fig.width=12, fig.height=12, fig.cap="Log-transformed scatter plot matrix reveals linear relationships hidden in raw data"}
# Create pairs plot (this may take a moment)
pairs_plot_log <- ggpairs(
  df_log,
  diag = list(continuous = wrap("barDiag", fill = "steelblue", bins = 30)),
  lower = list(continuous = wrap("points", size = 0.3, alpha = 0.3)),
  upper = list(continuous = wrap("points", size = 0.3, alpha = 0.3)),
  progress = FALSE
) +
  theme_minimal() +
  theme(axis.text = element_text(size = 5))

print(pairs_plot_log)
```

Log10(x+1) reduces visual skew. Stellar mass and radius are not perfectly linearly related; transit depth also depends on host-star radius and other factors. The following PCA uses these transformed variables, with logg and magnitude left unchanged.

---

# Section 6: Standardization

Now, we apply Z-score standardization:

$$z_j = \frac{x_j - \bar{x}_j}{\sigma_j}$$

This transforms each variable to mean 0 and variance 1, giving each input unit marginal variance before PCA.

```{r standardization}
# Apply Z-score standardization
df_scaled <- as.data.frame(scale(df_log))

# Verify standardization worked
scaled_variances <- df_scaled %>%
  summarise(across(everything(), ~ var(.x, na.rm = TRUE)))

cat("Verification - All variances after standardization:\n")
print(round(scaled_variances, 3))
```

All variances are now 1.0. The playing field is level.

---

# Section 7: Principal Component Analysis

```{r perform-pca}
# Perform PCA on standardized data
pca_result <- prcomp(
  df_scaled,
  center = FALSE,  # Already centered
  scale. = FALSE   # Already scaled
)

# Display summary
pca_summary <- summary(pca_result)
print(pca_summary)
```

## Reading the Results

- **Standard Deviation**: The square root of each eigenvalue (√λ)
- **Proportion of Variance**: How much variance this single component explains
- **Cumulative Proportion**: The running total—this is our north star

The first four components retain **`r round(sum(pca_result$sdev[1:4]^2)/sum(pca_result$sdev^2)*100, 1)`%** of total standardized variance. The printed table gives each contribution.

---

# Section 8: Interpreting the Components—Reading the PCA Equations

Each principal component is a **linear combination** of the transformed, standardized variables. The coefficients (loadings) tell us which variables contribute most.

```{r extract-loadings}
# Get loadings for first 4 components
loadings_matrix <- pca_result$rotation
print(round(loadings_matrix[, 1:4], 3))
```

Interpret component weights from the table above. Signs can reverse without changing the PCA geometry. Large absolute coefficients identify the variables contributing to a direction; names such as stellar scale or orbital properties are descriptive hypotheses.

```{r plot-loadings, fig.height=6, fig.cap="Which variables contribute most to each component?"}
# Convert loadings to long format
loadings_df <- as.data.frame(loadings_matrix[, 1:4]) %>%
  rownames_to_column(var = "Variable") %>%
  pivot_longer(cols = -Variable, names_to = "Component", values_to = "Loading")

loadings_plot <- ggplot(loadings_df, aes(x = reorder(Variable, Loading), y = Loading, fill = Component)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~ Component, scales = "free_x") +
  labs(
    title = "PCA Loadings: The Component Equations",
    x = "Variable",
    y = "Loading Coefficient"
  ) +
  theme_minimal() +
  theme(legend.position = "none", panel.spacing = grid::unit(1, "lines"),
        panel.spacing.x = grid::unit(2.5, "lines"))

print(loadings_plot)
```

---

# Section 9: How Many Components Do We Need?

We apply **four different criteria** to compare choices of retained dimension:

```{r dimensionality-reduction-criteria}
# Calculate eigenvalues and related metrics
eigenvalues <- pca_result$sdev^2
p <- length(eigenvalues)

pca_summary_data <- summary(pca_result)$importance

# Method 1: Percentage of Variance
cumulative_variance <- pca_summary_data["Cumulative Proportion", ]
variance <- pca_summary_data["Proportion of Variance", ]
cutoff_80 <- min(which(cumulative_variance >= 0.80))

# Method 2: Kaiser's Criterion (eigenvalue > 1)
kaiser_cutoff <- sum(eigenvalues > 1.0)

# Method 3: Jolliffe's Criterion (eigenvalue > 0.7)
jolliffe_cutoff <- sum(eigenvalues > 0.7)

cat("Method 1 (80% Variance):", cutoff_80, "components\n")
cat("Method 2 (Kaiser, λ > 1):", kaiser_cutoff, "components\n")
cat("Method 3 (Jolliffe, λ > 0.7):", jolliffe_cutoff, "components\n")
```

```{r scree-plot, fig.cap=c("Eigenvalues with Kaiser (1) and Jolliffe (0.7) reference lines", "Individual and cumulative variance, with the 80% reference threshold")}
# Create scree plot
variance_df <- data.frame(
  Component = 1:p,
  Eigenvalue = eigenvalues,
  Cumulative = cumulative_variance,
  Variance = variance
)

scree_plot <- ggplot(variance_df, aes(x = Component, y = Eigenvalue)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "blue", size = 3) +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0.7, linetype = "dotted", color = "green") +
  labs(
    title = "Scree Plot: Finding the Elbow",
    x = "Principal Component",
    y = "Eigenvalue (lambda)"
  ) +
  scale_x_continuous(breaks = 1:p) +
  theme_minimal()

print(scree_plot)

# Cumulative variance plot
cumulative_plot <- ggplot(variance_df, aes(x = Component, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +  # Barras azules
  geom_line(aes(y = Cumulative), color = "red", linewidth = 1) +      # Línea roja acumulada
  geom_point(aes(y = Cumulative), color = "red", size = 2) +     # Puntos rojos
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "black") +
  annotate("text", x = 2, y = 0.80, label = "80% Target", vjust = -0.5) +
  labs(
    title = "Cumulative Variance Explained",
    subtitle = "Bars: Individual Variance | Red Line: Cumulative",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  scale_x_continuous(breaks = 1:p) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal()

print(cumulative_plot)
```

The 80% threshold retains `r cutoff_80` components, Kaiser retains `r kaiser_cutoff`, and Jolliffe retains `r jolliffe_cutoff`. These are descriptive rules, not proofs of an optimal dimension.

---

# Section 10: Visualizing the PCA Space—Do Groups Separate?

Now for the moment of truth: Does PCA actually separate meaningful groups in our data?

```{r prepare-plot-data}
# Extract PCA scores
pca_scores_df <- as.data.frame(pca_result$x)

# Get labels (must match the exact same rows)
df_labels_clean <- df_selected[pca_rows, ]
stopifnot(nrow(pca_scores_df) == nrow(df_labels_clean))

plot_df <- cbind(pca_scores_df, df_labels_clean)
```

```{r pc1-v-pc2-pc3-scatter, fig.cap="PCA scores coloured by KOI disposition", message=FALSE, warning=FALSE}
library(plotly)
library(htmlwidgets)
library(webshot2)

# 1. Create the Plot
pca_3d_cloud <- plot_ly(
  data = plot_df,
  x = ~PC1, y = ~PC2, z = ~PC3,
  color = ~disposition, colors = "Dark2",
  type = "scatter3d", mode = "markers",
  marker = list(size = 1.5, opacity = 0.3, line = list(width = 0))
) %>%
  layout(
    title = "PCA Scores by KOI Disposition",
    scene = list(
      camera = list(eye = list(x = 1.65, y = 1.65, z = 1.65)),
      xaxis = list(title = "PC1"),
      yaxis = list(title = "PC2"),
      zaxis = list(title = "PC3")
    ),
    legend = list(title = list(text = "KOI Disposition"))
  )

# 2. Save paths (Save locally to project folder, NOT tempdir)
# This prevents permission issues and makes debugging easier
if (knitr::is_html_output()) {
  pca_3d_cloud
} else {
  print(ggplot(plot_df, aes(PC1, PC2, colour = disposition)) +
          geom_point(alpha = 0.3, size = 0.4) + theme_minimal() +
          labs(title = "PCA scores: PC1 and PC2"))
}

```

The plot shows overlap, but let's apply a **statistical test** to ask: Are the groups really different?

---

# Section 11: Exploratory Comparisons—The Kruskal-Wallis Test

```{r kruskal-wallis-tests}
# Test if groups differ significantly in PC space
pc1_test <- kruskal.test(PC1 ~ disposition, data = plot_df)
pc2_test <- kruskal.test(PC2 ~ disposition, data = plot_df)
pc3_test <- kruskal.test(PC3 ~ disposition, data = plot_df)

cat("Kruskal-Wallis Test - PC1 by Disposition:\n")
print(pc1_test)
cat("\nKruskal-Wallis Test - PC2 by Disposition:\n")
print(pc2_test)
cat("\nKruskal-Wallis Test - PC3 by Disposition:\n")
print(pc3_test)
cat("Holm-adjusted p-values for the three disposition comparisons:\n")
print(p.adjust(c(PC1=pc1_test$p.value, PC2=pc2_test$p.value, PC3=pc3_test$p.value), method="holm"))

```

**The Result:** Small p-values provide evidence that score distributions differ
in this sample. They do not quantify effect size, prove separability, or turn
visual overlap into noise; interpretation therefore also reports plots and
loading magnitudes.

---

# Section 12: The Loadings Plot—Visualizing Component Meanings

```{r loadings-biplot, fig.cap="Each arrow shows how original variables contribute to the new PCA axes"}
loadings_df <- as.data.frame(sweep(pca_result$rotation, 2, pca_result$sdev, "*")) %>%
  rownames_to_column(var = "Variable")

loadings_arrow_plot <- ggplot(loadings_df, aes(x = PC1, y = PC2, label = Variable)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.0), color = "gray", alpha = 0.3) +
  geom_segment(
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "blue"
  ) +
  geom_text_repel(
    segment.color = "gray",
    segment.size = 0.5
  ) +
  coord_fixed(ratio = 1) +
  labs(
    title = "Variable-PC Correlation Circle",
    subtitle = "How original variables project onto PC1 and PC2",
    x = "PC1 - Stellar Scale",
    y = "PC2 - Orbital Dynamics"
  ) +
  theme_minimal()

print(loadings_arrow_plot)
```

Arrow coordinates are correlations between each standardized variable and the PC scores. Angles in a two-dimensional projection only approximate relationships when both variables are well represented there. Orthogonality of PC scores does not establish independence of physical phenomena.

---

# Section 13: Comparing Components with Input-Derived Groups

Ridge plots describe score distributions across groups derived from the analysis inputs.

```{r ridge-plot-hot-star, fig.cap="PC-score distributions by input-derived temperature group"}
# Prepare data for ridge plots
plot_df$hot_star_label <- factor(plot_df$hot_star,
                                  levels = c(0, 1),
                                  labels = c("Cool Star", "Hot Star"))

plot_df_long <- plot_df %>%
  select(PC1, PC2, PC3, PC4, hot_star_label) %>%
  pivot_longer(
    cols = c(PC1, PC2, PC3, PC4),
    names_to = "Component",
    values_to = "Score"
  )

ridge_plot_hot <- ggplot(plot_df_long, aes(x = Score, y = hot_star_label, fill = hot_star_label)) +
  geom_density_ridges(alpha = 0.7) +
  facet_wrap(~ Component, scales = "free_x") +
  scale_fill_brewer(palette = "Set1") +
  coord_cartesian(xlim = c(-10, 10)) +
  labs(
    title = "Ridge Plot: Hot vs. Cool Stars Across PCA Components",
    x = "PC Score",
    y = "Stellar Temperature"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(ridge_plot_hot)
```

The ridge plots describe associations with temperature groups constructed from a PCA input. They do not independently validate the components.

```{r ridge-plot-insolation, fig.cap="PC-score distributions by input-derived insolation group"}
plot_df_long_insol <- plot_df %>%
  select(PC1, PC2, PC3, PC4, insolation_class) %>%
  mutate(insolation_class = factor(insolation_class, levels = c("low", "medium", "high"))) %>%
  pivot_longer(
    cols = c(PC1, PC2, PC3, PC4),
    names_to = "Component",
    values_to = "Score"
  )

ridge_plot_insol <- ggplot(plot_df_long_insol, aes(x = Score, y = insolation_class, fill = insolation_class)) +
  geom_density_ridges(alpha = 0.7) +
  facet_wrap(~ Component, scales = "free_x") +
  scale_fill_viridis_d(option = "C") +
  coord_cartesian(xlim = c(-10, 10)) +
  labs(
    title = "Ridge Plot: Insolation Class Across PCA Components",
    x = "PC Score",
    y = "Planetary Energy Level"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(ridge_plot_insol)
```

Insolation groups also reuse a PCA input. Their score distributions describe this sample and should be interpreted alongside the loadings.

---

# Section 14: Exploratory Distribution Comparisons

Let's run the Kruskal-Wallis test for all our domain-knowledge variables against all 4 components.

```{r comprehensive-tests}
# Create all combinations
variables <- c("large_planet", "hot_star", "insolation_class", "magnitude_class")
components <- c("PC1", "PC2", "PC3", "PC4")

all_combinations <- crossing(
  Variable_Group = variables,
  Component = components
)

# Run tests
all_tests_summary <- all_combinations %>%
  rowwise() %>%
  mutate(
    formula_str = paste(Component, "~", Variable_Group),
    test_result = list(tidy(kruskal.test(as.formula(formula_str), data = plot_df)))
  ) %>%
  ungroup() %>%
  unnest(test_result) %>%
  select(Variable_Group, Component, p.value) %>%
  mutate(p.value = p.adjust(p.value, method = "holm"))

# Create summary table
final_p_value_table <- all_tests_summary %>%
  mutate(p.value = format.pval(p.value, digits = 2, eps = 1e-100)) %>%
  pivot_wider(
    names_from = Component,
    values_from = p.value
  ) %>%
  select(Variable_Group, PC1, PC2, PC3, PC4)

knitr::kable(final_p_value_table, caption = "Holm-adjusted p-values for 16 exploratory Kruskal-Wallis comparisons")
```

## Interpreting the Results

The displayed p-values use Holm correction across all 16 comparisons. Small values indicate sample distribution differences, not separability or unbiased discovery. Derived groups reuse PCA inputs, and multiple KOIs can share a host star, so standard independent-observation p-values are descriptive approximations. A nonsignificant result does not prove independence.

---

# Conclusion

This report applies PCA to ten standardized inputs after log10(x+1) transformation of eight positive-scale variables. The first four PCs retain `r round(sum(pca_result$sdev[1:4]^2)/sum(pca_result$sdev^2)*100, 1)`% of variance. Loadings, score plots and derived groups describe this snapshot; they do not prove physical independence, classification accuracy or stellar evolutionary stages.

---

# Technical Appendix

## Session Information

```{r session-info}
session_lines <- capture.output(sessionInfo())
cat(paste(vapply(session_lines, function(line) paste(strwrap(line, width=78), collapse="\n"), character(1)), collapse="\n"))
```

## Complete Code Summary

This QMD file contains fully executable R code. In the HTML version, `code-fold: true` lets you view or hide the code.

Key libraries used:

- `tidyverse`: Data manipulation and visualization
- `ggplot2`: Advanced plotting
- `ggcorrplot`: Correlation heatmaps
- `GGally`: Pairs plots
- `plotly`: Interactive 3D visualization
- `ggridges`: Ridge plots for distribution visualization
- `broom`: Tidy statistical test results

---

**Created:** `r Sys.Date()`

**Analysis Purpose:** Dimensionality reduction and interpretation of exoplanet discovery data using Principal Component Analysis with explicit exploratory limitations.

