---
title: "The Matrix Chain Ordering Problem"
subtitle: "High-Performance Computing Challenge"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
    fig-width: 10
    fig-height: 6
    self-contained: true
    smooth-scroll: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Compiler flags: use an externally supplied configuration when available
# (for example, the Docker image enables OpenMP explicitly). Otherwise fall
# back to a fast local build that targets the current machine.
if (Sys.getenv("PKG_CXXFLAGS") == "") {
  Sys.setenv(PKG_CXXFLAGS = "-O3 -march=native")
}

library(Rcpp)
library(RcppArmadillo)
library(bench)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(knitr)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40", size = 12),
      panel.grid.minor = element_blank(),
      legend.position  = "top"
    )
)

palette_main <- c("#2E86AB", "#A23B72", "#F18F01", "#C73E1D", "#3B1F2B")

fmt_time <- function(x) {
  seconds <- as.numeric(x)
  if (seconds < 1e-6) return(paste0(round(seconds * 1e9, 1), " ns"))
  if (seconds < 1e-3) return(paste0(round(seconds * 1e6, 1), " \u00b5s"))
  if (seconds < 1) return(paste0(round(seconds * 1e3, 2), " ms"))
  paste0(round(seconds, 2), " s")
}

fmt_bytes_binary <- function(x) {
  bytes <- as.numeric(x)
  if (bytes < 1024) return(paste0(round(bytes, 1), " B"))
  if (bytes < 1024^2) return(paste0(round(bytes / 1024, 1), " KB"))
  if (bytes < 1024^3) return(paste0(round(bytes / 1024^2, 2), " MB"))
  paste0(round(bytes / 1024^3, 2), " GB")
}

fmt_ratio <- function(x) paste0(round(x, 1), "x")

fmt_relative_perf <- function(x) {
  if (x >= 1) {
    paste0(round(x, 1), "x faster")
  } else {
    paste0(round(1 / x, 1), "x slower")
  }
}

bench_value <- function(bm, label, column) {
  idx <- which(as.character(bm$expression) == label)
  bm[[column]][idx][1]
}

bench_cv <- function(bm) {
  sd_time <- vapply(bm$time, function(x) sd(as.numeric(x)), numeric(1))
  sd_time / as.numeric(bm$median)
}

blas_info <- extSoftVersion()[["BLAS"]]
```

# Introduction

This report addresses the **Matrix Chain Ordering Problem** (MCOP) as part of the High-Performance Computing course. The problem boils down to one question: given a chain of matrices to multiply, in what order should we group the multiplications to minimize computational cost?

Matrix multiplication is **associative**, so the final result does not depend on how we parenthesize the chain. But the number of scalar operations (and the size of intermediate matrices) can change wildly depending on the grouping. For $n$ matrices, the number of valid parenthesizations follows the Catalan numbers:

$$
C_{n-1} = \frac{1}{n}\binom{2(n-1)}{n-1}
$$

For $n = 5$ that gives us $C_4 = 14$ options, which is small enough to enumerate by brute force. For larger $n$, this grows exponentially and we need smarter algorithms.

The report is structured in four tasks:

1. **Task 1:** Find a compelling 5-matrix example where ordering matters, benchmark it in R.
2. **Task 2:** Implement a naive C++ multiplication using RcppArmadillo and compare.
3. **Task 3:** Build a C++ execution engine that multiplies matrices in any given order.
4. **Task 4:** Implement optimization algorithms (dynamic programming, parallelization, cache-aware heuristics) and compare everything end to end.

:::{.callout-note}
## How to read the benchmark tables

Each benchmark table in this report comes from a separate `bench::mark()` run executed during rendering. That means the **same technique can appear with slightly different absolute times in different sections**: cache state, garbage collection, BLAS thread scheduling, CPU frequency scaling, and the set of competing expressions all change from run to run.

So the safest interpretation is:

1. Compare entries **within the same table**.
2. Use medians rather than minimums.
3. Treat the **grand benchmark in Task 4** as the canonical cross-task comparison, because it places all approaches in one shared benchmark session.

To stabilize the measurements, every benchmark uses `min_time = 2`, so each expression runs for at least two seconds and accumulates iterations proportional to its speed. Each benchmark table includes a `CV (sd/median)` column so you can gauge stability at a glance; smaller values mean tighter timing dispersion. The `n_itr` column reports how many iterations were actually executed for each expression. I prefer `min_time` to a fixed `min_iterations` because a fixed iteration budget under-samples very fast operations and over-samples slow ones, but I also enforce `min_iterations = 50` to avoid single-iteration rows on slow cases. `bench::mark()` reports `n_itr` **after** optionally filtering out iterations that triggered garbage collection; to keep the floor intact, I disable GC filtering (`filter_gc = FALSE`) so the displayed `n_itr` matches the enforced minimum.

To reduce cold-start noise, the code warms up every strategy once before benchmarking. All benchmarks were produced on `r R.version$os` (`r R.version$arch`). Active BLAS: `r blas_info`.
:::

---

# Task 1: The Cost of Naive Ordering in R {#sec-task1}

The goal here is to find 5 matrices where R's default left-to-right evaluation (`A %*% B %*% C %*% D %*% E`) performs significantly worse than an optimal parenthesization. We need to design the dimensions carefully, enumerate all possible orderings, and benchmark the key cases.

## Choosing the dimensions

The trick is to alternate between **large** and **small** dimensions. This creates a "bottleneck" pattern where a naive left-to-right evaluation is forced to build huge intermediate matrices, while a smarter grouping can keep everything small by multiplying through the narrow dimensions first.

```{r}
#| label: task1-dims

# Dimension vector: A is p0×p1, B is p1×p2, ..., E is p4×p5
# We alternate large-small to create the bottleneck pattern
dims <- c(800, 5, 900, 8, 700, 6)

set.seed(42)
A <- matrix(rnorm(dims[1] * dims[2]), nrow = dims[1], ncol = dims[2])
B <- matrix(rnorm(dims[2] * dims[3]), nrow = dims[2], ncol = dims[3])
C <- matrix(rnorm(dims[3] * dims[4]), nrow = dims[3], ncol = dims[4])
D <- matrix(rnorm(dims[4] * dims[5]), nrow = dims[4], ncol = dims[5])
E <- matrix(rnorm(dims[5] * dims[6]), nrow = dims[5], ncol = dims[6])

tibble(
  Matrix     = LETTERS[1:5],
  Dimensions = paste0(dims[1:5], " × ", dims[2:6]),
  Elements   = dims[1:5] * dims[2:6]
) |> kable(
  format.args = list(big.mark = ","),
  align = "lcc"
)
```

Notice the pattern: $A$ is $800 \times 5$ (tall and skinny), $B$ is $5 \times 900$ (short and wide), $C$ is $900 \times 8$ (tall and skinny again), and so on. If R evaluates left to right, the very first step computes $AB$, which produces an $800 \times 900 = 720{,}000$ element intermediate. A smarter approach would start with $BC$ ($5 \times 8 = 40$ elements) or $DE$ ($8 \times 6 = 48$ elements), keeping intermediates tiny.

## Enumerating all 14 parenthesizations

Since $C_4 = 14$, we can just generate all of them, compute the theoretical operation count for each, and sort them. Each parenthesization is represented internally as a **split table**: a mapping from sub-chain `[i, j]` to the split index `k` where the final multiplication happens.

```{r}
#| label: task1-enumerate

# Count scalar multiplications for a given split table
count_ops <- function(dims, splits, i = 1, j = length(dims) - 1) {
  if (i == j) return(0)
  k <- splits[[paste(i, j)]]
  left  <- count_ops(dims, splits, i, k)
  right <- count_ops(dims, splits, k + 1, j)
  cost  <- dims[i] * dims[k + 1] * dims[j + 1]
  left + right + cost
}

# Generate all valid parenthesizations for n matrices
generate_all_parens <- function(n) {
  get_all <- function(i, j) {
    if (i == j) return(list(list()))
    combos <- list()
    for (k in i:(j - 1)) {
      lefts  <- get_all(i, k)
      rights <- get_all(k + 1, j)
      for (l in lefts) {
        for (r in rights) {
          combo <- c(l, r, setNames(list(k), paste(i, j)))
          combos[[length(combos) + 1]] <- combo
        }
      }
    }
    combos
  }
  raw <- get_all(1, n)
  lapply(raw, function(x) {
    splits <- list()
    for (nm in names(x)) splits[[nm]] <- x[[nm]]
    splits
  })
}

all_parens <- generate_all_parens(5)

# Build a human-readable label: (A · (B · C)) style
paren_label <- function(splits, i = 1, j = 5) {
  if (i == j) return(LETTERS[i])
  k <- splits[[paste(i, j)]]
  left  <- paren_label(splits, i, k)
  right <- paren_label(splits, k + 1, j)
  paste0("(", left, " · ", right, ")")
}

# Build the comparison table
paren_df <- tibble(
  id    = seq_along(all_parens),
  label = sapply(all_parens, paren_label),
  ops   = sapply(all_parens, function(s) count_ops(dims, s))
) |> arrange(ops) |>
  mutate(
    rank     = row_number(),
    is_best  = ops == min(ops),
    is_worst = ops == max(ops),
    is_naive = label == "((((A · B) · C) · D) · E)",
    highlight = case_when(
      is_best  ~ "★ Optimal",
      is_worst ~ "✗ Worst",
      is_naive ~ "→ Naive",
      TRUE     ~ ""
    )
  )

paren_df |>
  select(rank, label, ops, highlight) |>
  kable(
    col.names = c("Rank", "Parenthesization", "Scalar multiplications", ""),
    format.args = list(big.mark = ","),
    align = "clrc"
  )
```

The table tells a clear story. The optimal parenthesization `(A · ((B · C) · (D · E)))` requires only **93,840** scalar multiplications, while R's default left-to-right evaluation needs **17,200,000**, roughly **183 times more**. The worst case `(((A · B) · (C · D)) · E)` explodes to over **516 million** operations because it generates both an $800 \times 900$ and a $900 \times 700$ intermediate before even reaching the final product.

Why is the optimal ordering so cheap? It starts with the two "inner" products $B \cdot C$ ($5 \times 8$, costing 36,000 ops) and $D \cdot E$ ($8 \times 6$, costing 33,600 ops), which are both tiny. Then it combines them into a $5 \times 6$ matrix (240 ops), and only at the very end does it multiply the large $A$ into the result ($800 \times 5 \times 6 = 24{,}000$ ops). The narrow inner dimensions act as a funnel that keeps everything small.

```{r}
#| label: task1-ops-plot
#| fig-cap: "Operation count for all 14 parenthesizations of 5 matrices."
#| fig-height: 7

ggplot(paren_df, aes(x = reorder(label, ops), y = ops,
                     fill = case_when(
                       is_best  ~ "Optimal",
                       is_worst ~ "Worst",
                       is_naive ~ "Naive (L-to-R)",
                       TRUE     ~ "Other"
                     ))) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = format(ops, big.mark = ",")),
    hjust = -0.1, size = 3, fontface = "bold"
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(
    values = c("Optimal" = palette_main[1], "Worst" = palette_main[4],
               "Naive (L-to-R)" = palette_main[3], "Other" = "grey80"),
    name = NULL
  ) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Operation Count: All 14 Parenthesizations",
    subtitle = paste0("Chain dimensions: ", paste(dims, collapse = " × ")),
    x = NULL, y = "Scalar multiplications"
  )
```

The plot makes the disparity even more visual. The worst case is completely off the chart compared to everything else, and even the naive ordering is far from optimal. Most of the "reasonable" parenthesizations cluster in the low millions, but the optimal one sits at under 100K, an order of magnitude below the rest.

## Benchmarking with `bench::mark()`

Theory is nice, but let us see if the operation counts actually translate into real wall-clock time. We benchmark the three key cases: naive, optimal, and worst.

```{r}
#| label: task1-bench-setup

# Build R expressions from split tables so we can eval() them in the benchmark
build_expr <- function(splits, i = 1, j = 5) {
  if (i == j) return(as.name(LETTERS[i]))
  k <- splits[[paste(i, j)]]
  left  <- build_expr(splits, i, k)
  right <- build_expr(splits, k + 1, j)
  call("%*%", left, right)
}

best_splits  <- all_parens[[paren_df$id[paren_df$is_best][1]]]
worst_splits <- all_parens[[paren_df$id[paren_df$is_worst][1]]]

best_expr  <- build_expr(best_splits)
worst_expr <- build_expr(worst_splits)
naive_expr <- quote(A %*% B %*% C %*% D %*% E)
```

```{r}
#| label: task1-benchmark

invisible(eval(naive_expr))
invisible(eval(best_expr))
invisible(eval(worst_expr))
gc()

bm <- bench::mark(
  `Naive (L-to-R)` = eval(naive_expr),
  `Optimal`        = eval(best_expr),
  `Worst`          = eval(worst_expr),
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm |>
  mutate(
    expression = as.character(expression),
    `CV (sd/median)` = round(bench_cv(bm), 3)
  ) |>
  select(expression, min, median, `CV (sd/median)`, mem_alloc, n_itr) |>
  kable(caption = "Benchmark: time and memory for each parenthesization strategy",
        align = "lccccc")
```

The benchmark confirms the same ordering we saw in the theoretical FLOP counts. In this render, the median times are `r fmt_time(bench_value(bm, "Optimal", "median"))` for the optimal ordering, `r fmt_time(bench_value(bm, "Naive (L-to-R)", "median"))` for the naive ordering, and `r fmt_time(bench_value(bm, "Worst", "median"))` for the worst case. The exact values will vary with the machine and BLAS backend, but the pattern is stable: the optimal plan is dramatically faster because it avoids both extra multiplications and huge intermediates. R-side memory tracking shows the same trend, with `r fmt_bytes_binary(bench_value(bm, "Optimal", "mem_alloc"))` for the optimal case versus `r fmt_bytes_binary(bench_value(bm, "Naive (L-to-R)", "mem_alloc"))` and `r fmt_bytes_binary(bench_value(bm, "Worst", "mem_alloc"))` for the naive and worst cases. The CV column is the coefficient of variation (`sd/median`), and `n_itr` shows how many iterations were run after applying the timing and iteration floors. Note that the optimal ordering consistently shows a higher CV than the naive and worst cases: at ~`r fmt_time(bench_value(bm, "Optimal", "median"))`, a single OS scheduling interrupt or cache eviction represents a larger fraction of the total measurement than it does for operations that take tens of milliseconds. The absolute jitter is similar across strategies; only the relative measure (CV) amplifies it for the fast case.

```{r}
#| label: task1-violin
#| fig-cap: "Violin plot of execution times (log scale). The three strategies span almost four orders of magnitude."

autoplot(bm, type = "violin") +
  labs(
    title    = "Execution Time Distribution",
    subtitle = "Each strategy benchmarked for at least 2 seconds",
    x = NULL
  ) +
  scale_fill_manual(values = palette_main[c(3, 1, 4)]) +
  theme(legend.position = "none")
```

The violin plot shows how tight the distributions are: in this render there is no overlap between the three strategies. The optimal ordering lives entirely in the sub-millisecond range, the naive clusters around `r fmt_time(bench_value(bm, "Naive (L-to-R)", "median"))`, and the worst case reaches `r fmt_time(bench_value(bm, "Worst", "median"))`. The observed time ratio between naive and worst is `r fmt_ratio(as.numeric(bench_value(bm, "Worst", "median")) / as.numeric(bench_value(bm, "Naive (L-to-R)", "median")))`, somewhat below the theoretical 30x suggested by raw FLOP counts — BLAS threads absorb part of the penalty when accumulating large intermediates, which softens the wall-clock gap relative to the scalar-operation ratio. This is not noise or measurement error; the difference is structural.

## Why does it matter? Intermediate matrix sizes

The raw operation count does not tell the whole story. A bad ordering also means **huge intermediate matrices** that eat memory and trash the CPU cache. Let us trace the exact dimensions of every intermediate produced by each strategy.

```{r}
#| label: task1-intermediates

# Trace intermediate matrix dimensions for a given split table
trace_intermediates <- function(splits, dims, i = 1, j = 5) {
  if (i == j) return(list(intermediates = list(), dim = c(dims[i], dims[i + 1])))
  k <- splits[[paste(i, j)]]
  left  <- trace_intermediates(splits, dims, i, k)
  right <- trace_intermediates(splits, dims, k + 1, j)
  result_dim <- c(dims[i], dims[j + 1])
  intermediates <- c(
    left$intermediates, right$intermediates,
    list(list(
      rows = result_dim[1], cols = result_dim[2],
      elements = result_dim[1] * result_dim[2],
      bytes = result_dim[1] * result_dim[2] * 8
    ))
  )
  list(intermediates = intermediates, dim = result_dim)
}

naive_splits_full <- all_parens[[paren_df$id[paren_df$is_naive][1]]]

strategies <- list(
  "Optimal"        = best_splits,
  "Naive (L-to-R)" = naive_splits_full,
  "Worst"          = worst_splits
)

inter_df <- bind_rows(lapply(names(strategies), function(strat) {
  inters <- trace_intermediates(strategies[[strat]], dims)$intermediates
  tibble(
    strategy  = strat,
    step      = seq_along(inters),
    dim_label = sapply(inters, \(x) paste0(x$rows, " × ", x$cols)),
    elements  = sapply(inters, \(x) x$elements),
    kb        = sapply(inters, \(x) x$bytes / 1024)
  )
}))

inter_df |>
  kable(
    col.names = c("Strategy", "Step", "Intermediate size", "Elements", "Memory (KB)"),
    format.args = list(big.mark = ","),
    digits = 1,
    align = "clcrc",
    caption = "Intermediate matrices generated at each multiplication step"
  )
```

This table is the key to understanding everything. Look at the optimal strategy: its first three intermediates are $5 \times 8$, $8 \times 6$, and $5 \times 6$, all under 50 elements. Only the final product ($800 \times 6$) is moderately sized. The naive strategy, on the other hand, starts with an $800 \times 900$ monster (720,000 elements, 5,625 KB) on the very first step.

```{r}
#| label: task1-memory-plot
#| fig-cap: "Peak and total memory allocated for intermediate matrices. The optimal ordering uses orders of magnitude less memory."
#| fig-height: 5

mem_summary <- inter_df |>
  summarize(
    peak_kb  = max(kb),
    total_kb = sum(kb),
    .by = strategy
  ) |>
  pivot_longer(cols = c(peak_kb, total_kb),
               names_to = "metric", values_to = "kb") |>
  mutate(
    metric = if_else(metric == "peak_kb", "Peak intermediate", "Total (all intermediates)"),
    strategy = factor(strategy, levels = c("Optimal", "Naive (L-to-R)", "Worst"))
  )

ggplot(mem_summary, aes(x = strategy, y = kb, fill = strategy)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(kb, 1), " KB")),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(
    values = c("Optimal" = palette_main[1], "Naive (L-to-R)" = palette_main[3],
               "Worst" = palette_main[4])
  ) +
  scale_y_continuous(labels = label_comma(suffix = " KB"),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Memory Cost of Intermediate Matrices",
    subtitle = "Peak = largest single intermediate; Total = sum of all intermediates",
    x = NULL, y = NULL
  ) +
  theme(legend.position = "none")
```

The chart makes the difference concrete. The optimal strategy's peak intermediate is just **37.5 KB** (the final $800 \times 6$ result), while the naive and worst both peak at **5,625 KB** ($800 \times 900$). In total, the optimal strategy allocates about **38.4 KB** across all intermediates, versus over **10,000 KB** for the naive. That is a ~263x difference in total memory, which directly impacts cache behavior: the optimal intermediates fit comfortably in L1 cache (typically 32–64 KB per core), while the naive ones spill into L2 or even L3.

## Summary

To wrap up Task 1:

```{r}
#| label: task1-summary

peak_kb <- function(strategy) {
  mem_summary$kb[mem_summary$strategy == strategy &
                   mem_summary$metric == "Peak intermediate"][1]
}

task1_summary <- tibble(
  Metric = c("Scalar ops", "Median time", "Memory allocated", "Peak intermediate"),
  Optimal = c(
    format(min(paren_df$ops), big.mark = ","),
    fmt_time(bench_value(bm, "Optimal", "median")),
    fmt_bytes_binary(bench_value(bm, "Optimal", "mem_alloc")),
    paste0(round(peak_kb("Optimal"), 1), " KB")
  ),
  `Naive (L-to-R)` = c(
    format(paren_df$ops[paren_df$is_naive], big.mark = ","),
    fmt_time(bench_value(bm, "Naive (L-to-R)", "median")),
    fmt_bytes_binary(bench_value(bm, "Naive (L-to-R)", "mem_alloc")),
    paste0(round(peak_kb("Naive (L-to-R)"), 1), " KB")
  ),
  Worst = c(
    format(max(paren_df$ops), big.mark = ","),
    fmt_time(bench_value(bm, "Worst", "median")),
    fmt_bytes_binary(bench_value(bm, "Worst", "mem_alloc")),
    paste0(round(peak_kb("Worst"), 1), " KB")
  ),
  `Naive/Optimal ratio` = c(
    fmt_ratio(paren_df$ops[paren_df$is_naive] / min(paren_df$ops)),
    fmt_ratio(as.numeric(bench_value(bm, "Naive (L-to-R)", "median")) /
                as.numeric(bench_value(bm, "Optimal", "median"))),
    fmt_ratio(as.numeric(bench_value(bm, "Naive (L-to-R)", "mem_alloc")) /
                as.numeric(bench_value(bm, "Optimal", "mem_alloc"))),
    fmt_ratio(peak_kb("Naive (L-to-R)") / peak_kb("Optimal"))
  )
)

task1_summary |>
  kable(align = "lrrrr", caption = "Task 1 summary derived from the rendered benchmark")
```

The conclusion is straightforward: **parenthesization order is not a minor optimization**. For matrices with heterogeneous dimensions, the wrong ordering can easily cost two to three orders of magnitude in both time and memory. R's default left-to-right `%*%` evaluation offers no protection against this. In the following tasks, we will build tools to automatically find and execute the optimal ordering.

---

# Task 2: C++ Naive Multiplication with Armadillo {#sec-task2}

Now we move from R to C++. The question is: does switching to a compiled language fix the ordering problem? Spoiler: not really. The bottleneck is algorithmic (how many FLOPs you do), not just interpretive overhead. But Armadillo does help in other ways, particularly memory management.

## The C++ naive function

We implement a simple function that takes a list of matrices and multiplies them left to right, exactly like R's `%*%` chaining. The function uses RcppArmadillo, which wraps LAPACK/BLAS for the heavy lifting. We also disable Armadillo's bounds checking with `ARMA_NO_DEBUG` for maximum speed.

```{Rcpp}
//| label: task2-cpp-naive

// [[Rcpp::depends(RcppArmadillo)]]
#define ARMA_NO_DEBUG
#include <RcppArmadillo.h>
using namespace Rcpp;

// [[Rcpp::export]]
arma::mat naive_multiply_cpp(const List& matrices) {
  int n = matrices.size();
  // Start with the first matrix and keep folding the chain left to right.
  arma::mat result = as<arma::mat>(matrices[0]);
  for (int i = 1; i < n; i++) {
    // This mirrors A %*% B %*% C ... exactly: no regrouping involved.
    result = result * as<arma::mat>(matrices[i]);
  }
  return result;
}
```

The function is intentionally simple. It accepts a generic `List` from R, casts each element to an `arma::mat`, and accumulates the product left to right. There is no reordering, no parenthesization logic. This mirrors what R does with `A %*% B %*% C %*% D %*% E`, but in compiled C++ with Armadillo's memory management instead of R's allocator.

## Benchmarking C++ naive vs. R

We compare three cases: R naive (left to right), R optimal (manually parenthesized), and C++ naive (left to right via Armadillo). This lets us isolate the effect of switching language while keeping the same algorithm.

```{r}
#| label: task2-benchmark

mat_list <- list(A, B, C, D, E)

invisible(eval(naive_expr))
invisible(eval(best_expr))
invisible(naive_multiply_cpp(mat_list))
gc()

bm_task2 <- bench::mark(
  `R naive`     = eval(naive_expr),
  `R optimal`   = eval(best_expr),
  `C++ naive`   = naive_multiply_cpp(mat_list),
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm_task2 |>
  mutate(
    expression = as.character(expression),
    `CV (sd/median)` = round(bench_cv(bm_task2), 3)
  ) |>
  select(expression, min, median, `CV (sd/median)`, mem_alloc, n_itr) |>
  kable(caption = "Task 2 benchmark: R vs. C++ naive multiplication",
        align = "lccccc")
```

In this benchmark run, the C++ naive function is modestly faster than R’s naive evaluation. Both perform the same sequence of multiplications (left to right), so both produce the same enormous intermediate matrices and do the same number of FLOPs. The speed improvement comes from two sources:

1. **Lower overhead**: Armadillo avoids R's SEXP allocation and garbage collector for intermediates. In R, each `%*%` call allocates a new R matrix object, which goes through R's memory management. Armadillo handles intermediate storage internally in C++ heap memory.
2. **Same BLAS backend**: both R and Armadillo ultimately call the same underlying BLAS routines for the actual matrix multiplication. This is why the speed difference is modest rather than dramatic.

The memory column tells the most interesting story. Here, R's naive approach reports about 10 MB allocated (all the large R matrix objects for intermediates), while `C++ naive` stays at only about 43 KB on the R side because Armadillo manages its own memory below R's tracking radar. The 38.61 KB shown for `R optimal` is small for a different reason: that ordering never materializes the huge intermediates in the first place. The actual memory consumption inside C++ is similar (those huge intermediates still exist), but R's allocator is not involved.

```{r}
#| label: task2-violin
#| fig-cap: "Execution time comparison: R naive vs. R optimal vs. C++ naive. The C++ version is only a modest improvement for the naive ordering, while the optimal parenthesization dominates both."

autoplot(bm_task2, type = "violin") +
  labs(
    title    = "C++ Naive vs. R: Does the Language Matter?",
    subtitle = "Same naive ordering, different runtime environments",
    x = NULL
  ) +
  scale_fill_manual(values = palette_main[c(3, 1, 4)]) +
  theme(legend.position = "none")
```

The violin plot makes the key point of this task visually clear: **switching to C++ does not solve the ordering problem**. The two naive strategies live in the same performance band, while the manually parenthesized R solution remains much faster because it is solving a different computational problem: it performs far fewer scalar multiplications.

```{r}
#| label: task2-ratio-plot
#| fig-cap: "Speedup ratios relative to R naive. Reordering the multiplications matters far more than switching language."

bm_ratios <- bm_task2 |>
  mutate(
    expression = as.character(expression),
    median_ns  = as.numeric(median)
  ) |>
  mutate(
    baseline  = median_ns[expression == "R naive"],
    speedup   = baseline / median_ns
  )

ggplot(bm_ratios, aes(x = reorder(expression, -speedup), y = speedup,
                       fill = expression)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(speedup, 1), "x")),
            vjust = -0.3, size = 4, fontface = "bold") +
  scale_fill_manual(values = c(
    "R naive"    = palette_main[3],
    "R optimal"  = palette_main[1],
    "C++ naive"  = palette_main[4]
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Speedup Relative to R Naive",
    subtitle = "Algorithm beats language: optimal ordering wins by orders of magnitude",
    x = NULL, y = "Speedup factor (higher = faster)"
  ) +
  theme(legend.position = "none")
```

## Discussion

The results of Task 2 confirm an important principle: **algorithm dominates implementation**. Switching from R to C++ with the same naive ordering keeps us in the same broad performance band, while reordering the multiplications in pure R changes the cost by orders of magnitude. The lesson is clear:

- A bad algorithm in a fast language (C++ naive) loses to a good algorithm in a slow language (R optimal).
- Any C++ advantage is secondary: lower allocation overhead, no GC pressure, tighter memory management. For small chains like ours ($n = 5$), this does not add up to much.
- Where C++ *will* shine is in Tasks 3 and 4, where we combine the optimal ordering with compiled execution. That gives us both the algorithmic advantage and the implementation advantage.

The next step is obvious: we need a way to execute *any* ordering efficiently in C++, not just the hard-coded naive one. That is what Task 3's stack machine is for.

---

# Task 3: A General-Purpose Execution Engine {#sec-task3}

In Task 2 we hard-coded the naive ordering into C++. That is obviously not scalable: we cannot write a different C++ function for each of the 14 parenthesizations, let alone for chains of 20 or 100 matrices. What we need is a **generic execution engine** that takes any ordering as input and executes it efficiently.

The conceptual issue is this: a parenthesization such as `(A * ((B * C) * (D * E)))` is easy for a human to read, but awkward to hand over to a low-level function. C++ wants a compact, explicit execution plan, not a string that must be parsed or a separate hard-coded function for every case.

So from Task 3 onward we separate the problem into two stages:

1. **Planning:** represent the order of multiplications in a machine-friendly format.
2. **Execution:** feed that plan to a generic C++ engine that performs the multiplications in exactly that order.

## The RPN approach

The standard way to do this is to encode the parenthesization as a **Reverse Polish Notation (RPN)** vector, which is just the post-order traversal of the expression tree. If you have not seen RPN before, the key idea is simple: operands appear first, and the operation appears only after the data it needs is already available. That makes it perfect for stack-based execution.

In our encoding:

- A positive integer $i$ means "push matrix $A_i$ onto the stack."
- The value $-1$ means "pop two matrices, multiply them, push the result."

For example, the optimal ordering `(A * ((B * C) * (D * E)))` becomes:

$$
\underbrace{1}_{\text{push A}},\;
\underbrace{2, 3, -1}_{\text{B*C}},\;
\underbrace{4, 5, -1}_{\text{D*E}},\;
\underbrace{-1}_{\text{(BC)*(DE)}},\;
\underbrace{-1}_{\text{A*(BCDE)}}
$$

This representation has two practical advantages. First, it is compact: for $n$ matrices, the vector has length $2n - 1$. Second, it is executable without recursion or expression parsing. We first build a helper in R to convert any split table into an RPN vector:

```{r}
#| label: task3-rpn-builder

# Post-order traversal of the parenthesization tree -> RPN vector
# Positive integers = 1-based matrix indices, -1 = multiply opcode
build_rpn <- function(splits, i = 1, j = length(dims) - 1) {
  if (i == j) return(i)
  # Each split table entry stores where the final multiplication happens.
  k <- splits[[paste(i, j)]]
  c(build_rpn(splits, i, k), build_rpn(splits, k + 1, j), -1L)
}

rpn_optimal <- build_rpn(best_splits)
rpn_naive   <- build_rpn(naive_splits_full)

tibble(
  Ordering = c("Optimal", "Naive (L-to-R)"),
  Parenthesization = c(
    paren_df$label[paren_df$is_best],
    paren_df$label[paren_df$is_naive]
  ),
  `RPN vector` = c(
    paste(rpn_optimal, collapse = ", "),
    paste(rpn_naive, collapse = ", ")
  )
) |> kable(align = "llc")
```

The naive ordering pushes each matrix and immediately multiplies: `1, 2, -1, 3, -1, ...` (strictly left-associative). The optimal ordering pushes groups first and multiplies them internally before combining: `1, 2, 3, -1, 4, 5, -1, -1, -1`.

It is useful to see why this works mechanically. For the optimal RPN vector:

1. `1` pushes `A`.
2. `2, 3, -1` pushes `B` and `C`, then replaces them with `BC`.
3. `4, 5, -1` pushes `D` and `E`, then replaces them with `DE`.
4. The next `-1` multiplies the two intermediate results, producing `(BC)(DE)`.
5. The final `-1` multiplies `A` by that result.

At every multiplication step, the two operands we need are already on top of the stack. That is exactly why post-order traversal and stack evaluation fit together so well.

## The C++ stack machine

The executor is a classic stack machine. A **stack** is a last-in, first-out structure: the last matrix pushed is the first one popped. That is exactly what we need here, because each `-1` token should combine the two most recently available sub-results.

The loop is therefore conceptually very simple:

1. Read the next token.
2. If it is positive, push the corresponding matrix.
3. If it is `-1`, pop the top two matrices, multiply them, and push the result back.

We also use `std::move()` when popping from the stack to avoid unnecessary copies of large intermediate matrices. In plain language, move semantics let us transfer ownership of a matrix buffer instead of duplicating all of its data.

```{Rcpp}
//| label: task3-stack-machine

// [[Rcpp::depends(RcppArmadillo)]]
#define ARMA_NO_DEBUG
#include <RcppArmadillo.h>
#include <stack>
#include <vector>
using namespace Rcpp;

// [[Rcpp::export]]
arma::mat execute_rpn(const List& matrices, const IntegerVector& rpn) {
  int n = matrices.size();
  
  // --- Zero-copy pre-conversion ---
  // Build lightweight arma::mat wrappers that REUSE R's memory.
  // The constructor mat(ptr, n_rows, n_cols, copy_aux_mem=false, strict=true)
  // points Armadillo directly at R's column-major SEXP buffer,
  // avoiding the deep copy that as<arma::mat>() would perform.
  std::vector<arma::mat> mats;
  mats.reserve(n);
  for (int i = 0; i < n; i++) {
    NumericMatrix rm = matrices[i];
    arma::mat wrapper(rm.begin(), rm.nrow(), rm.ncol(), false, true);
    mats.push_back(wrapper);
  }
  
  // --- Stack machine ---
  std::stack<arma::mat> stk;
  
  for (int t = 0; t < rpn.size(); t++) {
    int token = rpn[t];
    if (token >= 0) {
      // Positive: push a COPY of the wrapper (cheap: the input matrices
      // must not be modified, so we copy once here; intermediates are
      // then moved, not copied, for the rest of the computation).
      stk.push(mats[token - 1]);
    } else {
      // -1: pop two, multiply, push result
      arma::mat right = std::move(stk.top()); stk.pop();
      arma::mat left  = std::move(stk.top()); stk.pop();
      stk.push(left * right);
    }
  }
  
  return stk.top();
}
```

A few things worth noting about this implementation:

1. **Zero-copy input conversion**: a naive approach would call `as<arma::mat>(matrices[token - 1])` inside the token loop, which performs a **deep copy** on every push. For a 500 MB matrix, that would double memory usage and turn the benchmark into a memory-bandwidth test rather than a computation test. Instead, we pre-convert all matrices into lightweight `arma::mat` wrappers using Armadillo's advanced constructor with `copy_aux_mem = false`. These wrappers point directly at R's column-major SEXP memory, eliminating the R-to-C++ transfer overhead entirely.
2. **`std::move(stk.top())`**: when we pop a matrix from the stack, we use move semantics instead of copying. For an $800 \times 900$ intermediate, this means transferring a pointer (8 bytes) instead of copying 5.7 MB of data.
3. **`const List&`**: the input list is passed by reference, so the R matrices are not copied when entering the C++ function.
4. **Token-driven loop**: the function does not know or care what the "optimal" ordering is. It simply follows instructions. This is what makes it generic: any valid RPN vector produces the correct result.

```{r}
#| label: task3-verify

# Verify the stack machine gives the right answer for both orderings
res_rpn_opt   <- execute_rpn(mat_list, rpn_optimal)
res_rpn_naive <- execute_rpn(mat_list, rpn_naive)

tibble(
  Test = c("RPN optimal vs R optimal", "RPN naive vs R naive"),
  `Max absolute difference` = c(
    max(abs(res_rpn_opt - eval(best_expr))),
    max(abs(res_rpn_naive - eval(naive_expr)))
  )
) |> kable(align = "lc")
```

Both comparisons return zero maximum difference, so the stack machine is not just faster or more flexible: it is numerically identical to the corresponding R evaluation.

## Benchmarking: stack machine vs. everything else

Now the interesting question: how does the generic stack machine compare to the hard-coded approaches from Tasks 1 and 2? We test five configurations.

```{r}
#| label: task3-benchmark

invisible(eval(naive_expr))
invisible(eval(best_expr))
invisible(naive_multiply_cpp(mat_list))
invisible(execute_rpn(mat_list, rpn_naive))
invisible(execute_rpn(mat_list, rpn_optimal))
gc()

bm_task3 <- bench::mark(
  `R naive`            = eval(naive_expr),
  `R optimal`          = eval(best_expr),
  `C++ naive (Task 2)` = naive_multiply_cpp(mat_list),
  `C++ RPN naive`      = execute_rpn(mat_list, rpn_naive),
  `C++ RPN optimal`    = execute_rpn(mat_list, rpn_optimal),
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm_task3 |>
  mutate(
    expression = as.character(expression),
    `CV (sd/median)` = round(bench_cv(bm_task3), 3)
  ) |>
  select(expression, min, median, `CV (sd/median)`, mem_alloc, n_itr) |>
  kable(caption = "Task 3 benchmark: stack machine vs. previous approaches",
        align = "lccccc")
```

There are two comparisons worth making here:

1. **C++ RPN naive vs C++ naive (Task 2):** the two are very close in the table above. That means the extra abstraction of reading an instruction vector and using a stack does not materially hurt performance.

2. **C++ RPN optimal vs R optimal:** both are much faster than the naive variants because both are executing the same good multiplication order. Any small difference now comes from execution overhead, not from algorithmic cost.

The key takeaway is that **the stack machine buys us generality almost for free**. We can now execute any parenthesization encoded as an RPN vector, which is exactly the interface Task 4 will need.

```{r}
#| label: task3-comparison-plot
#| fig-cap: "Full comparison across Tasks 1--3. The stack machine (RPN) matches the hard-coded approaches while being fully generic."

bm_plot_df <- bm_task3 |>
  mutate(
    expression = as.character(expression),
    median_us  = as.numeric(median) * 1e6,
    group = case_when(
      grepl("naive", expression, ignore.case = TRUE) &
        !grepl("optimal", expression, ignore.case = TRUE) ~ "Naive ordering",
      TRUE ~ "Optimal ordering"
    )
  )

ggplot(bm_plot_df, aes(x = reorder(expression, median_us),
                        y = median_us, fill = group)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = if_else(
    median_us > 1000,
    paste0(round(median_us / 1000, 1), " ms"),
    paste0(round(median_us), " \u00b5s")
  )), hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip(clip = "off") +
  scale_fill_manual(
    values = c("Naive ordering" = palette_main[3], "Optimal ordering" = palette_main[1]),
    name = NULL
  ) +
  scale_y_log10(
    labels = function(x) if_else(x >= 1000, paste0(x/1000, " ms"), paste0(x, " \u00b5s")),
    expand = expansion(mult = c(0, 0.3))
  ) +
  labs(
    title    = "Full Benchmark: Tasks 1 through 3",
    subtitle = "Log scale. Ordering dominates; the stack machine adds no overhead.",
    x = NULL, y = "Median time (log scale)"
  )
```

The log-scale bar chart tells the whole story of Tasks 1--3. The naive approaches cluster together, and the optimal approaches cluster together. The separating factor is the **algorithm** (which multiplications we choose), not the **language** (R vs C++) or the **execution mechanism** (direct loop vs stack machine).

## Discussion

Task 3 gives us a powerful building block: a generic, zero-overhead execution engine that can run any parenthesization encoded as an RPN vector. This is important because:

- We no longer need to generate and `eval()` R expressions, which was a bit of a hack in Task 1.
- The same C++ function handles any chain length, not just 5 matrices.
- The RPN encoding is compact ($2n - 1$ integers) and easy to produce from a DP algorithm.

What we still lack is the algorithm that *finds* the optimal RPN vector automatically. Right now we are computing it in R by enumerating all 14 parenthesizations, which is $O(C_n)$ and only feasible for tiny $n$. For larger chains, we need dynamic programming. That is the subject of Task 4.

---

# Task 4: Finding the Optimal Order Automatically {#sec-task4}

This is the core algorithmic task. We need to go from "try all parenthesizations" (exponential) to something that scales. The classical answer is dynamic programming in $O(n^3)$, which we implement first in R, then in C++. After that, we push further: a parallel wavefront variant that exploits multi-core CPUs, and a resource-aware cost function inspired by recent HPC literature that accounts for memory hierarchy effects.

From a teaching point of view, this section answers three different questions:

1. **Why does dynamic programming apply here at all?**
2. **How do we reconstruct an actual multiplication plan, not just a cost?**
3. **Why might a hardware-aware objective differ from pure FLOP counting?**

## The Godbole algorithm: $O(n^3)$ dynamic programming

Dynamic programming works because the matrix-chain problem has **optimal substructure** and **overlapping subproblems**.

- **Optimal substructure** means that if the best way to compute $A_i \cdots A_j$ splits at position $k$, then the left side $A_i \cdots A_k$ and the right side $A_{k+1} \cdots A_j$ must themselves also be solved optimally.
- **Overlapping subproblems** means the same sub-chains appear again and again across different parenthesizations. Instead of recomputing them exponentially many times, we compute each interval once and store the answer.

The idea is therefore to fill two tables bottom-up:

- **Cost table $m[i,j]$**: the minimum number of scalar multiplications needed to compute the sub-chain $A_i \cdots A_j$.
- **Split table $s[i,j]$**: the index $k$ where we should split the sub-chain to achieve that minimum.

The recurrence is:

$$
m[i, j] = \min_{i \le k < j} \left\{ m[i, k] + m[k+1, j] + p_{i-1} \cdot p_k \cdot p_j \right\}
$$

with base case $m[i, i] = 0$. We iterate by chain length: first all pairs ($L = 2$), then triples ($L = 3$), and so on up to the full chain ($L = n$). Each cell only depends on shorter intervals, so a bottom-up traversal guarantees that every dependency has already been computed when we need it.

This is the same dependency pattern we will later exploit for parallelization: every diagonal of the table can be computed once the previous diagonals are known.

### R implementation

```{r}
#| label: task4-dp-r

mcm_dp_r <- function(dims) {
  n <- length(dims) - 1
  m <- matrix(0, nrow = n, ncol = n)
  s <- matrix(0L, nrow = n, ncol = n)
  
  # Fill the DP tables by diagonal: short chains first, long chains later.
  for (len in 2:n) {
    for (i in 1:(n - len + 1)) {
      j <- i + len - 1
      m[i, j] <- Inf
      for (k in i:(j - 1)) {
        cost <- m[i, k] + m[k + 1, j] + dims[i] * dims[k + 1] * dims[j + 1]
        if (cost < m[i, j]) {
          m[i, j] <- cost
          # Remember the split so we can rebuild the execution plan later.
          s[i, j] <- k
        }
      }
    }
  }
  
  list(cost = m, splits = s, objective_value = m[1, n])
}

# Helper: convert split matrix to RPN vector
build_rpn_from_matrix <- function(s, i = 1, j = nrow(s)) {
  if (i == j) return(i)
  k <- s[i, j]
  c(build_rpn_from_matrix(s, i, k), build_rpn_from_matrix(s, k + 1, j), -1L)
}

dp_r <- mcm_dp_r(dims)
rpn_from_dp <- build_rpn_from_matrix(dp_r$splits)

tibble(
  Source = c("Brute-force (Task 1)", "DP (Task 4)"),
  `Optimal ops` = c(min(paren_df$ops), dp_r$objective_value),
  `RPN vector` = c(paste(rpn_optimal, collapse = ", "),
                    paste(rpn_from_dp, collapse = ", "))
) |> kable(format.args = list(big.mark = ","), align = "lrc")
```

The DP gives exactly the same answer as our brute-force enumeration: **93,840 operations**, with the same RPN vector `[1, 2, 3, -1, 4, 5, -1, -1, -1]`. This is expected for $n = 5$, but the DP runs in $O(n^3)$ instead of $O(C_{n-1})$, so it scales to thousands of matrices where brute force is hopeless.

### Three C++ implementations

Now we implement three progressively more sophisticated versions in C++:

1. **Standard DP**: direct translation of the Godbole algorithm with flat arrays for cache-friendly access.
2. **Parallel wavefront DP**: same algorithm, but the inner loop over starting index $i$ is parallelized with OpenMP for each diagonal $L$. All cells on the same diagonal are independent (they depend only on shorter sub-chains already computed on previous diagonals), so they can be computed simultaneously across CPU cores.
3. **Resource-aware DP**: inspired by Nissim, Schwartz & Shabo (2024), this variant replaces the pure FLOP cost with a penalized cost that accounts for **memory hierarchy effects**. When an intermediate matrix exceeds a cache size threshold, the cost is multiplied by a penalty factor. The intuition: even if an ordering has fewer FLOPs, if it creates intermediates that spill out of L2/L3 cache, the resulting cache thrashing may make it slower in practice than an ordering with slightly more FLOPs but better data locality.

The important distinction is that the first two algorithms optimize the **same mathematical objective** as the textbook problem: scalar multiplication count. The third optimizes a **different surrogate objective**: scalar work plus hardware penalties. That is why later on we will compare both the recovered RPN vector and the objective value, rather than calling everything "minimum operations".

To isolate the OpenMP contribution from the algorithmic improvements, I check whether the report was compiled with OpenMP enabled and benchmark the wavefront code with one thread versus several threads on a larger optimizer-only case.

```{Rcpp}
//| label: task4-dp-cpp

// [[Rcpp::depends(RcppArmadillo)]]
#define ARMA_NO_DEBUG
#include <RcppArmadillo.h>
#include <vector>
#include <stack>
#include <limits>
#ifdef _OPENMP
  #include <omp.h>
#endif
using namespace Rcpp;

// Helper: reconstruct RPN from flat split table -- ITERATIVE version.
// Avoids deep recursion that could overflow the call stack for large n
// (e.g. n > 10,000 in genomic simulations). Uses an explicit work-stack
// that simulates post-order traversal with a state machine.
void build_rpn_iterative(int root_i, int root_j,
                         const std::vector<int>& s, int n,
                         std::vector<int>& rpn) {
  rpn.clear();
  // State machine: 0 = enter node, 1 = left child done, 2 = right child done
  struct Frame { int i; int j; int state; };
  std::stack<Frame> stk;
  stk.push({root_i, root_j, 0});
  while (!stk.empty()) {
    Frame& f = stk.top();
    if (f.i == f.j) {
      rpn.push_back(f.i + 1);    // 1-based for R
      stk.pop();
    } else if (f.state == 0) {
      int k = s[f.i * n + f.j];
      f.state = 1;
      stk.push({f.i, k, 0});     // push left subtree
    } else if (f.state == 1) {
      int k = s[f.i * n + f.j];
      f.state = 2;
      stk.push({k + 1, f.j, 0}); // push right subtree
    } else {
      rpn.push_back(-1);          // post-order: emit multiply after children
      stk.pop();
    }
  }
}

// 1. Standard O(n^3) DP with flat arrays
//    Uses `double` for costs (safe up to 2^53) and `long long` for
//    intermediate size calculations to avoid int32 overflow when
//    dimensions exceed ~46,000.
// [[Rcpp::export]]
List mcm_dp_cpp(const IntegerVector& dims) {
  int n = dims.size() - 1;
  std::vector<double> m(n * n, 0.0);
  std::vector<int> s(n * n, 0);
  
  for (int len = 2; len <= n; len++) {
    for (int i = 0; i < n - len + 1; i++) {
      int j = i + len - 1;
      // Flat index: cell (i, j) lives at i * n + j.
      m[i * n + j] = std::numeric_limits<double>::infinity();
      for (int k = i; k < j; k++) {
        double cost = m[i * n + k] + m[(k + 1) * n + j] +
          (double)dims[i] * (double)dims[k + 1] * (double)dims[j + 1];
        if (cost < m[i * n + j]) {
          m[i * n + j] = cost;
          s[i * n + j] = k;
        }
      }
    }
  }
  
  std::vector<int> rpn;
  rpn.reserve(2 * n - 1);
  build_rpn_iterative(0, n - 1, s, n, rpn);
  
  return List::create(
    Named("objective_value") = m[0 * n + (n - 1)],
    Named("rpn") = IntegerVector(rpn.begin(), rpn.end())
  );
}

// Report whether this build actually has OpenMP enabled.
// This keeps the notebook honest: a "parallel" code path compiled without
// OpenMP is just sequential code with ignored pragmas.
// [[Rcpp::export]]
List openmp_status_cpp() {
  #ifdef _OPENMP
  return List::create(
    Named("enabled") = true,
    Named("max_threads") = omp_get_max_threads(),
    Named("version") = _OPENMP
  );
  #else
  return List::create(
    Named("enabled") = false,
    Named("max_threads") = 1,
    Named("version") = NA_INTEGER
  );
  #endif
}

// 2. Parallel wavefront DP (OpenMP on each diagonal)
//    `min_cells_parallel` avoids spawning a parallel region for tiny diagonals,
//    where the OpenMP overhead would cost more than the work itself.
// [[Rcpp::export]]
List mcm_dp_parallel(const IntegerVector& dims,
                     int n_threads = 0,
                     int min_cells_parallel = 64) {
  int n = dims.size() - 1;
  std::vector<double> m(n * n, 0.0);
  std::vector<int> s(n * n, 0);
  
  #ifdef _OPENMP
  if (n_threads > 0) omp_set_num_threads(n_threads);
  #endif
  
  for (int len = 2; len <= n; len++) {
    int num_cells = n - len + 1;
    // Every cell on the same diagonal has the same amount of work, so a
    // static schedule keeps the runtime overhead low and the code simple.
    #pragma omp parallel for schedule(static) if(num_cells >= min_cells_parallel)
    for (int i = 0; i < num_cells; i++) {
      int j = i + len - 1;
      double best = std::numeric_limits<double>::infinity();
      int best_k = i;
      for (int k = i; k < j; k++) {
        double cost = m[i * n + k] + m[(k + 1) * n + j] +
          (double)dims[i] * (double)dims[k + 1] * (double)dims[j + 1];
        if (cost < best) {
          best = cost;
          best_k = k;
        }
      }
      m[i * n + j] = best;
      s[i * n + j] = best_k;
    }
  }
  
  std::vector<int> rpn;
  rpn.reserve(2 * n - 1);
  build_rpn_iterative(0, n - 1, s, n, rpn);
  
  return List::create(
    Named("objective_value") = m[0 * n + (n - 1)],
    Named("rpn") = IntegerVector(rpn.begin(), rpn.end())
  );
}

// 3. Resource-aware DP: penalizes intermediates that exceed cache
//    Inspired by the IO-complexity model of Hong & Kung (1981) and
//    adapted from the communication-cost framework of Nissim et al. (2024).
//    Uses a step-function penalty across two cache tiers (L2 and L3),
//    plus a fixed SIMD overhead term that penalizes very small
//    multiplications where BLAS call overhead and poor vectorization
//    dominate the actual computation.
// [[Rcpp::export]]
List mcm_dp_resource_aware(const IntegerVector& dims,
                            double l2_bytes     = 262144,
                            double l3_bytes     = 8388608,
                            double l2_penalty   = 1.5,
                            double l3_penalty   = 3.0,
                            double simd_overhead = 500.0) {
  int n = dims.size() - 1;
  std::vector<double> m(n * n, 0.0);
  std::vector<int> s(n * n, 0);
  
  for (int len = 2; len <= n; len++) {
    for (int i = 0; i < n - len + 1; i++) {
      int j = i + len - 1;
      double best = std::numeric_limits<double>::infinity();
      int best_k = i;
      for (int k = i; k < j; k++) {
        double flops = (double)dims[i] * (double)dims[k + 1] * (double)dims[j + 1];
        
        // Two-tier cache penalty: L2 and L3
        auto tier_penalty = [&](double bytes) -> double {
          if (bytes > l3_bytes) return l3_penalty;
          if (bytes > l2_bytes) return l2_penalty;
          return 1.0;
        };
        
        double result_bytes = (double)dims[i] * (double)dims[j + 1] * 8.0;
        double left_bytes   = (double)dims[i] * (double)dims[k + 1] * 8.0;
        double right_bytes  = (double)dims[k + 1] * (double)dims[j + 1] * 8.0;
        
        double penalty = tier_penalty(result_bytes)
                       * tier_penalty(left_bytes)
                       * tier_penalty(right_bytes);
        
        // Fixed SIMD overhead: each matrix multiply pays a constant cost
        // for the BLAS dgemm call setup, register spills, and poor
        // vectorization on tiny operands. This teaches the optimizer
        // to prefer fewer, larger multiplications over many tiny ones.
        double cost = m[i * n + k] + m[(k + 1) * n + j]
                    + flops * penalty + simd_overhead;
        if (cost < best) {
          best = cost;
          best_k = k;
        }
      }
      m[i * n + j] = best;
      s[i * n + j] = best_k;
    }
  }
  
  std::vector<int> rpn;
  rpn.reserve(2 * n - 1);
  build_rpn_iterative(0, n - 1, s, n, rpn);
  
  return List::create(
    Named("objective_value") = m[0 * n + (n - 1)],
    Named("rpn") = IntegerVector(rpn.begin(), rpn.end())
  );
}
```

A few implementation notes:

- **Flat arrays**: all three variants store the $m$ and $s$ tables as `std::vector<double>` indexed by `i * n + j` rather than nested vectors or Armadillo matrices. This gives contiguous memory layout and better cache behavior during the triple-nested-loop traversal. The cost entries use `double` (safe up to $2^{53}$), and all dimension products are explicitly cast to `double` before multiplication to avoid `int32` overflow when dimensions exceed ~46,000.
- **Iterative RPN reconstruction**: instead of the textbook recursive approach, we use an explicit `std::stack<Frame>` that simulates post-order traversal iteratively. Recursive reconstruction consumes one call-stack frame per level of the expression tree; for chains of $n > 10{,}000$ matrices (common in auto-generated genomic or simulation pipelines), this would exhaust the OS call stack (~8 MB on Linux/macOS) and segfault. The iterative version moves this bookkeeping to heap-allocated memory, which is effectively unlimited.
- **`#pragma omp parallel for`**: the parallel variant annotates the loop over starting index `i` for each chain length `len`. Since all cells on the same diagonal are independent, this is safe without any synchronization. I use `schedule(static)` because every cell on a given diagonal performs the same number of split evaluations. The `min_cells_parallel` guard avoids paying thread-management overhead on tiny diagonals, while still allowing the report to exercise the real threaded path on medium-sized chains.
- **Two-tier cache penalty + SIMD overhead**: the resource-aware variant uses a **step function** with two tiers (L2 at 256 KB, penalty 1.5x; L3 at 8 MB, penalty 3x), plus a **fixed per-multiplication overhead** (`simd_overhead`, default 500 pseudo-FLOPs). The tiered penalty models memory hierarchy effects following the IO-complexity framework of Hong & Kung (1981). The SIMD overhead term models a different physical reality: on modern CPUs with AVX2/AVX-512, very small matrix multiplications (e.g., $4 \times 4$) underutilize the vector registers and pay a disproportionate cost for the BLAS `dgemm` call setup, pipeline flush, and register spills. Adding a constant per split teaches the optimizer to prefer fewer, larger multiplications over many tiny ones, which is exactly how BLAS performs best.

### Parallel configuration in this render

```{r}
#| label: task4-openmp-status

openmp_info <- openmp_status_cpp()
parallel_threads <- if (isTRUE(openmp_info$enabled)) {
  min(4L, as.integer(openmp_info$max_threads))
} else {
  1L
}
parallel_min_cells <- 64L

tibble(
  `OpenMP enabled` = if (isTRUE(openmp_info$enabled)) "Yes" else "No",
  `Threads used` = parallel_threads,
  `Activation threshold` = paste(parallel_min_cells, "cells per diagonal"),
  `Recommended render path` = if (isTRUE(openmp_info$enabled)) "Current build is fine" else "Use Docker"
) |> kable(
  caption = "Parallel configuration used in this render",
  align = "lccc"
)
```

`r if (isTRUE(openmp_info$enabled)) { "OpenMP is active in this render: the parallel benchmarks below reflect real multi-threaded execution. The Docker image is provided as an alternative reproducibility path for environments where libomp is not available." } else { "If OpenMP enabled is No, the wavefront implementation still returns the correct answer, but the pragmas are ignored and the code runs sequentially. To reproduce the threaded benchmark, build and run the provided Docker image: docker build -t mcop . && docker run -v \\\"$(pwd)/output:/output\\\" mcop" }`

### Verification: same ordering, different objectives

```{r}
#| label: task4-verify

r1 <- mcm_dp_cpp(dims)
r2 <- mcm_dp_parallel(
  dims,
  n_threads = parallel_threads,
  min_cells_parallel = parallel_min_cells
)
r3 <- mcm_dp_resource_aware(dims)  # uses L2=256KB, L3=8MB defaults

tibble(
  Algorithm = c("Standard DP", "Parallel wavefront DP", "Resource-aware DP"),
  `Objective optimized` = c("Scalar ops", "Scalar ops", "Penalized cost"),
  `Objective value` = c(r1$objective_value, r2$objective_value, r3$objective_value),
  `RPN vector` = c(
    paste(r1$rpn, collapse = ", "),
    paste(r2$rpn, collapse = ", "),
    paste(r3$rpn, collapse = ", ")
  ),
  `Matches brute-force?` = c(
    identical(r1$rpn, as.integer(rpn_optimal)),
    identical(r2$rpn, as.integer(rpn_optimal)),
    identical(r3$rpn, as.integer(rpn_optimal))
  )
) |> kable(format.args = list(big.mark = ","), align = "llccc")
```

All three recover the same RPN vector as our brute-force enumeration from Task 1, so they agree on the **ordering**. The difference is in what the reported objective means. For the standard and parallel versions it is the textbook scalar-operation count. For the resource-aware version it is a penalized surrogate cost, so the numeric value is not directly comparable to "operations" even when the chosen ordering is the same. For our bottleneck-pattern chain, the FLOP-optimal ordering also happens to be cache-friendly, so the resource-aware penalties do not alter the best plan.

The resource-aware approach produces different orderings on chains with more uniform, large dimensions, where multiple near-equal-cost parenthesizations exist but differ in their memory footprint.

## The grand benchmark: everything together

Now we bring everything from Tasks 1--4 together in a single benchmark. For the Task 4 entries, we measure the **full pipeline**: optimizer + executor, so the cost of finding the optimal ordering is included.

```{r}
#| label: task4-grand-benchmark

invisible(eval(naive_expr))
invisible(eval(best_expr))
invisible(naive_multiply_cpp(mat_list))
invisible(execute_rpn(mat_list, rpn_optimal))
invisible(execute_rpn(mat_list, r1$rpn))
invisible(execute_rpn(mat_list, r2$rpn))
invisible(execute_rpn(mat_list, r3$rpn))
gc()

bm_grand <- bench::mark(
  `R naive`                   = eval(naive_expr),
  `R optimal (hand)`          = eval(best_expr),
  `C++ naive (Task 2)`        = naive_multiply_cpp(mat_list),
  `C++ RPN optimal (Task 3)`  = execute_rpn(mat_list, rpn_optimal),
  `R DP + execute`            = {
    dp <- mcm_dp_r(dims)
    rpn <- build_rpn_from_matrix(dp$splits)
    execute_rpn(mat_list, rpn)
  },
  `C++ DP + execute`          = {
    dp <- mcm_dp_cpp(dims)
    execute_rpn(mat_list, dp$rpn)
  },
  `C++ parallel DP + execute` = {
    dp <- mcm_dp_parallel(
      dims,
      n_threads = parallel_threads,
      min_cells_parallel = parallel_min_cells
    )
    execute_rpn(mat_list, dp$rpn)
  },
  `C++ resource-aware + exec` = {
    dp <- mcm_dp_resource_aware(dims)
    execute_rpn(mat_list, dp$rpn)
  },
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm_grand |>
  mutate(
    expression = as.character(expression),
    `CV (sd/median)` = round(bench_cv(bm_grand), 3)
  ) |>
  select(expression, min, median, `CV (sd/median)`, mem_alloc, n_itr) |>
  kable(
    caption = "Grand benchmark: all approaches from Tasks 1--4",
    align = "lccccc"
  )
```

The results tell a clear story. The exact medians should be read from the table because they depend on the current machine, but the qualitative pattern is robust: all the "DP + execute" pipelines stay very close to the hand-optimized baselines, while the naive approaches remain far slower. In other words, for a chain this short, **finding the optimal plan is cheap enough that almost all the runtime is still spent in the matrix multiplications themselves**. It is also expected that `R optimal (hand)` and `C++ DP + execute` land in the same region: once the ordering is fixed, both dispatch the same sequence of BLAS calls on the same data, so the runtime difference reduces to the small overhead of the RPN executor and the DP optimizer — consistent with what Task 2 already showed for the naive case.

### Decomposition: optimize vs. execute

One natural concern with the grand benchmark is that it conflates two distinct costs: finding the optimal ordering and executing it. In many real-world scenarios (e.g., training a neural network), you optimize the chain **once** and execute it **millions of times** with different data. The next benchmark separates the two to confirm that the DP overhead is not distorting the comparison:

```{r}
#| label: task4-decompose

invisible(mcm_dp_cpp(dims))
invisible(execute_rpn(mat_list, r1$rpn))
gc()

bm_decompose <- bench::mark(
  `C++ DP (optimize only)` = mcm_dp_cpp(dims),
  `C++ execute (pre-computed RPN)` = execute_rpn(mat_list, r1$rpn),
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm_decompose |>
  mutate(
    expression = as.character(expression),
    `CV (sd/median)` = round(bench_cv(bm_decompose), 3)
  ) |>
  select(expression, min, median, `CV (sd/median)`, mem_alloc) |>
  kable(caption = "Decomposed: optimization vs. execution (n = 5)",
        align = "lcccc")
```

For $n = 5$, the optimization-only benchmark is tiny compared with execution. In this render it is `r fmt_time(bench_value(bm_decompose, "C++ DP (optimize only)", "median"))` versus `r fmt_time(bench_value(bm_decompose, "C++ execute (pre-computed RPN)", "median"))`, so optimization is only about `r round(100 * as.numeric(bench_value(bm_decompose, "C++ DP (optimize only)", "median")) / as.numeric(bench_value(bm_decompose, "C++ execute (pre-computed RPN)", "median")), 1)`% of the execution cost. That makes the plan-computation overhead essentially negligible when the same multiplication pattern is reused many times.

```{r}
#| label: task4-grand-plot
#| fig-cap: "Grand benchmark across all tasks. The DP optimization overhead is invisible at this scale."
#| fig-height: 7

bm_plot <- bm_grand |>
  mutate(
    expression = as.character(expression),
    median_us  = as.numeric(median) * 1e6,
    task = case_when(
      grepl("Task 2", expression) ~ "Task 2",
      grepl("Task 3", expression) ~ "Task 3",
      grepl("DP|resource|parallel", expression) ~ "Task 4",
      TRUE ~ "Task 1"
    )
  )

ggplot(bm_plot, aes(x = reorder(expression, median_us),
                     y = median_us, fill = task)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = if_else(
    median_us > 1000,
    paste0(round(median_us / 1000, 1), " ms"),
    paste0(round(median_us), " \u00b5s")
  )), hjust = -0.1, size = 3.3, fontface = "bold") +
  coord_flip(clip = "off") +
  scale_fill_manual(
    values = c("Task 1" = palette_main[3], "Task 2" = palette_main[4],
               "Task 3" = palette_main[2], "Task 4" = palette_main[1]),
    name = "Task"
  ) +
  scale_y_log10(
    labels = function(x) if_else(x >= 1000, paste0(x/1000, " ms"), paste0(x, " \u00b5s")),
    expand = expansion(mult = c(0, 0.35))
  ) +
  labs(
    title    = "The Full Picture: Tasks 1 through 4",
    subtitle = "Log scale. Automatic optimization (Task 4) matches hand-tuned performance.",
    x = NULL, y = "Median execution time (log scale)"
  )
```

## How fast is the optimizer itself? Scaling to larger chains

For our 5-matrix chain, the DP overhead is negligible. But what happens as $n$ grows? This matters because in real applications (e.g., deep learning, scientific simulations), chains of 20, 50, or even hundreds of matrices are common.

We benchmark the four DP implementations (R, C++ standard, C++ parallel, C++ resource-aware) on chains of increasing length with randomly generated bottleneck dimensions. Only the **dimension vectors** matter here; the optimizer does not need actual matrix entries, so this section isolates planning cost cleanly.

```{r}
#| label: task4-scaling

set.seed(123)
scaling_results <- list()

for (n in c(5, 10, 20, 50, 100, 200)) {
  # Random bottleneck dimensions: alternate large/small
  test_dims <- integer(n + 1)
  for (i in seq_along(test_dims)) {
    test_dims[i] <- if (i %% 2 == 1) sample(100:1000, 1) else sample(2:20, 1)
  }
  
  bm_s <- bench::mark(
    `R DP`               = mcm_dp_r(test_dims),
    `C++ DP`             = mcm_dp_cpp(test_dims),
    `C++ parallel DP`    = mcm_dp_parallel(
      test_dims,
      n_threads = parallel_threads,
      min_cells_parallel = parallel_min_cells
    ),
    `C++ resource-aware` = mcm_dp_resource_aware(test_dims),
    check = FALSE,
    min_time = 2,
    min_iterations = 50,
    filter_gc = FALSE,
    max_iterations = 100
  )
  
  scaling_results[[as.character(n)]] <- bm_s |>
    mutate(
      expression = as.character(expression),
      n = n,
      sd_time = vapply(time, function(x) sd(as.numeric(x)), numeric(1))
    ) |>
    select(expression, n, median, sd_time)
}

scale_df <- bind_rows(scaling_results) |>
  mutate(
    median_us = as.numeric(median) * 1e6,
    cv = as.numeric(sd_time) / as.numeric(median)
  )
```

```{r}
#| label: task4-scaling-table

scale_table <- scale_df |>
  mutate(time_label = case_when(
    median_us < 1000 ~ paste0(round(median_us, 1), " \u00b5s"),
    median_us < 1e6  ~ paste0(round(median_us / 1000, 2), " ms"),
    TRUE             ~ paste0(round(median_us / 1e6, 2), " s")
  )) |>
  select(expression, n, time_label) |>
  pivot_wider(names_from = expression, values_from = time_label)

scale_cv_table <- scale_df |>
  mutate(cv_label = round(cv, 3)) |>
  select(expression, n, cv_label) |>
  pivot_wider(names_from = expression, values_from = cv_label) |>
  rename_with(~ paste0(.x, " CV"), -n)

scale_table <- scale_table |>
  left_join(scale_cv_table, by = "n")

# Dynamic crossover detection: find where parallel first beats sequential C++ DP
crossover_df <- scale_df |>
  filter(expression %in% c("C++ DP", "C++ parallel DP")) |>
  select(n, expression, median_us) |>
  pivot_wider(names_from = expression, values_from = median_us) |>
  rename(seq_us = `C++ DP`, par_us = `C++ parallel DP`) |>
  arrange(n) |>
  mutate(par_wins = par_us < seq_us)

# Count sign flips to detect monotone vs non-monotone crossover
flips <- crossover_df |>
  mutate(prev_par_wins = lag(par_wins)) |>
  filter(!is.na(prev_par_wins), par_wins != prev_par_wins)

n_flips <- nrow(flips)

if (n_flips == 0) {
  # parallel never beats sequential (or always beats — check which)
  crossover_case <- if (all(!crossover_df$par_wins)) "never" else "always"
} else if (n_flips == 1) {
  crossover_case <- "clean"
  cross_n_hi   <- flips$n[1]
  cross_n_lo   <- crossover_df$n[which(crossover_df$n == cross_n_hi) - 1]
  cross_seq_lo <- crossover_df$seq_us[crossover_df$n == cross_n_lo]
  cross_par_lo <- crossover_df$par_us[crossover_df$n == cross_n_lo]
  cross_seq_hi <- crossover_df$seq_us[crossover_df$n == cross_n_hi]
  cross_par_hi <- crossover_df$par_us[crossover_df$n == cross_n_hi]
} else {
  crossover_case <- "nonmonotone"
  # summarise: which n values does parallel actually win?
  par_wins_at <- crossover_df$n[crossover_df$par_wins]
  par_loses_at <- crossover_df$n[!crossover_df$par_wins]
}

 parallel_crossover_text <- if (crossover_case == "clean") {
  paste0(
    "In this benchmark a clean crossover appears between $n = ", cross_n_lo, "$ and $n = ", cross_n_hi,
    "$: at $n = ", cross_n_lo, "$ the parallel variant is still slower (",
    fmt_time(cross_par_lo * 1e-6), " vs ", fmt_time(cross_seq_lo * 1e-6), " for sequential), and from $n = ",
    cross_n_hi, "$ onward it consistently pulls ahead (", fmt_time(cross_par_hi * 1e-6), " vs ",
    fmt_time(cross_seq_hi * 1e-6), " for sequential)."
  )
} else if (crossover_case == "nonmonotone") {
  paste0(
    "In this benchmark the advantage is non-monotone: parallel wins at $n \\in \\{",
    paste(par_wins_at, collapse = ", "), "\\}$ but loses again at $n \\in \\{",
    paste(par_loses_at, collapse = ", "),
    "\\}$. This reflects the sensitivity of OpenMP thread scheduling to diagonal width on this hardware — the kernel is too light per cell to amortize thread overhead consistently at intermediate chain lengths."
  )
} else if (crossover_case == "never") {
  paste0(
    "Within the measured range up to $n = ", max(crossover_df$n),
    "$, the parallel variant does not outperform the sequential C++ DP at any measured point — diagonals remain too narrow to offset thread overhead."
  )
} else {
  "The parallel variant is faster than sequential C++ DP across all measured $n$ in this render."
}

scale_table |>
  kable(caption = "DP optimization time and CV by chain length and algorithm",
        col.names = c("n", "R DP", "C++ DP", "C++ parallel", "C++ resource-aware",
                      "R DP CV", "C++ DP CV", "C++ parallel CV", "C++ resource-aware CV"),
        align = "ccccccccc")
```

At $n = 5$, all variants are effectively free. But the gap widens quickly:

- By $n = 200$, the table shows that the compiled DP implementations are much faster than the R version. In this render, `R DP` takes `r scale_table[scale_table$n == 200, "R DP"][[1]]` while `C++ DP` takes `r scale_table[scale_table$n == 200, "C++ DP"][[1]]`, which is about a `r fmt_ratio(scale_df$median_us[scale_df$expression == "R DP" & scale_df$n == 200] / scale_df$median_us[scale_df$expression == "C++ DP" & scale_df$n == 200])` speedup.
- The parallel wavefront version now uses the same OpenMP configuration reported above. At small $n$, thread-management overhead makes it slower than the sequential C++ DP: each diagonal is narrow, so the work available per thread is tiny. `r parallel_crossover_text` The scaling table does not extend beyond $n = `r max(scale_df$n)`$, so the exact behaviour for larger chains is not measured here.
- The resource-aware variant is consistently slower than standard C++ DP because it evaluates a richer objective function at every candidate split. That extra work is the price of modeling hardware effects rather than counting FLOPs alone.

### A direct parallel benchmark

The scaling table mixes several effects at once: language, objective function, and thread count. To isolate the OpenMP effect, I benchmark the **same wavefront implementation** twice on a larger optimizer-only problem: once with one thread, once with the thread count selected above. The only thing that changes is the number of threads.

```{r}
#| label: task4-parallel-demo

set.seed(2026)
parallel_demo_n <- 320L
parallel_demo_dims <- integer(parallel_demo_n + 1L)

for (i in seq_along(parallel_demo_dims)) {
  parallel_demo_dims[i] <- if (i %% 2 == 1) sample(150:1200, 1) else sample(2:30, 1)
}

if (isTRUE(openmp_info$enabled) && parallel_threads > 1L) {
  invisible(mcm_dp_parallel(
    parallel_demo_dims,
    n_threads = 1L,
    min_cells_parallel = parallel_min_cells
  ))
  invisible(mcm_dp_parallel(
    parallel_demo_dims,
    n_threads = parallel_threads,
    min_cells_parallel = parallel_min_cells
  ))
  gc()

  bm_parallel_demo <- bench::mark(
    `Wavefront DP (1 thread)` = mcm_dp_parallel(
      parallel_demo_dims,
      n_threads = 1L,
      min_cells_parallel = parallel_min_cells
    ),
    `Wavefront DP (parallel)` = mcm_dp_parallel(
      parallel_demo_dims,
      n_threads = parallel_threads,
      min_cells_parallel = parallel_min_cells
    ),
    check = FALSE,
    min_time = 2,
    min_iterations = 50,
    filter_gc = FALSE,
    max_iterations = 100
  )

  bm_parallel_demo |>
    mutate(
      expression = as.character(expression),
      `CV (sd/median)` = round(bench_cv(bm_parallel_demo), 3)
    ) |>
    select(expression, min, median, `CV (sd/median)`, mem_alloc, n_itr) |>
    kable(
      caption = "Direct OpenMP benchmark: same wavefront DP, one thread vs parallel",
      align = "lccccc"
    )
} else {
  tibble(
    Message = "OpenMP is not enabled in this render, so the wavefront benchmark falls back to one thread. Use Docker to reproduce the threaded benchmark."
  ) |>
    kable(caption = "Direct OpenMP benchmark status", align = "l")
}
```

This benchmark is intentionally optimizer-only. The point is not to claim that OpenMP changes the end-to-end Task 4 result for a 5-matrix chain; the point is to show that the wavefront code really can fan out across cores once the diagonals are wide enough. The speedup is modest rather than dramatic, which is expected: the kernel is doing lightweight scalar DP updates, not large BLAS calls.

```{r}
#| label: task4-scaling-plot
#| fig-cap: "DP optimization time vs chain length. Both axes are log-scaled, showing the expected cubic growth."

ggplot(scale_df, aes(x = n, y = median_us, color = expression)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_log10() +
  scale_y_log10(labels = function(x) {
    case_when(
      x < 1000 ~ paste0(round(x), " \u00b5s"),
      x < 1e6  ~ paste0(round(x / 1000, 1), " ms"),
      TRUE     ~ paste0(round(x / 1e6, 2), " s")
    )
  }) +
  scale_color_manual(values = c(
    "R DP"               = palette_main[3],
    "C++ DP"             = palette_main[1],
    "C++ parallel DP"    = palette_main[2],
    "C++ resource-aware" = palette_main[4]
  ), name = NULL) +
  labs(
    title    = "Optimizer Scaling: O(n\u00b3) in Action",
    subtitle = "Log-log plot. The cubic growth is evident; C++ gains a constant-factor advantage.",
    x = "Chain length (n)", y = "Median optimization time"
  )
```

The log-log plot shows all four lines growing with roughly the same slope (the $n^3$ exponent), separated by constant factors. The R implementation sits one order of magnitude above the C++ variants, and the resource-aware variant sits slightly above the standard C++ DP. The parallel curve stays close to the standard one because this DP kernel is relatively light per cell; the dedicated benchmark above is the cleaner way to see the effect of extra threads.

## Beyond Godbole: a brief note on $O(n \log n)$ methods

For completeness, it is worth mentioning that the $O(n^3)$ dynamic programming approach is not the theoretical limit. Hu and Shing (1982, 1984) showed that the MCOP is isomorphic to the **minimum-weight triangulation of a convex polygon**, and exploited this geometric structure to achieve $O(n \log n)$ time. The key idea is that each parenthesization corresponds to a triangulation: the $n$ matrices map to the sides of an $(n+1)$-sided polygon, and each multiplication corresponds to a triangle with cost equal to the product of its three vertex weights.

However, the Hu-Shing algorithm is significantly harder to implement correctly (it requires circular linked lists, specialized stacks, and careful handling of degenerate cases), and its constant factors are large enough that it only outperforms the $O(n^3)$ DP for $n > 1000$ or so. Schwartz and Weiss (2019) even found a flaw in the original correctness proof, which required a nontrivial fix. For our purposes ($n = 5$ to $n = 200$), the cubic DP is more than fast enough, and the C++ implementation keeps us comfortably under a second even at $n = 200$.

## Discussion

Task 4 brings together all the pieces:

1. **The optimizer finds the optimal ordering automatically.** The $O(n^3)$ DP produces the same RPN vector as our brute-force enumeration, but scales to chains of any practical size. The scaling table shows that even when the problem size grows to $n = 200$, the compiled DP remains comfortably fast.

2. **The full pipeline (optimize + execute) is as fast as hand-tuned code for this example.** In the grand benchmark, the automated C++ approaches stay in the same performance region as the hand-written optimal baselines. That is exactly what we wanted from the assignment: automation without sacrificing the main performance gain.

3. **Compilation matters more than parallelization** at this scale. The largest constant-factor improvement comes from moving the optimizer from R to C++ and storing the DP tables in contiguous arrays. The direct OpenMP benchmark confirms a real speedup, but it stays modest because each DP cell update is tiny compared with the cost profile of dense matrix multiplication.

4. **The resource-aware cost function models real hardware, not just FLOPs.** It accounts for two physical effects that a pure FLOP counter ignores: (a) cache-tier penalties that approximate memory bandwidth costs when intermediates spill out of L2 or L3, and (b) a fixed SIMD overhead per multiplication that penalizes tiny operands where BLAS call setup and poor vectorization dominate. For our bottleneck-pattern chain, these corrections do not change the optimal ordering because it is already cache-friendly and involves few multiplications. The value emerges on chains with more uniform, large dimensions, where multiple near-equal-cost orderings exist and the tiebreaker is memory locality or the number of BLAS calls.

The overall conclusion across all four tasks:

```{r}
#| label: task4-overall-summary

grand_speedup <- function(label) {
  as.numeric(bench_value(bm_grand, "R naive", "median")) /
    as.numeric(bench_value(bm_grand, label, "median"))
}

overall_summary <- tibble(
  `What we did` = c(
    "Hand-parenthesized R (Task 1)",
    "C++ with same naive ordering (Task 2)",
    "C++ stack machine + pre-computed optimal (Task 3)",
    "C++ DP optimizer + stack machine (Task 4)"
  ),
  `Relative to R naive` = c(
    fmt_relative_perf(grand_speedup("R optimal (hand)")),
    fmt_relative_perf(grand_speedup("C++ naive (Task 2)")),
    fmt_relative_perf(grand_speedup("C++ RPN optimal (Task 3)")),
    fmt_relative_perf(grand_speedup("C++ DP + execute"))
  )
)

overall_summary |>
  kable(align = "lr", caption = "Overall relative performance derived from the grand benchmark")
```

:::{.callout-note}
## Sanity check on the results

The benchmark patterns are internally consistent with the theory developed in the report:

1. Whenever two strategies are compared **inside the same benchmark table**, the good parenthesization beats the naive one by roughly the factor suggested by the scalar-operation counts.
2. The naive R and naive C++ versions stay in the same broad performance region, which is expected because both ultimately spend most of their time inside the same BLAS kernels once the multiplication order is fixed.
3. In the optimization-only microbenchmark, planning takes microseconds while executing dense matrix products takes hundreds of microseconds to milliseconds, so the optimizer is not the bottleneck for the problem sizes studied here.

That is why the cross-section timing drift is acceptable: the absolute medians move a bit across benchmark sessions, but the qualitative conclusions remain stable.
:::

```{r}
#| label: task4-t3-vs-t4-text
#| include: false

t3 <- as.numeric(bench_value(bm_grand, "C++ RPN optimal (Task 3)", "median"))
t4 <- as.numeric(bench_value(bm_grand, "C++ DP + execute", "median"))
dp <- as.numeric(bench_value(bm_decompose, "C++ DP (optimize only)", "median"))

t3_vs_t4_text <- if (t3 < t4) {
  paste0(
    "The same logic explains why `C++ RPN optimal (Task 3)` (", fmt_time(t3),
    ") edges out `C++ DP + execute` (", fmt_time(t4),
    ") in the overall summary table: Task 3 passes a pre-computed RPN vector with zero optimization cost at benchmark time, while Task 4 folds in the ",
    fmt_time(dp), " DP call on top of the same executor."
  )
} else {
  paste0(
    "In this render `C++ DP + execute` (", fmt_time(t4),
    ") is marginally faster than `C++ RPN optimal (Task 3)` (", fmt_time(t3),
    "). The difference (", fmt_time(abs(t4 - t3)),
    ") is within measurement noise — both pipelines execute the same BLAS calls, and the ",
    fmt_time(dp), " DP overhead is smaller than the run-to-run jitter at these timescales."
  )
}
```

:::{.callout-note}
## Technical Note: the "Hand-Parenthesization Paradox"

A curious observation in these results is that the hand-parenthesized baseline comes out measurably ahead of the fully automated pipeline (`r fmt_time(bench_value(bm_grand, "R optimal (hand)", "median"))` vs `r fmt_time(bench_value(bm_grand, "C++ DP + execute", "median"))`) in this tiny 5-matrix example. This is not a failure of the later algorithms, but a reflection of two concrete HPC effects:

1. **The efficiency of BLAS.** R's `%*%` operator is not "slow R code"; it is a thin wrapper around highly optimized C/Fortran libraries (BLAS/LAPACK). When the multiplication order is correct, the overhead of the R interpreter becomes negligible compared to the billions of floating-point operations performed by the CPU.

2. **RPN executor overhead, not DP cost.** The decomposition benchmark shows the DP itself costs only `r fmt_time(bench_value(bm_decompose, "C++ DP (optimize only)", "median"))` — a negligible fraction of the gap. The real cost is the stack machine: for a 5-matrix chain (just 4 multiplications), each token requires a push/pop, a type conversion, and an indirect dispatch. That constant per-operation overhead of the generic executor exceeds the R interpreter overhead when BLAS dominates the runtime. The gap shrinks as the chain grows and BLAS work dwarfs executor bookkeeping. `r t3_vs_t4_text`

The real victory of Task 4 is **scalability**. While hand-tuning is feasible for 5 matrices, it is impossible for 100. The C++ implementation gets extremely close to the hand-written optimum while providing a fully automated, crash-safe, and cache-aware solution that scales to any practical chain length.
:::

The message is consistent: **algorithm first, language second**. Reordering the chain is where the large win comes from; C++ then helps us package that win into an automated and scalable implementation.

---

# Limitations and Future Improvements {.unnumbered}

No solution is complete without honest self-assessment. Here are the main limitations of this implementation and how they could be addressed in a production setting:

**Shallow-copy overhead at the R/C++ boundary.** The `execute_rpn` engine uses zero-copy `arma::mat` wrappers (via `copy_aux_mem = false`) to point directly at R's SEXP memory, avoiding the deep-copy penalty of `as<arma::mat>()`. However, when a matrix is pushed onto the execution stack, it still creates a shallow copy of the wrapper metadata (a few pointers and size fields). For truly gigabyte-scale matrices this is negligible, but an even more aggressive approach would use `Rcpp::XPtr<arma::mat>` to pass raw heap pointers between R and C++ without touching the R allocator at all.

**Thread overhead in the parallel wavefront DP.** The work inside each DP cell is still just a few scalar operations and comparisons, so this is not the kind of kernel that scales like `dgemm`. Even with the lower `min_cells_parallel = 64` threshold used in the report, the threaded version only starts to help on wider diagonals, and the speedup remains modest on laptop-sized benchmarks. A more adaptive strategy would calibrate the threshold at startup and potentially switch between sequential, OpenMP, and task-based runtimes depending on the problem size.

**Iterative RPN reconstruction.** The RPN path reconstruction uses an explicit `std::stack<Frame>` instead of recursion. This is deliberate: for chains of $n > 10{,}000$ matrices (common in auto-generated genomic or simulation pipelines), a recursive implementation would exhaust the OS call stack (~8 MB on Linux/macOS) and segfault. The iterative version moves this bookkeeping to heap memory, making it safe for arbitrarily long chains.

**Simplified cache and SIMD model.** The resource-aware DP uses a two-tier step function (L2 and L3 penalties) plus a fixed per-multiplication SIMD overhead. This captures the two dominant hardware effects (memory hierarchy and BLAS call granularity), but it remains an approximation. A faithful model would follow the **IO-complexity framework** of Hong & Kung (1981), using a continuous cost function $C = \alpha F + \beta M + \gamma D$ where $F$ is FLOPs, $M$ is data movement volume, and $D$ is cache-line evictions. Moreover, the model ignores **eviction policies**: multiplying two small matrices that fit in cache can still evict previously loaded data if the cache is shared. A full cache simulation (e.g., using the CME model of Christ et al., 2013) is out of scope for this report but would be the natural next step.

**Integer overflow risk.** The DP cost tables use `double` (safe up to $2^{53} \approx 9 \times 10^{15}$), and all dimension products are explicitly cast to `double` before multiplication. However, the dimension vector itself is stored as `int`. For matrices exceeding $46{,}340 \times 46{,}340$, the element count overflows a 32-bit integer. A production implementation would use `size_t` or `int64_t` throughout.

**Mixed precision opportunity.** All computations use `double` (64-bit). In HPC and deep learning contexts, `float` (32-bit) or even `half` (16-bit) arithmetic can double or quadruple throughput when full precision is not critical. Armadillo supports `arma::fmat` for single-precision, and the stack machine could be templated to handle both. We did not pursue this here because the focus is on the ordering problem, not precision trade-offs.

**Benchmark methodology.** The benchmarks measure warm-cache performance (repeated iterations with `bench::mark`). Because each section uses its own benchmark session, absolute medians for the same technique can drift slightly across tables. Cold-start behavior (first execution after data load) can also differ significantly due to TLB misses and cache warming. For a fully rigorous comparison, one would alternate configurations within the same benchmark loop to equalize cache state, or use hardware performance counters (`perf stat` on Linux) to measure L1/L2/L3 miss rates directly.

---

# References {.unnumbered}

1. Cormen, T. H., Leiserson, C. E., Rivest, R. L., & Stein, C. (2009). *Introduction to Algorithms* (3rd ed.). MIT Press. [Publisher page](https://mitpress.mit.edu/9780262033848/introduction-to-algorithms/)

2. Godbole, S. S. (1973). On efficient computation of matrix chain products. *IEEE Transactions on Computers, C-22*(9), 864--866. [https://doi.org/10.1109/TC.1973.5009182](https://doi.org/10.1109/TC.1973.5009182)

3. Hu, T. C., & Shing, M. T. (1982). Computation of matrix chain products. Part I. *SIAM Journal on Computing, 11*(2), 362--373. [https://doi.org/10.1137/0211028](https://doi.org/10.1137/0211028)

4. Hu, T. C., & Shing, M. T. (1984). Computation of matrix chain products. Part II. *SIAM Journal on Computing, 13*(2), 228--251. [https://doi.org/10.1137/0213017](https://doi.org/10.1137/0213017)

5. Schwartz, O., & Weiss, E. (2019). Revisiting "Computation of matrix chain products". *SIAM Journal on Computing, 48*(5), 1481--1486. [https://doi.org/10.1137/18M1195401](https://doi.org/10.1137/18M1195401)

6. Nissim, R., Schwartz, O., & Shabo, R. (2024). *Challenges in parallel matrix chain multiplication*. JSSPP 2024. [Author PDF](https://www.cs.huji.ac.il/~odedsc/papers/JSSPP24-Chain-MM.pdf)

7. Eddelbuettel, D., & Sanderson, C. (2014). RcppArmadillo: Accelerating R with high-performance C++ linear algebra. *Computational Statistics & Data Analysis, 71*, 1054--1063. [https://doi.org/10.1016/j.csda.2013.02.005](https://doi.org/10.1016/j.csda.2013.02.005). Package vignette: [RcppArmadillo introduction](https://cran.r-project.org/package=RcppArmadillo/vignettes/RcppArmadillo-intro.pdf)

8. Hong, J. W., & Kung, H. T. (1981). I/O complexity: The red-blue pebble game. *Proceedings of the 13th Annual ACM Symposium on Theory of Computing*, 326--333. [https://doi.org/10.1145/800076.802486](https://doi.org/10.1145/800076.802486)

9. Matrix chain multiplication. (2026). In *Wikipedia*. Retrieved February 14, 2026, from [https://en.wikipedia.org/wiki/Matrix_chain_multiplication](https://en.wikipedia.org/wiki/Matrix_chain_multiplication)
