---
title: "The Matrix Chain Ordering Problem"
subtitle: "High-Performance Computing Challenge"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
    fig-width: 10
    fig-height: 6
    self-contained: true
    smooth-scroll: true
execute:
  warning: true
  message: false
---

```{r}
#| label: setup
#| include: false

# Compiler flags: use an externally supplied configuration when available
# (for example, the Docker image enables OpenMP explicitly). Otherwise fall
# back to a fast local build that targets the current machine.
if (Sys.getenv("PKG_CXXFLAGS") == "") {
  Sys.setenv(PKG_CXXFLAGS = "-O3 -march=native")
}

library(Rcpp)
library(RcppArmadillo)
library(bench)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(knitr)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40", size = 12),
      panel.grid.minor = element_blank(),
      legend.position  = "top"
    )
)

palette_main <- c("#2E86AB", "#A23B72", "#F18F01", "#C73E1D", "#3B1F2B")

fmt_time <- function(x) {
  seconds <- as.numeric(x)
  if (seconds < 1e-6) return(paste0(round(seconds * 1e9, 1), " ns"))
  if (seconds < 1e-3) return(paste0(round(seconds * 1e6, 1), " \u00b5s"))
  if (seconds < 1) return(paste0(round(seconds * 1e3, 2), " ms"))
  paste0(round(seconds, 2), " s")
}

fmt_bytes_binary <- function(x) {
  bytes <- as.numeric(x)
  if (bytes < 1024) return(paste0(round(bytes, 1), " B"))
  if (bytes < 1024^2) return(paste0(round(bytes / 1024, 1), " KiB"))
  if (bytes < 1024^3) return(paste0(round(bytes / 1024^2, 2), " MiB"))
  paste0(round(bytes / 1024^3, 2), " GiB")
}

fmt_ratio <- function(x) paste0(round(x, 1), "x")

fmt_relative_perf <- function(x) {
  if (x >= 1) {
    paste0(round(x, 1), "x faster")
  } else {
    paste0(round(1 / x, 1), "x slower")
  }
}

bench_value <- function(bm, label, column) {
  idx <- which(as.character(bm$expression) == label)
  bm[[column]][idx][1]
}

bench_cv <- function(bm) {
  sd_time <- vapply(bm$time, function(x) sd(as.numeric(x)), numeric(1))
  sd_time / vapply(bm$time, function(x) mean(as.numeric(x)), numeric(1))
}

blas_info <- extSoftVersion()[["BLAS"]]
```

# Introduction

This report addresses the **Matrix Chain Ordering Problem** (MCOP) as part of the High-Performance Computing course. The problem boils down to one question: given a chain of matrices to multiply, in what order should we group the multiplications to minimize computational cost?

Matrix multiplication is **associative**, in exact arithmetic; floating-point results may differ slightly with parenthesization. But the number of scalar operations (and the size of intermediate matrices) can change wildly depending on the grouping. For $n$ matrices, the number of valid parenthesizations follows the Catalan numbers:

$$
C_{n-1} = \frac{1}{n}\binom{2(n-1)}{n-1}
$$

For $n = 5$ that gives us $C_4 = 14$ options, which is small enough to enumerate by brute force. For larger $n$, this grows exponentially and we need smarter algorithms.

The report is structured in four tasks:

1. **Task 1:** Find a compelling 5-matrix example where ordering matters, benchmark it in R.
2. **Task 2:** Implement a naive C++ multiplication using RcppArmadillo and compare.
3. **Task 3:** Build a C++ execution engine that multiplies matrices in any given order.
4. **Task 4:** Implement optimization algorithms (dynamic programming, parallelization, cache-aware heuristics) and compare everything end to end.

:::{.callout-note}
## How to read the benchmark tables

Each benchmark table in this report comes from a separate `bench::mark()` run executed during rendering. That means the **same technique can appear with slightly different absolute times in different sections**: cache state, garbage collection, BLAS thread scheduling, CPU frequency scaling, and the set of competing expressions all change from run to run.

So the safest interpretation is:

1. Compare entries **within the same table**.
2. Use medians rather than minimums.
3. Treat the **grand benchmark in Task 4** as the canonical cross-task comparison, because it places all approaches in one shared benchmark session.

Benchmarks request `min_time = 2` and at least 50 iterations. The two scaling sweeps also cap execution at 100 iterations, so fast expressions there may run for less than two seconds. Read the measured iteration counts and dispersion alongside each timing. Each benchmark table includes a `CV (sd/mean)` column so you can gauge stability at a glance; smaller values mean tighter timing dispersion. The `n_itr` column reports how many iterations were actually executed for each expression. I prefer `min_time` to a fixed `min_iterations` because a fixed iteration budget under-samples very fast operations and over-samples slow ones, but I also enforce `min_iterations = 50` to avoid single-iteration rows on slow cases. `bench::mark()` reports `n_itr` **after** optionally filtering out iterations that triggered garbage collection; to keep the floor intact, I disable GC filtering (`filter_gc = FALSE`) so the displayed `n_itr` matches the enforced minimum.

To reduce cold-start noise, the code warms up every strategy once before benchmarking. All benchmarks were produced on `r R.version$os` (`r R.version$arch`). Active BLAS: `r blas_info`.
:::

---

# Task 1: The Cost of Naive Ordering in R {#sec-task1}

The goal here is to find 5 matrices where R's default left-to-right evaluation (`A %*% B %*% C %*% D %*% E`) performs significantly worse than an optimal parenthesization. We need to design the dimensions carefully, enumerate all possible orderings, and benchmark the key cases.

## Choosing the dimensions

The trick is to alternate between **large** and **small** dimensions. This creates a "bottleneck" pattern where a naive left-to-right evaluation is forced to build huge intermediate matrices, while a smarter grouping can keep everything small by multiplying through the narrow dimensions first.

```{r}
#| label: task1-dims

# Dimension vector: A is p0×p1, B is p1×p2, ..., E is p4×p5
# We alternate large-small to create the bottleneck pattern
dims <- c(800, 5, 900, 8, 700, 6)

set.seed(42)
A <- matrix(rnorm(dims[1] * dims[2]), nrow = dims[1], ncol = dims[2])
B <- matrix(rnorm(dims[2] * dims[3]), nrow = dims[2], ncol = dims[3])
C <- matrix(rnorm(dims[3] * dims[4]), nrow = dims[3], ncol = dims[4])
D <- matrix(rnorm(dims[4] * dims[5]), nrow = dims[4], ncol = dims[5])
E <- matrix(rnorm(dims[5] * dims[6]), nrow = dims[5], ncol = dims[6])

tibble(
  Matrix     = LETTERS[1:5],
  Dimensions = paste0(dims[1:5], " × ", dims[2:6]),
  Elements   = dims[1:5] * dims[2:6]
) |> kable(
  format.args = list(big.mark = ","),
  align = "lcc"
)
```

Notice the pattern: $A$ is $800 \times 5$ (tall and skinny), $B$ is $5 \times 900$ (short and wide), $C$ is $900 \times 8$ (tall and skinny again), and so on. If R evaluates left to right, the very first step computes $AB$, which produces an $800 \times 900 = 720{,}000$ element intermediate. A smarter approach would start with $BC$ ($5 \times 8 = 40$ elements) or $DE$ ($8 \times 6 = 48$ elements), keeping intermediates tiny.

## Enumerating all 14 parenthesizations

Since $C_4 = 14$, we can just generate all of them, compute the theoretical operation count for each, and sort them. Each parenthesization is represented internally as a **split table**: a mapping from sub-chain `[i, j]` to the split index `k` where the final multiplication happens.

```{r}
#| label: task1-enumerate

# Count scalar multiplications for a given split table
count_ops <- function(dims, splits, i = 1, j = length(dims) - 1) {
  if (i == j) return(0)
  k <- splits[[paste(i, j)]]
  left  <- count_ops(dims, splits, i, k)
  right <- count_ops(dims, splits, k + 1, j)
  cost  <- dims[i] * dims[k + 1] * dims[j + 1]
  left + right + cost
}

# Generate all valid parenthesizations for n matrices
generate_all_parens <- function(n) {
  get_all <- function(i, j) {
    if (i == j) return(list(list()))
    combos <- list()
    for (k in i:(j - 1)) {
      lefts  <- get_all(i, k)
      rights <- get_all(k + 1, j)
      for (l in lefts) {
        for (r in rights) {
          combo <- c(l, r, setNames(list(k), paste(i, j)))
          combos[[length(combos) + 1]] <- combo
        }
      }
    }
    combos
  }
  raw <- get_all(1, n)
  lapply(raw, function(x) {
    splits <- list()
    for (nm in names(x)) splits[[nm]] <- x[[nm]]
    splits
  })
}

all_parens <- generate_all_parens(5)

# Build a human-readable label: (A · (B · C)) style
paren_label <- function(splits, i = 1, j = 5) {
  if (i == j) return(LETTERS[i])
  k <- splits[[paste(i, j)]]
  left  <- paren_label(splits, i, k)
  right <- paren_label(splits, k + 1, j)
  paste0("(", left, " · ", right, ")")
}

# Build the comparison table
paren_df <- tibble(
  id    = seq_along(all_parens),
  label = sapply(all_parens, paren_label),
  ops   = sapply(all_parens, function(s) count_ops(dims, s))
) |> arrange(ops) |>
  mutate(
    rank     = row_number(),
    is_best  = ops == min(ops),
    is_worst = ops == max(ops),
    is_naive = label == "((((A · B) · C) · D) · E)",
    highlight = case_when(
      is_best  ~ "★ Optimal",
      is_worst ~ "✗ Worst",
      is_naive ~ "→ Naive",
      TRUE     ~ ""
    )
  )

paren_df |>
  select(rank, label, ops, highlight) |>
  kable(
    col.names = c("Rank", "Parenthesization", "Scalar multiplications", ""),
    format.args = list(big.mark = ","),
    align = "clrc"
  )
```

The table tells a clear story. The optimal parenthesization `(A · ((B · C) · (D · E)))` requires only **93,840** scalar multiplications, while R's default left-to-right evaluation needs **17,200,000**, roughly **183 times more**. The worst case `(((A · B) · (C · D)) · E)` explodes to over **516 million** operations because it generates both an $800 \times 900$ and a $900 \times 700$ intermediate before even reaching the final product.

Why is the optimal ordering so cheap? It starts with the two "inner" products $B \cdot C$ ($5 \times 8$, costing 36,000 ops) and $D \cdot E$ ($8 \times 6$, costing 33,600 ops), which are both tiny. Then it combines them into a $5 \times 6$ matrix (240 ops), and only at the very end does it multiply the large $A$ into the result ($800 \times 5 \times 6 = 24{,}000$ ops). The narrow inner dimensions act as a funnel that keeps everything small.

```{r}
#| label: task1-ops-plot
#| fig-cap: "Operation count for all 14 parenthesizations of 5 matrices."
#| fig-height: 7

ggplot(paren_df, aes(x = reorder(label, ops), y = ops,
                     fill = case_when(
                       is_best  ~ "Optimal",
                       is_worst ~ "Worst",
                       is_naive ~ "Naive (L-to-R)",
                       TRUE     ~ "Other"
                     ))) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = format(ops, big.mark = ",")),
    hjust = -0.1, size = 3, fontface = "bold"
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(
    values = c("Optimal" = palette_main[1], "Worst" = palette_main[4],
               "Naive (L-to-R)" = palette_main[3], "Other" = "grey80"),
    name = NULL
  ) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Operation Count: All 14 Parenthesizations",
    subtitle = paste0("Chain dimensions: ", paste(dims, collapse = " × ")),
    x = NULL, y = "Scalar multiplications"
  )
```

The plot makes the disparity even more visual. The worst case is completely off the chart compared to everything else, and even the naive ordering is far from optimal. Most of the "reasonable" parenthesizations cluster in the low millions, but the optimal one sits at under 100K, an order of magnitude below the rest.

## Benchmarking with `bench::mark()`

Theory is nice, but let us see if the operation counts actually translate into real wall-clock time. We benchmark the three key cases: naive, optimal, and worst.

```{r}
#| label: task1-bench-setup

# Build R expressions from split tables so we can eval() them in the benchmark
build_expr <- function(splits, i = 1, j = 5) {
  if (i == j) return(as.name(LETTERS[i]))
  k <- splits[[paste(i, j)]]
  left  <- build_expr(splits, i, k)
  right <- build_expr(splits, k + 1, j)
  call("%*%", left, right)
}

best_splits  <- all_parens[[paren_df$id[paren_df$is_best][1]]]
worst_splits <- all_parens[[paren_df$id[paren_df$is_worst][1]]]

best_expr  <- build_expr(best_splits)
worst_expr <- build_expr(worst_splits)
naive_expr <- quote(A %*% B %*% C %*% D %*% E)
```

```{r}
#| label: task1-benchmark

invisible(eval(naive_expr))
invisible(eval(best_expr))
invisible(eval(worst_expr))
gc()

bm <- bench::mark(
  `Naive (L-to-R)` = eval(naive_expr),
  `Optimal`        = eval(best_expr),
  `Worst`          = eval(worst_expr),
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm |>
  mutate(
    expression = as.character(expression),
    `CV (sd/mean)` = round(bench_cv(bm), 3)
  ) |>
  select(expression, min, median, `CV (sd/mean)`, mem_alloc, n_itr) |>
  kable(caption = "Benchmark: time and memory for each parenthesization strategy",
        align = "lccccc")
```

This render measures median times of `r fmt_time(bench_value(bm, "Optimal", "median"))`, `r fmt_time(bench_value(bm, "Naive (L-to-R)", "median"))` and `r fmt_time(bench_value(bm, "Worst", "median"))` for optimal, naive and worst scalar-cost plans. CV is the sample standard deviation divided by the mean iteration time. Timing dispersion has several possible causes; this experiment does not identify them independently.

```{r}
#| label: task1-violin
#| fig-cap: "Violin plot of execution times (log scale). The three strategies span almost four orders of magnitude."

autoplot(bm, type = "violin") +
  labs(
    title    = "Execution Time Distribution",
    subtitle = "Each strategy benchmarked for at least 2 seconds",
    x = NULL
  ) +
  scale_fill_manual(values = palette_main[c(3, 1, 4)]) +
  theme(legend.position = "none")
```

The violin plots display the timing distributions from this run. The median worst/naive ratio is `r fmt_ratio(as.numeric(bench_value(bm, "Worst", "median")) / as.numeric(bench_value(bm, "Naive (L-to-R)", "median")))`. Multiplication counts exclude additions and are not FLOP totals; neither those counts nor matrix dimensions alone determine elapsed time.

## Why does it matter? Intermediate matrix sizes

The raw operation count does not tell the whole story. A bad ordering also means **huge intermediate matrices** that eat memory and trash the CPU cache. Let us trace the exact dimensions of every intermediate produced by each strategy.

```{r}
#| label: task1-intermediates

# Trace intermediate matrix dimensions for a given split table
trace_intermediates <- function(splits, dims, i = 1, j = 5) {
  if (i == j) return(list(intermediates = list(), dim = c(dims[i], dims[i + 1])))
  k <- splits[[paste(i, j)]]
  left  <- trace_intermediates(splits, dims, i, k)
  right <- trace_intermediates(splits, dims, k + 1, j)
  result_dim <- c(dims[i], dims[j + 1])
  intermediates <- c(
    left$intermediates, right$intermediates,
    list(list(
      rows = result_dim[1], cols = result_dim[2],
      elements = result_dim[1] * result_dim[2],
      bytes = result_dim[1] * result_dim[2] * 8
    ))
  )
  list(intermediates = intermediates, dim = result_dim)
}

naive_splits_full <- all_parens[[paren_df$id[paren_df$is_naive][1]]]

strategies <- list(
  "Optimal"        = best_splits,
  "Naive (L-to-R)" = naive_splits_full,
  "Worst"          = worst_splits
)

inter_df <- bind_rows(lapply(names(strategies), function(strat) {
  inters <- trace_intermediates(strategies[[strat]], dims)$intermediates
  tibble(
    strategy  = strat,
    step      = seq_along(inters),
    dim_label = sapply(inters, \(x) paste0(x$rows, " × ", x$cols)),
    elements  = sapply(inters, \(x) x$elements),
    kb        = sapply(inters, \(x) x$bytes / 1024)
  )
}))

inter_df |>
  kable(
    col.names = c("Strategy", "Step", "Intermediate size", "Elements", "Memory (KiB)"),
    format.args = list(big.mark = ","),
    digits = 1,
    align = "clcrc",
    caption = "Intermediate matrices generated at each multiplication step"
  )
```

This table is the key to understanding everything. Look at the optimal strategy: its first three intermediates are $5 \times 8$, $8 \times 6$, and $5 \times 6$, all under 50 elements. Only the final product ($800 \times 6$) is moderately sized. The naive strategy, on the other hand, starts with an $800 \times 900$ monster (720,000 elements, 5,625 KB) on the very first step.

```{r}
#| label: task1-memory-plot
#| fig-cap: "Peak and total memory allocated for intermediate matrices. The optimal ordering uses orders of magnitude less memory."
#| fig-height: 5

mem_summary <- inter_df |>
  summarize(
    peak_kb  = max(kb),
    total_kb = sum(kb),
    .by = strategy
  ) |>
  pivot_longer(cols = c(peak_kb, total_kb),
               names_to = "metric", values_to = "kb") |>
  mutate(
    metric = if_else(metric == "peak_kb", "Peak intermediate", "Total (all intermediates)"),
    strategy = factor(strategy, levels = c("Optimal", "Naive (L-to-R)", "Worst"))
  )

ggplot(mem_summary, aes(x = strategy, y = kb, fill = strategy)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(kb, 1), " KiB")),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(
    values = c("Optimal" = palette_main[1], "Naive (L-to-R)" = palette_main[3],
               "Worst" = palette_main[4])
  ) +
  scale_y_continuous(labels = label_comma(suffix = " KiB"),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Memory Cost of Intermediate Matrices",
    subtitle = "Peak = largest single intermediate; Total = sum of all intermediates",
    x = NULL, y = NULL
  ) +
  theme(legend.position = "none")
```

The largest individual output buffer is 37.5 KiB for the optimal plan and 5,625 KiB for both naive and worst plans. The table sums intermediate-buffer sizes, including the final result; it does not measure simultaneously live memory, process peak memory, or cache residency.

## Summary

To wrap up Task 1:

```{r}
#| label: task1-summary

peak_kb <- function(strategy) {
  mem_summary$kb[mem_summary$strategy == strategy &
                   mem_summary$metric == "Peak intermediate"][1]
}

task1_summary <- tibble(
  Metric = c("Scalar ops", "Median time", "Memory allocated", "Peak intermediate"),
  Optimal = c(
    format(min(paren_df$ops), big.mark = ","),
    fmt_time(bench_value(bm, "Optimal", "median")),
    fmt_bytes_binary(bench_value(bm, "Optimal", "mem_alloc")),
    paste0(round(peak_kb("Optimal"), 1), " KiB")
  ),
  `Naive (L-to-R)` = c(
    format(paren_df$ops[paren_df$is_naive], big.mark = ","),
    fmt_time(bench_value(bm, "Naive (L-to-R)", "median")),
    fmt_bytes_binary(bench_value(bm, "Naive (L-to-R)", "mem_alloc")),
    paste0(round(peak_kb("Naive (L-to-R)"), 1), " KiB")
  ),
  Worst = c(
    format(max(paren_df$ops), big.mark = ","),
    fmt_time(bench_value(bm, "Worst", "median")),
    fmt_bytes_binary(bench_value(bm, "Worst", "mem_alloc")),
    paste0(round(peak_kb("Worst"), 1), " KiB")
  ),
  `Naive/Optimal ratio` = c(
    fmt_ratio(paren_df$ops[paren_df$is_naive] / min(paren_df$ops)),
    fmt_ratio(as.numeric(bench_value(bm, "Naive (L-to-R)", "median")) /
                as.numeric(bench_value(bm, "Optimal", "median"))),
    fmt_ratio(as.numeric(bench_value(bm, "Naive (L-to-R)", "mem_alloc")) /
                as.numeric(bench_value(bm, "Optimal", "mem_alloc"))),
    fmt_ratio(peak_kb("Naive (L-to-R)") / peak_kb("Optimal"))
  )
)

task1_summary |>
  kable(align = "lrrrr", caption = "Task 1 summary derived from the rendered benchmark")
```

The conclusion is straightforward: **parenthesization order is not a minor optimization**. For matrices with heterogeneous dimensions, the wrong ordering can easily cost two to three orders of magnitude in both time and memory. R's default left-to-right `%*%` evaluation offers no protection against this. In the following tasks, we will build tools to automatically find and execute the optimal ordering.

---

# Task 2: C++ Naive Multiplication with Armadillo {#sec-task2}

Now we move from R to C++. The question is: does switching to a compiled language fix the ordering problem? Spoiler: not really. The bottleneck is algorithmic (how many FLOPs you do), not just interpretive overhead. But Armadillo does help in other ways, particularly memory management.

## The C++ naive function

We implement a simple function that takes a list of matrices and multiplies them left to right, exactly like R's `%*%` chaining. The function uses RcppArmadillo, which wraps LAPACK/BLAS for the heavy lifting. Input checks reject empty, nonfinite or dimensionally incompatible matrices.

```{Rcpp}
//| label: task2-cpp-naive

// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace Rcpp;

// [[Rcpp::export]]
arma::mat naive_multiply_cpp(const List& matrices) {
  int n = matrices.size();
  if (n == 0) stop("The matrix chain must not be empty");
  // Start with the first matrix and keep folding the chain left to right.
  arma::mat result = as<arma::mat>(matrices[0]);
  if (result.n_elem == 0 || !result.is_finite()) stop("Matrices must be nonempty and finite");
  for (int i = 1; i < n; i++) {
    // This mirrors A %*% B %*% C ... exactly: no regrouping involved.
    arma::mat next = as<arma::mat>(matrices[i]);
    if (next.n_elem == 0 || !next.is_finite()) stop("Matrices must be nonempty and finite");
    if (result.n_cols != next.n_rows) stop("Incompatible matrix dimensions");
    result = result * next;
  }
  return result;
}
```

The function is intentionally simple. It accepts a generic `List` from R, casts each element to an `arma::mat`, and accumulates the product left to right. There is no reordering, no parenthesization logic. This mirrors what R does with `A %*% B %*% C %*% D %*% E`, but in compiled C++ with Armadillo's memory management instead of R's allocator.

## Benchmarking C++ naive vs. R

We compare three cases: R naive (left to right), R optimal (manually parenthesized), and C++ naive (left to right via Armadillo). This lets us isolate the effect of switching language while keeping the same algorithm.

```{r}
#| label: task2-benchmark

mat_list <- list(A, B, C, D, E)

invisible(eval(naive_expr))
invisible(eval(best_expr))
invisible(naive_multiply_cpp(mat_list))
gc()

bm_task2 <- bench::mark(
  `R naive`     = eval(naive_expr),
  `R optimal`   = eval(best_expr),
  `C++ naive`   = naive_multiply_cpp(mat_list),
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm_task2 |>
  mutate(
    expression = as.character(expression),
    `CV (sd/mean)` = round(bench_cv(bm_task2), 3)
  ) |>
  select(expression, min, median, `CV (sd/mean)`, mem_alloc, n_itr) |>
  kable(caption = "Task 2 benchmark: R vs. C++ naive multiplication",
        align = "lccccc")
```

The table compares R and C++ with the same left-to-right ordering. Their median ratio is `r fmt_ratio(as.numeric(bench_value(bm_task2, "R naive", "median")) / as.numeric(bench_value(bm_task2, "C++ naive", "median")))`; values above 1 favor C++. Both perform the same scalar work, while conversion, allocation and library dispatch can differ.

The memory column records allocations visible to R: `r fmt_bytes_binary(bench_value(bm_task2, "R naive", "mem_alloc"))` for R naive and `r fmt_bytes_binary(bench_value(bm_task2, "C++ naive", "mem_alloc"))` for C++ naive. It omits native Armadillo heap allocations and therefore cannot compare total process memory.

```{r}
#| label: task2-violin
#| fig-cap: "Execution time comparison: R naive vs. R optimal vs. C++ naive. The C++ version is only a modest improvement for the naive ordering, while the optimal parenthesization dominates both."

autoplot(bm_task2, type = "violin") +
  labs(
    title    = "C++ Naive vs. R: Does the Language Matter?",
    subtitle = "Same naive ordering, different runtime environments",
    x = NULL
  ) +
  scale_fill_manual(values = palette_main[c(3, 1, 4)]) +
  theme(legend.position = "none")
```

The violin plot makes the key point of this task visually clear: **switching to C++ does not solve the ordering problem**. The two naive strategies live in the same performance band, while the manually parenthesized R solution remains much faster because it is solving a different computational problem: it performs far fewer scalar multiplications.

```{r}
#| label: task2-ratio-plot
#| fig-cap: "Speedup ratios relative to R naive. Reordering the multiplications matters far more than switching language."

bm_ratios <- bm_task2 |>
  mutate(
    expression = as.character(expression),
    median_ns  = as.numeric(median)
  ) |>
  mutate(
    baseline  = median_ns[expression == "R naive"],
    speedup   = baseline / median_ns
  )

ggplot(bm_ratios, aes(x = reorder(expression, -speedup), y = speedup,
                       fill = expression)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(speedup, 1), "x")),
            vjust = -0.3, size = 4, fontface = "bold") +
  scale_fill_manual(values = c(
    "R naive"    = palette_main[3],
    "R optimal"  = palette_main[1],
    "C++ naive"  = palette_main[4]
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Speedup Relative to R Naive",
    subtitle = "Algorithm beats language: optimal ordering wins by orders of magnitude",
    x = NULL, y = "Speedup factor (higher = faster)"
  ) +
  theme(legend.position = "none")
```

## Discussion

The results of Task 2 confirm an important principle: **algorithm dominates implementation**. Switching from R to C++ with the same naive ordering keeps us in the same broad performance band, while reordering the multiplications in pure R changes the cost by orders of magnitude. The lesson is clear:

- A bad algorithm in a fast language (C++ naive) loses to a good algorithm in a slow language (R optimal).
- Any C++ advantage is secondary: lower allocation overhead, no GC pressure, tighter memory management. For small chains like ours ($n = 5$), this does not add up to much.
- Where C++ *will* shine is in Tasks 3 and 4, where we combine the optimal ordering with compiled execution. That gives us both the algorithmic advantage and the implementation advantage.

The next step is obvious: we need a way to execute *any* ordering efficiently in C++, not just the hard-coded naive one. That is what Task 3's stack machine is for.

---

# Task 3: A General-Purpose Execution Engine {#sec-task3}

In Task 2 we hard-coded the naive ordering into C++. That is obviously not scalable: we cannot write a different C++ function for each of the 14 parenthesizations, let alone for chains of 20 or 100 matrices. What we need is a **generic execution engine** that takes any ordering as input and executes it efficiently.

The conceptual issue is this: a parenthesization such as `(A * ((B * C) * (D * E)))` is easy for a human to read, but awkward to hand over to a low-level function. C++ wants a compact, explicit execution plan, not a string that must be parsed or a separate hard-coded function for every case.

So from Task 3 onward we separate the problem into two stages:

1. **Planning:** represent the order of multiplications in a machine-friendly format.
2. **Execution:** feed that plan to a generic C++ engine that performs the multiplications in exactly that order.

## The RPN approach

The standard way to do this is to encode the parenthesization as a **Reverse Polish Notation (RPN)** vector, which is just the post-order traversal of the expression tree. If you have not seen RPN before, the key idea is simple: operands appear first, and the operation appears only after the data it needs is already available. That makes it perfect for stack-based execution.

In our encoding:

- A positive integer $i$ means "push matrix $A_i$ onto the stack."
- The value $-1$ means "pop two matrices, multiply them, push the result."

For example, the optimal ordering `(A * ((B * C) * (D * E)))` becomes:

$$
\underbrace{1}_{\text{push A}},\;
\underbrace{2, 3, -1}_{\text{B*C}},\;
\underbrace{4, 5, -1}_{\text{D*E}},\;
\underbrace{-1}_{\text{(BC)*(DE)}},\;
\underbrace{-1}_{\text{A*(BCDE)}}
$$

This representation has two practical advantages. First, it is compact: for $n$ matrices, the vector has length $2n - 1$. Second, it is executable without recursion or expression parsing. We first build a helper in R to convert any split table into an RPN vector:

```{r}
#| label: task3-rpn-builder

# Post-order traversal of the parenthesization tree -> RPN vector
# Positive integers = 1-based matrix indices, -1 = multiply opcode
build_rpn <- function(splits, i = 1, j = length(dims) - 1) {
  if (i == j) return(i)
  # Each split table entry stores where the final multiplication happens.
  k <- splits[[paste(i, j)]]
  c(build_rpn(splits, i, k), build_rpn(splits, k + 1, j), -1L)
}

rpn_optimal <- build_rpn(best_splits)
rpn_naive   <- build_rpn(naive_splits_full)

tibble(
  Ordering = c("Optimal", "Naive (L-to-R)"),
  Parenthesization = c(
    paren_df$label[paren_df$is_best],
    paren_df$label[paren_df$is_naive]
  ),
  `RPN vector` = c(
    paste(rpn_optimal, collapse = ", "),
    paste(rpn_naive, collapse = ", ")
  )
) |> kable(align = "llc")
```

The naive ordering pushes each matrix and immediately multiplies: `1, 2, -1, 3, -1, ...` (strictly left-associative). The optimal ordering pushes groups first and multiplies them internally before combining: `1, 2, 3, -1, 4, 5, -1, -1, -1`.

It is useful to see why this works mechanically. For the optimal RPN vector:

1. `1` pushes `A`.
2. `2, 3, -1` pushes `B` and `C`, then replaces them with `BC`.
3. `4, 5, -1` pushes `D` and `E`, then replaces them with `DE`.
4. The next `-1` multiplies the two intermediate results, producing `(BC)(DE)`.
5. The final `-1` multiplies `A` by that result.

At every multiplication step, the two operands we need are already on top of the stack. That is exactly why post-order traversal and stack evaluation fit together so well.

## The C++ stack machine

The executor is a classic stack machine. A **stack** is a last-in, first-out structure: the last matrix pushed is the first one popped. That is exactly what we need here, because each `-1` token should combine the two most recently available sub-results.

The loop is therefore conceptually very simple:

1. Read the next token.
2. If it is positive, push the corresponding matrix.
3. If it is `-1`, pop the top two matrices, multiply them, and push the result back.

We also use `std::move()` when popping from the stack to avoid unnecessary copies of large intermediate matrices. In plain language, move semantics let us transfer ownership of a matrix buffer instead of duplicating all of its data.

```{Rcpp}
//| label: task3-stack-machine

// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
#include <stack>
#include <vector>
using namespace Rcpp;

// [[Rcpp::export]]
arma::mat execute_rpn(const List& matrices, const IntegerVector& rpn) {
  int n = matrices.size();
  if (n == 0 || rpn.size() != 2LL * n - 1) stop("RPN must encode one nonempty matrix chain");
  int next_operand = 1;
  
  // --- R-memory wrappers for input conversion ---
  // Build lightweight arma::mat wrappers that initially reuse R's memory.
  // The constructor mat(ptr, n_rows, n_cols, copy_aux_mem=false, strict=true)
  // points Armadillo directly at R's column-major SEXP buffer,
  // avoiding the deep copy that as<arma::mat>() would perform.
  std::vector<arma::mat> mats;
  mats.reserve(n);
  for (int i = 0; i < n; i++) {
    NumericMatrix rm = matrices[i];
    arma::mat wrapper(rm.begin(), rm.nrow(), rm.ncol(), false, true);
    if (wrapper.n_elem == 0 || !wrapper.is_finite()) stop("Matrices must be nonempty and finite");
    mats.push_back(wrapper); // Own a copy; the temporary wrapper is not retained.
  }
  
  // --- Stack machine ---
  std::stack<arma::mat> stk;
  
  for (int t = 0; t < rpn.size(); t++) {
    int token = rpn[t];
    if (token > 0) {
      if (token > n || token != next_operand++) stop("RPN operands must occur once in original chain order");
      // Positive: push an owned working copy; intermediates are subsequently moved.
      stk.push(mats[token - 1]);
    } else if (token == -1) {
      if (stk.size() < 2) stop("Malformed RPN: multiplication needs two matrices");
      // -1: pop two, multiply, push result
      arma::mat right = std::move(stk.top()); stk.pop();
      arma::mat left  = std::move(stk.top()); stk.pop();
      if (left.n_cols != right.n_rows) stop("Incompatible matrix dimensions");
      stk.push(left * right);
    } else {
      stop("Malformed RPN token");
    }
  }

  if (stk.size() != 1 || next_operand != n + 1) stop("Malformed RPN: expected one final matrix");
  return stk.top();
}
```

A few things worth noting about this implementation:

1. **Input conversion**: the initial wrappers use R's column-major memory and
avoid repeated `as<arma::mat>()` conversion. The stack still owns working
copies when it evaluates a plan, so this is not an end-to-end zero-copy
executor; memory claims should be verified with profiling.
2. **`std::move(stk.top())`**: when we pop a matrix from the stack, we use move semantics instead of copying. For an $800 \times 900$ intermediate, this means transferring a pointer (8 bytes) instead of copying 5.7 MB of data.
3. **`const List&`**: the input list is passed by reference, so the R matrices are not copied when entering the C++ function.
4. **Token-driven loop**: the function does not know or care what the "optimal" ordering is. It simply follows instructions. This is what makes it generic: any valid RPN vector produces the correct result.

```{r}
#| label: task3-verify

# Verify the stack machine gives the right answer for both orderings
res_rpn_opt   <- execute_rpn(mat_list, rpn_optimal)
res_rpn_naive <- execute_rpn(mat_list, rpn_naive)

tibble(
  Test = c("RPN optimal vs R optimal", "RPN naive vs R naive"),
  `Max absolute difference` = c(
    max(abs(res_rpn_opt - eval(best_expr))),
    max(abs(res_rpn_naive - eval(naive_expr)))
  )
) |> kable(align = "lc")
```

The table reports actual absolute differences. Assertions compare all 14 execution plans with R under a relative tolerance; exact bitwise identity is not required across BLAS backends.

## Benchmarking: stack machine vs. everything else

Now the interesting question: how does the generic stack machine compare to the hard-coded approaches from Tasks 1 and 2? We test five configurations.

```{r}
#| label: task3-benchmark

invisible(eval(naive_expr))
invisible(eval(best_expr))
invisible(naive_multiply_cpp(mat_list))
invisible(execute_rpn(mat_list, rpn_naive))
invisible(execute_rpn(mat_list, rpn_optimal))
gc()

bm_task3 <- bench::mark(
  `R naive`            = eval(naive_expr),
  `R optimal`          = eval(best_expr),
  `C++ naive (Task 2)` = naive_multiply_cpp(mat_list),
  `C++ RPN naive`      = execute_rpn(mat_list, rpn_naive),
  `C++ RPN optimal`    = execute_rpn(mat_list, rpn_optimal),
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm_task3 |>
  mutate(
    expression = as.character(expression),
    `CV (sd/mean)` = round(bench_cv(bm_task3), 3)
  ) |>
  select(expression, min, median, `CV (sd/mean)`, mem_alloc, n_itr) |>
  kable(caption = "Task 3 benchmark: stack machine vs. previous approaches",
        align = "lccccc")
```

There are two comparisons worth making here:

1. **C++ RPN naive vs C++ naive (Task 2):** the two are very close in the table above. That means the extra abstraction of reading an instruction vector and using a stack does not materially hurt performance.

2. **C++ RPN optimal vs R optimal:** both are much faster than the naive variants because both are executing the same good multiplication order. Any small difference now comes from execution overhead, not from algorithmic cost.

The key takeaway is that **the stack machine buys us generality almost for free**. We can now execute any parenthesization encoded as an RPN vector, which is exactly the interface Task 4 will need.

```{r}
#| label: task3-comparison-plot
#| fig-cap: "Full comparison across Tasks 1--3. The stack machine (RPN) matches the hard-coded approaches while being fully generic."

bm_plot_df <- bm_task3 |>
  mutate(
    expression = as.character(expression),
    median_us  = as.numeric(median) * 1e6,
    group = case_when(
      grepl("naive", expression, ignore.case = TRUE) &
        !grepl("optimal", expression, ignore.case = TRUE) ~ "Naive ordering",
      TRUE ~ "Optimal ordering"
    )
  )

ggplot(bm_plot_df, aes(x = reorder(expression, median_us),
                        y = median_us, fill = group)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = if_else(
    median_us > 1000,
    paste0(round(median_us / 1000, 1), " ms"),
    paste0(round(median_us), " \u00b5s")
  )), hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip(clip = "off") +
  scale_fill_manual(
    values = c("Naive ordering" = palette_main[3], "Optimal ordering" = palette_main[1]),
    name = NULL
  ) +
  scale_y_log10(
    labels = function(x) if_else(x >= 1000, paste0(x/1000, " ms"), paste0(x, " \u00b5s")),
    expand = expansion(mult = c(0, 0.3))
  ) +
  labs(
    title    = "Full Benchmark: Tasks 1 through 3",
    subtitle = "Log scale. Ordering dominates; compare the measured executor overhead.",
    x = NULL, y = "Median time (log scale)"
  )
```

The log-scale bar chart tells the whole story of Tasks 1--3. The naive approaches cluster together, and the optimal approaches cluster together. The separating factor is the **algorithm** (which multiplications we choose), not the **language** (R vs C++) or the **execution mechanism** (direct loop vs stack machine).

## Discussion

Task 3 gives us a powerful building block: a generic execution engine that can run any parenthesization encoded as an RPN vector. This is important because:

- We no longer need to generate and `eval()` R expressions, which was a bit of a hack in Task 1.
- The same C++ function handles any chain length, not just 5 matrices.
- The RPN encoding is compact ($2n - 1$ integers) and easy to produce from a DP algorithm.

What we still lack is the algorithm that *finds* the optimal RPN vector automatically. Right now we are computing it in R by enumerating all 14 parenthesizations, which is $O(C_n)$ and only feasible for tiny $n$. For larger chains, we need dynamic programming. That is the subject of Task 4.

---

# Task 4: Finding the Optimal Order Automatically {#sec-task4}

This is the core algorithmic task. We need to go from "try all parenthesizations" (exponential) to something that scales. The classical answer is dynamic programming in $O(n^3)$, which we implement first in R, then in C++. After that, we push further: a parallel wavefront variant that exploits multi-core CPUs, and a resource-aware cost function inspired by recent HPC literature that accounts for memory hierarchy effects.

From a teaching point of view, this section answers three different questions:

1. **Why does dynamic programming apply here at all?**
2. **How do we reconstruct an actual multiplication plan, not just a cost?**
3. **Why might a hardware-aware objective differ from pure FLOP counting?**

## The Godbole algorithm: $O(n^3)$ dynamic programming

Dynamic programming works because the matrix-chain problem has **optimal substructure** and **overlapping subproblems**.

- **Optimal substructure** means that if the best way to compute $A_i \cdots A_j$ splits at position $k$, then the left side $A_i \cdots A_k$ and the right side $A_{k+1} \cdots A_j$ must themselves also be solved optimally.
- **Overlapping subproblems** means the same sub-chains appear again and again across different parenthesizations. Instead of recomputing them exponentially many times, we compute each interval once and store the answer.

The idea is therefore to fill two tables bottom-up:

- **Cost table $m[i,j]$**: the minimum number of scalar multiplications needed to compute the sub-chain $A_i \cdots A_j$.
- **Split table $s[i,j]$**: the index $k$ where we should split the sub-chain to achieve that minimum.

The recurrence is:

$$
m[i, j] = \min_{i \le k < j} \left\{ m[i, k] + m[k+1, j] + p_{i-1} \cdot p_k \cdot p_j \right\}
$$

with base case $m[i, i] = 0$. We iterate by chain length: first all pairs ($L = 2$), then triples ($L = 3$), and so on up to the full chain ($L = n$). Each cell only depends on shorter intervals, so a bottom-up traversal guarantees that every dependency has already been computed when we need it.

This is the same dependency pattern we will later exploit for parallelization: every diagonal of the table can be computed once the previous diagonals are known.

### R implementation

```{r}
#| label: task4-dp-r

mcm_dp_r <- function(dims) {
  if (length(dims) < 2L || any(!is.finite(dims)) || any(dims <= 0) || any(dims != floor(dims)) || any(dims > .Machine$integer.max)) {
    stop("dims must contain at least two positive finite dimensions")
  }
  dims <- as.double(dims) # Avoid integer overflow in products supplied as integer vectors.
  n <- length(dims) - 1
  m <- matrix(0, nrow = n, ncol = n)
  s <- matrix(0L, nrow = n, ncol = n)
  if (n == 1L) return(list(cost = m, splits = s, objective_value = 0))
  
  # Fill the DP tables by diagonal: short chains first, long chains later.
  for (len in 2:n) {
    for (i in 1:(n - len + 1)) {
      j <- i + len - 1
      m[i, j] <- Inf
      for (k in i:(j - 1)) {
        cost <- m[i, k] + m[k + 1, j] + dims[i] * dims[k + 1] * dims[j + 1]
        if (cost < m[i, j]) {
          m[i, j] <- cost
          # Remember the split so we can rebuild the execution plan later.
          s[i, j] <- k
        }
      }
    }
  }
  
  list(cost = m, splits = s, objective_value = m[1, n])
}

# Helper: convert split matrix to RPN vector
build_rpn_from_matrix <- function(s, i = 1, j = nrow(s)) {
  if (i == j) return(i)
  k <- s[i, j]
  c(build_rpn_from_matrix(s, i, k), build_rpn_from_matrix(s, k + 1, j), -1L)
}

dp_r <- mcm_dp_r(dims)
rpn_from_dp <- build_rpn_from_matrix(dp_r$splits)

tibble(
  Source = c("Brute-force (Task 1)", "DP (Task 4)"),
  `Optimal ops` = c(min(paren_df$ops), dp_r$objective_value),
  `RPN vector` = c(paste(rpn_optimal, collapse = ", "),
                    paste(rpn_from_dp, collapse = ", "))
) |> kable(format.args = list(big.mark = ","), align = "lrc")
```

The DP gives exactly the same answer as our brute-force enumeration: **93,840 operations**, with the same RPN vector `[1, 2, 3, -1, 4, 5, -1, -1, -1]`. This is expected for $n = 5$, but the DP runs in $O(n^3)$ instead of $O(C_{n-1})$, so it scales to thousands of matrices where brute force is hopeless.

### Three C++ implementations

Now we implement three progressively more sophisticated versions in C++:

1. **Standard DP**: direct translation of the Godbole algorithm with flat arrays for cache-friendly access.
2. **Parallel wavefront DP**: same algorithm, but the inner loop over starting index $i$ is parallelized with OpenMP for each diagonal $L$. All cells on the same diagonal are independent (they depend only on shorter sub-chains already computed on previous diagonals), so they can be computed simultaneously across CPU cores.
3. **Resource-aware DP**: inspired by Nissim, Schwartz & Shabo (2024), this variant replaces the pure FLOP cost with a penalized cost that accounts for **memory hierarchy effects**. When an intermediate matrix exceeds a cache size threshold, the cost is multiplied by a penalty factor. The intuition: even if an ordering has fewer FLOPs, if it creates intermediates that spill out of L2/L3 cache, the resulting cache thrashing may make it slower in practice than an ordering with slightly more FLOPs but better data locality.

The important distinction is that the first two algorithms optimize the **same mathematical objective** as the textbook problem: scalar multiplication count. The third optimizes a **different surrogate objective**: scalar work plus hardware penalties. That is why later on we will compare both the recovered RPN vector and the objective value, rather than calling everything "minimum operations".

To isolate the OpenMP contribution from the algorithmic improvements, I check whether the report was compiled with OpenMP enabled and benchmark the wavefront code with one thread versus several threads on a larger optimizer-only case.

```{Rcpp}
//| label: task4-dp-cpp

// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
#include <vector>
#include <stack>
#include <limits>
#include <cmath>
#include <climits>
#ifdef _OPENMP
  #include <omp.h>
#endif
using namespace Rcpp;

int checked_chain_size(const NumericVector& dims) {
  if (dims.size() < 2 || dims.size() > 46341)
    stop("Supply 2 to 46341 positive integer dimensions");
  for (double d : dims) {
    if (!std::isfinite(d) || d <= 0 || d > INT_MAX || d != std::floor(d))
      stop("Dimensions must be positive finite integers within the int32 range");
  }
  return static_cast<int>(dims.size()) - 1;
}

// Helper: reconstruct RPN from flat split table -- ITERATIVE version.
// Avoids deep recursion that could overflow the call stack for large n
// (e.g. n > 10,000 in genomic simulations). Uses an explicit work-stack
// that simulates post-order traversal with a state machine.
void build_rpn_iterative(int root_i, int root_j,
                         const std::vector<int>& s, int n,
                         std::vector<int>& rpn) {
  rpn.clear();
  // State machine: 0 = enter node, 1 = left child done, 2 = right child done
  struct Frame { int i; int j; int state; };
  std::stack<Frame> stk;
  stk.push({root_i, root_j, 0});
  while (!stk.empty()) {
    Frame& f = stk.top();
    if (f.i == f.j) {
      rpn.push_back(f.i + 1);    // 1-based for R
      stk.pop();
    } else if (f.state == 0) {
      int k = s[f.i * n + f.j];
      f.state = 1;
      stk.push({f.i, k, 0});     // push left subtree
    } else if (f.state == 1) {
      int k = s[f.i * n + f.j];
      f.state = 2;
      stk.push({k + 1, f.j, 0}); // push right subtree
    } else {
      rpn.push_back(-1);          // post-order: emit multiply after children
      stk.pop();
    }
  }
}

// 1. Standard O(n^3) DP with flat arrays
//    Uses `double` for costs (exact integer representation up to 2^53) and floating-point
//    dimension products; table sizes are validated before allocation.
// [[Rcpp::export]]
List mcm_dp_cpp(const NumericVector& dims) {
  int n = checked_chain_size(dims);
  std::vector<double> m(n * n, 0.0);
  std::vector<int> s(n * n, 0);
  
  for (int len = 2; len <= n; len++) {
    for (int i = 0; i < n - len + 1; i++) {
      int j = i + len - 1;
      // Flat index: cell (i, j) lives at i * n + j.
      m[i * n + j] = std::numeric_limits<double>::infinity();
      for (int k = i; k < j; k++) {
        double cost = m[i * n + k] + m[(k + 1) * n + j] +
          (double)dims[i] * (double)dims[k + 1] * (double)dims[j + 1];
        if (cost < m[i * n + j]) {
          m[i * n + j] = cost;
          s[i * n + j] = k;
        }
      }
    }
  }
  
  if (!std::isfinite(m[n - 1])) stop("DP objective overflow");
  std::vector<int> rpn;
  rpn.reserve(2 * n - 1);
  build_rpn_iterative(0, n - 1, s, n, rpn);
  
  return List::create(
    Named("objective_value") = m[0 * n + (n - 1)],
    Named("rpn") = IntegerVector(rpn.begin(), rpn.end())
  );
}

// Report whether this build actually has OpenMP enabled.
// This keeps the notebook honest: a "parallel" code path compiled without
// OpenMP is just sequential code with ignored pragmas.
// [[Rcpp::export]]
List openmp_status_cpp() {
  #ifdef _OPENMP
  return List::create(
    Named("enabled") = true,
    Named("max_threads") = omp_get_max_threads(),
    Named("version") = _OPENMP
  );
  #else
  return List::create(
    Named("enabled") = false,
    Named("max_threads") = 1,
    Named("version") = NA_INTEGER
  );
  #endif
}

// 2. Parallel wavefront DP (OpenMP on each diagonal)
//    `min_cells_parallel` avoids spawning a parallel region for tiny diagonals,
//    where the OpenMP overhead would cost more than the work itself.
// [[Rcpp::export]]
List mcm_dp_parallel(const NumericVector& dims,
                     int n_threads = 0,
                     int min_cells_parallel = 64) {
  int n = checked_chain_size(dims);
  std::vector<double> m(n * n, 0.0);
  std::vector<int> s(n * n, 0);
  
  if (n_threads < 0 || min_cells_parallel < 1) stop("Invalid parallel configuration");
  int actual_threads = 1;
  #ifdef _OPENMP
  actual_threads = n_threads > 0 ? n_threads : omp_get_max_threads();
  #endif
  
  for (int len = 2; len <= n; len++) {
    int num_cells = n - len + 1;
    // Every cell on the same diagonal has the same amount of work, so a
    // static schedule keeps the runtime overhead low and the code simple.
    #pragma omp parallel for schedule(static) num_threads(actual_threads) if(actual_threads > 1 && num_cells >= min_cells_parallel)
    for (int i = 0; i < num_cells; i++) {
      int j = i + len - 1;
      double best = std::numeric_limits<double>::infinity();
      int best_k = i;
      for (int k = i; k < j; k++) {
        double cost = m[i * n + k] + m[(k + 1) * n + j] +
          (double)dims[i] * (double)dims[k + 1] * (double)dims[j + 1];
        if (cost < best) {
          best = cost;
          best_k = k;
        }
      }
      m[i * n + j] = best;
      s[i * n + j] = best_k;
    }
  }
  
  if (!std::isfinite(m[n - 1])) stop("DP objective overflow");
  std::vector<int> rpn;
  rpn.reserve(2 * n - 1);
  build_rpn_iterative(0, n - 1, s, n, rpn);
  
  return List::create(
    Named("objective_value") = m[0 * n + (n - 1)],
    Named("rpn") = IntegerVector(rpn.begin(), rpn.end())
  );
}

// 3. Resource-aware DP: penalizes intermediates that exceed cache
//    Inspired by the IO-complexity model of Hong & Kung (1981) and
//    motivated by communication costs; it is not the algorithm of Nissim et al. (2024).
//    Uses a step-function penalty across two cache tiers (L2 and L3),
//    plus a dimension-dependent small-product penalty that penalizes very small
//    multiplications where BLAS call overhead and poor vectorization
//    dominate the actual computation.
// [[Rcpp::export]]
List mcm_dp_resource_aware(const NumericVector& dims,
                            double l2_bytes     = 262144,
                            double l3_bytes     = 8388608,
                            double l2_penalty   = 1.5,
                            double l3_penalty   = 3.0,
                            double simd_overhead = 500.0) {
  if (!std::isfinite(l2_bytes) || !std::isfinite(l3_bytes) || l2_bytes <= 0 || l3_bytes < l2_bytes ||
      !std::isfinite(l2_penalty) || !std::isfinite(l3_penalty) || l2_penalty < 1 || l3_penalty < l2_penalty ||
      !std::isfinite(simd_overhead) || simd_overhead < 0) stop("Invalid surrogate-cost parameters");
  int n = checked_chain_size(dims);
  std::vector<double> m(n * n, 0.0);
  std::vector<int> s(n * n, 0);
  
  for (int len = 2; len <= n; len++) {
    for (int i = 0; i < n - len + 1; i++) {
      int j = i + len - 1;
      double best = std::numeric_limits<double>::infinity();
      int best_k = i;
      for (int k = i; k < j; k++) {
        double flops = (double)dims[i] * (double)dims[k + 1] * (double)dims[j + 1];
        
        // Two-tier cache penalty: L2 and L3
        auto tier_penalty = [&](double bytes) -> double {
          if (bytes > l3_bytes) return l3_penalty;
          if (bytes > l2_bytes) return l2_penalty;
          return 1.0;
        };
        
        double result_bytes = (double)dims[i] * (double)dims[j + 1] * 8.0;
        double left_bytes   = (double)dims[i] * (double)dims[k + 1] * 8.0;
        double right_bytes  = (double)dims[k + 1] * (double)dims[j + 1] * 8.0;
        
        double penalty = tier_penalty(result_bytes)
                       * tier_penalty(left_bytes)
                       * tier_penalty(right_bytes);
        
        // Small products have a disproportionately large setup cost.  Unlike
        // a constant term, this dimension-dependent penalty can affect the
        // selected parenthesization.
        double small_product_penalty = simd_overhead * (1.0 + 1e6 / std::max(flops, 1.0));
        double cost = m[i * n + k] + m[(k + 1) * n + j]
                    + flops * penalty + small_product_penalty;
        if (cost < best) {
          best = cost;
          best_k = k;
        }
      }
      m[i * n + j] = best;
      s[i * n + j] = best_k;
    }
  }
  
  if (!std::isfinite(m[n - 1])) stop("DP objective overflow");
  std::vector<int> rpn;
  rpn.reserve(2 * n - 1);
  build_rpn_iterative(0, n - 1, s, n, rpn);
  
  return List::create(
    Named("objective_value") = m[0 * n + (n - 1)],
    Named("rpn") = IntegerVector(rpn.begin(), rpn.end())
  );
}
```

A few implementation notes:

- **Flat arrays**: all three variants store the $m$ and $s$ tables as `std::vector<double>` indexed by `i * n + j` rather than nested vectors or Armadillo matrices. This gives contiguous memory layout and better cache behavior during the triple-nested-loop traversal. The cost entries use `double` (safe up to $2^{53}$), and all dimension products are explicitly cast to `double` before multiplication to avoid `int32` overflow when dimensions exceed ~46,000.
- **Iterative RPN reconstruction**: instead of the textbook recursive approach, we use an explicit `std::stack<Frame>` that simulates post-order traversal iteratively. Recursive reconstruction consumes one call-stack frame per level of the expression tree; for chains of $n > 10{,}000$ matrices (common in auto-generated genomic or simulation pipelines), this would exhaust the OS call stack (~8 MB on Linux/macOS) and segfault. The iterative version moves this bookkeeping to heap-allocated memory, which is effectively unlimited.
- **`#pragma omp parallel for`**: the parallel variant annotates the loop over starting index `i` for each chain length `len`. Since all cells on the same diagonal are independent, this is safe without any synchronization. I use `schedule(static)` because every cell on a given diagonal performs the same number of split evaluations. The `min_cells_parallel` guard avoids paying thread-management overhead on tiny diagonals, while still allowing the report to exercise the real threaded path on medium-sized chains.
- **Two-tier cache penalty + small-product heuristic**: the resource-aware
  variant uses configurable L2/L3 thresholds plus a dimension-dependent
  penalty for tiny products. It is a hardware-sensitive heuristic, not a
  calibrated performance model; benchmarked execution time is the criterion
  for accepting a selected plan.

### Parallel configuration in this render

```{r}
#| label: task4-openmp-status

openmp_info <- openmp_status_cpp()
parallel_threads <- if (isTRUE(openmp_info$enabled)) {
  min(4L, as.integer(openmp_info$max_threads))
} else {
  1L
}
parallel_min_cells <- 64L

tibble(
  `OpenMP enabled` = if (isTRUE(openmp_info$enabled)) "Yes" else "No",
  `Threads used` = parallel_threads,
  `Activation threshold` = paste(parallel_min_cells, "cells per diagonal"),
  `Recommended render path` = if (isTRUE(openmp_info$enabled)) "Current build is fine" else "Use Docker"
) |> kable(
  caption = "Parallel configuration used in this render",
  align = "lccc"
)
```

`r if (isTRUE(openmp_info$enabled)) { "OpenMP is active in this render: the parallel benchmarks below reflect real multi-threaded execution. The Docker image is provided as an alternative reproducibility path for environments where libomp is not available." } else { "If OpenMP enabled is No, the wavefront implementation still returns the correct answer, but the pragmas are ignored and the code runs sequentially. To reproduce the threaded benchmark, build and run the provided Docker image: docker build -t mcop . && docker run -v \\\"$(pwd)/output:/output\\\" mcop" }`

### Verification: same ordering, different objectives

```{r}
#| label: task4-verify

r1 <- mcm_dp_cpp(dims)
r2 <- mcm_dp_parallel(
  dims,
  n_threads = parallel_threads,
  min_cells_parallel = parallel_min_cells
)
r3 <- mcm_dp_resource_aware(dims)  # uses L2=256KB, L3=8MB defaults

tibble(
  Algorithm = c("Standard DP", "Parallel wavefront DP", "Resource-aware DP"),
  `Objective optimized` = c("Scalar ops", "Scalar ops", "Penalized cost"),
  `Objective value` = c(r1$objective_value, r2$objective_value, r3$objective_value),
  `RPN vector` = c(
    paste(r1$rpn, collapse = ", "),
    paste(r2$rpn, collapse = ", "),
    paste(r3$rpn, collapse = ", ")
  ),
  `Matches brute-force?` = c(
    identical(r1$rpn, as.integer(rpn_optimal)),
    identical(r2$rpn, as.integer(rpn_optimal)),
    identical(r3$rpn, as.integer(rpn_optimal))
  )
) |> kable(format.args = list(big.mark = ","), align = "llccc")
```

The table reports whether each selected plan matches the enumerated scalar-cost optimum. Standard and parallel DP minimize multiplication count. The resource-aware variant minimizes its stated surrogate and may select a different plan; its objective is not an operation count.

The surrogate is tested against exhaustive enumeration for this small example. Its cache thresholds and penalties are illustrative defaults, not measurements of the current processor or evidence of a runtime improvement.

```{r}
#| label: correctness-checks
expect_error <- function(expr) stopifnot(inherits(try(force(expr), silent = TRUE), "try-error"))
for (plan in all_parens) {
  reference <- eval(build_expr(plan))
  actual <- execute_rpn(mat_list, build_rpn(plan))
  stopifnot(isTRUE(all.equal(actual, reference, tolerance = 1e-10)))
}
set.seed(301)
for (nn in 2:7) {
  dd <- sample(2:20, nn + 1L, replace = TRUE)
  expected <- min(vapply(generate_all_parens(nn), function(plan) count_ops(dd, plan), numeric(1)))
  stopifnot(mcm_dp_r(dd)$objective_value == expected,
            mcm_dp_cpp(dd)$objective_value == expected,
            mcm_dp_parallel(dd, 2L, 1L)$objective_value == expected)
}
stopifnot(mcm_dp_r(c(50000L,50000L,50000L))$objective_value == 125000000000000,
          mcm_dp_cpp(c(7,9))$objective_value == 0,
          identical(mcm_dp_cpp(c(7,9))$rpn, 1L),
          isTRUE(all.equal(naive_multiply_cpp(mat_list), eval(naive_expr), tolerance = 1e-10)))
for (bad in list(numeric(), 4, c(2,0), c(NA,3), c(2.5,4), c(Inf,3))) {
  expect_error(mcm_dp_r(bad)); expect_error(mcm_dp_cpp(bad))
  expect_error(mcm_dp_parallel(bad)); expect_error(mcm_dp_resource_aware(bad))
}
expect_error(naive_multiply_cpp(list()))
expect_error(naive_multiply_cpp(list(matrix(1,2,3), matrix(1,2,2))))
expect_error(execute_rpn(mat_list, c(1L,1L,-1L)))
expect_error(execute_rpn(mat_list, c(-1L,1:5,-1L,-1L,-1L)))
expect_error(mcm_dp_resource_aware(dims, l2_penalty = NA_real_))
# Independent exhaustive evaluation of the custom surrogate over all 14 plans.
surrogate_cost <- function(plan, i = 1L, j = length(dims)-1L) {
  if (i == j) return(0)
  k <- plan[[paste(i,j)]]
  work <- dims[i]*dims[k+1L]*dims[j+1L]
  buffers <- c(dims[i]*dims[j+1L], dims[i]*dims[k+1L], dims[k+1L]*dims[j+1L])*8
  tiers <- ifelse(buffers > 8388608, 3, ifelse(buffers > 262144, 1.5, 1))
  surrogate_cost(plan,i,k)+surrogate_cost(plan,k+1L,j)+work*prod(tiers)+500*(1+1e6/max(work,1))
}
stopifnot(isTRUE(all.equal(r3$objective_value, min(vapply(all_parens,surrogate_cost,numeric(1))))))
```

## The grand benchmark: everything together

Now we bring everything from Tasks 1--4 together in a single benchmark. For the Task 4 entries, we measure the **full pipeline**: optimizer + executor, so the cost of finding the optimal ordering is included.

```{r}
#| label: task4-grand-benchmark

invisible(eval(naive_expr))
invisible(eval(best_expr))
invisible(naive_multiply_cpp(mat_list))
invisible(execute_rpn(mat_list, rpn_optimal))
invisible(execute_rpn(mat_list, r1$rpn))
invisible(execute_rpn(mat_list, r2$rpn))
invisible(execute_rpn(mat_list, r3$rpn))
gc()

bm_grand <- bench::mark(
  `R naive`                   = eval(naive_expr),
  `R optimal (hand)`          = eval(best_expr),
  `C++ naive (Task 2)`        = naive_multiply_cpp(mat_list),
  `C++ RPN optimal (Task 3)`  = execute_rpn(mat_list, rpn_optimal),
  `R DP + execute`            = {
    dp <- mcm_dp_r(dims)
    rpn <- build_rpn_from_matrix(dp$splits)
    execute_rpn(mat_list, rpn)
  },
  `C++ DP + execute`          = {
    dp <- mcm_dp_cpp(dims)
    execute_rpn(mat_list, dp$rpn)
  },
  `C++ parallel DP + execute` = {
    dp <- mcm_dp_parallel(
      dims,
      n_threads = parallel_threads,
      min_cells_parallel = parallel_min_cells
    )
    execute_rpn(mat_list, dp$rpn)
  },
  `C++ resource-aware + exec` = {
    dp <- mcm_dp_resource_aware(dims)
    execute_rpn(mat_list, dp$rpn)
  },
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm_grand |>
  mutate(
    expression = as.character(expression),
    `CV (sd/mean)` = round(bench_cv(bm_grand), 3)
  ) |>
  select(expression, min, median, `CV (sd/mean)`, mem_alloc, n_itr) |>
  kable(
    caption = "Grand benchmark: all approaches from Tasks 1--4",
    align = "lccccc"
  )
```

The grand table compares all implementations in one benchmark session. The Task 4 entries include planning and execution; the hand-parenthesized and precomputed-RPN entries exclude planning. Their measured differences should be read from this table without assuming equality or a fixed ranking.

### Decomposition: optimize vs. execute

One natural concern with the grand benchmark is that it conflates two distinct costs: finding the optimal ordering and executing it. In many real-world scenarios (e.g., training a neural network), you optimize the chain **once** and execute it **millions of times** with different data. The next benchmark separates the two to confirm that the DP overhead is not distorting the comparison:

```{r}
#| label: task4-decompose

invisible(mcm_dp_cpp(dims))
invisible(execute_rpn(mat_list, r1$rpn))
gc()

bm_decompose <- bench::mark(
  `C++ DP (optimize only)` = mcm_dp_cpp(dims),
  `C++ execute (pre-computed RPN)` = execute_rpn(mat_list, r1$rpn),
  check = FALSE,
  min_time = 2,
  min_iterations = 50,
  filter_gc = FALSE
)

bm_decompose |>
  mutate(
    expression = as.character(expression),
    `CV (sd/mean)` = round(bench_cv(bm_decompose), 3)
  ) |>
  select(expression, min, median, `CV (sd/mean)`, mem_alloc) |>
  kable(caption = "Decomposed: optimization vs. execution (n = 5)",
        align = "lcccc")
```

For $n = 5$, the optimization-only benchmark is tiny compared with execution. In this render it is `r fmt_time(bench_value(bm_decompose, "C++ DP (optimize only)", "median"))` versus `r fmt_time(bench_value(bm_decompose, "C++ execute (pre-computed RPN)", "median"))`, so optimization is only about `r round(100 * as.numeric(bench_value(bm_decompose, "C++ DP (optimize only)", "median")) / as.numeric(bench_value(bm_decompose, "C++ execute (pre-computed RPN)", "median")), 1)`% of the execution cost. That makes the plan-computation overhead essentially negligible when the same multiplication pattern is reused many times.

```{r}
#| label: task4-grand-plot
#| fig-cap: "Grand benchmark across all tasks. The DP optimization overhead is invisible at this scale."
#| fig-height: 7

bm_plot <- bm_grand |>
  mutate(
    expression = as.character(expression),
    median_us  = as.numeric(median) * 1e6,
    task = case_when(
      grepl("Task 2", expression) ~ "Task 2",
      grepl("Task 3", expression) ~ "Task 3",
      grepl("DP|resource|parallel", expression) ~ "Task 4",
      TRUE ~ "Task 1"
    )
  )

ggplot(bm_plot, aes(x = reorder(expression, median_us),
                     y = median_us, fill = task)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = if_else(
    median_us > 1000,
    paste0(round(median_us / 1000, 1), " ms"),
    paste0(round(median_us), " \u00b5s")
  )), hjust = -0.1, size = 3.3, fontface = "bold") +
  coord_flip(clip = "off") +
  scale_fill_manual(
    values = c("Task 1" = palette_main[3], "Task 2" = palette_main[4],
               "Task 3" = palette_main[2], "Task 4" = palette_main[1]),
    name = "Task"
  ) +
  scale_y_log10(
    labels = function(x) if_else(x >= 1000, paste0(x/1000, " ms"), paste0(x, " \u00b5s")),
    expand = expansion(mult = c(0, 0.35))
  ) +
  labs(
    title    = "The Full Picture: Tasks 1 through 4",
    subtitle = "Log scale. Optimization and execution are both included for Task 4.",
    x = NULL, y = "Median execution time (log scale)"
  )
```

## How fast is the optimizer itself? Scaling to larger chains

For our 5-matrix chain, the DP overhead is negligible. But what happens as $n$ grows? This matters because in real applications (e.g., deep learning, scientific simulations), chains of 20, 50, or even hundreds of matrices are common.

We benchmark the four DP implementations (R, C++ standard, C++ parallel, C++ resource-aware) on chains of increasing length with randomly generated bottleneck dimensions. Only the **dimension vectors** matter here; the optimizer does not need actual matrix entries, so this section isolates planning cost cleanly.

```{r}
#| label: task4-scaling

set.seed(123)
scaling_results <- list()

for (n in c(5, 10, 20, 50, 100, 200)) {
  # Random bottleneck dimensions: alternate large/small
  test_dims <- integer(n + 1)
  for (i in seq_along(test_dims)) {
    test_dims[i] <- if (i %% 2 == 1) sample(100:1000, 1) else sample(2:20, 1)
  }
  
  bm_s <- bench::mark(
    `R DP`               = mcm_dp_r(test_dims),
    `C++ DP`             = mcm_dp_cpp(test_dims),
    `C++ parallel DP`    = mcm_dp_parallel(
      test_dims,
      n_threads = parallel_threads,
      min_cells_parallel = parallel_min_cells
    ),
    `C++ resource-aware` = mcm_dp_resource_aware(test_dims),
    check = FALSE,
    min_time = 2,
    min_iterations = 50,
    filter_gc = FALSE,
    max_iterations = 100
  )
  
  scaling_results[[as.character(n)]] <- bm_s |>
    mutate(
      expression = as.character(expression),
      n = n,
      sd_time = vapply(time, function(x) sd(as.numeric(x)), numeric(1)),
      mean_time = vapply(time, function(x) mean(as.numeric(x)), numeric(1))
    ) |>
    select(expression, n, median, sd_time, mean_time)
}

scale_df <- bind_rows(scaling_results) |>
  mutate(
    median_us = as.numeric(median) * 1e6,
    cv = as.numeric(sd_time) / as.numeric(mean_time)
  )
```

```{r}
#| label: task4-scaling-table

scale_table <- scale_df |>
  mutate(time_label = case_when(
    median_us < 1000 ~ paste0(round(median_us, 1), " \u00b5s"),
    median_us < 1e6  ~ paste0(round(median_us / 1000, 2), " ms"),
    TRUE             ~ paste0(round(median_us / 1e6, 2), " s")
  )) |>
  select(expression, n, time_label) |>
  pivot_wider(names_from = expression, values_from = time_label)

scale_cv_table <- scale_df |>
  mutate(cv_label = round(cv, 3)) |>
  select(expression, n, cv_label) |>
  pivot_wider(names_from = expression, values_from = cv_label) |>
  rename_with(~ paste0(.x, " CV"), -n)

scale_table <- scale_table |>
  left_join(scale_cv_table, by = "n")

# Dynamic crossover detection: find where parallel first beats sequential C++ DP
crossover_df <- scale_df |>
  filter(expression %in% c("C++ DP", "C++ parallel DP")) |>
  select(n, expression, median_us) |>
  pivot_wider(names_from = expression, values_from = median_us) |>
  rename(seq_us = `C++ DP`, par_us = `C++ parallel DP`) |>
  arrange(n) |>
  mutate(par_wins = par_us < seq_us)

# Count sign flips to detect monotone vs non-monotone crossover
flips <- crossover_df |>
  mutate(prev_par_wins = lag(par_wins)) |>
  filter(!is.na(prev_par_wins), par_wins != prev_par_wins)

n_flips <- nrow(flips)

if (n_flips == 0) {
  # parallel never beats sequential (or always beats — check which)
  crossover_case <- if (all(!crossover_df$par_wins)) "never" else "always"
} else if (n_flips == 1) {
  crossover_case <- "clean"
  cross_n_hi   <- flips$n[1]
  cross_n_lo   <- crossover_df$n[which(crossover_df$n == cross_n_hi) - 1]
  cross_seq_lo <- crossover_df$seq_us[crossover_df$n == cross_n_lo]
  cross_par_lo <- crossover_df$par_us[crossover_df$n == cross_n_lo]
  cross_seq_hi <- crossover_df$seq_us[crossover_df$n == cross_n_hi]
  cross_par_hi <- crossover_df$par_us[crossover_df$n == cross_n_hi]
} else {
  crossover_case <- "nonmonotone"
  # summarise: which n values does parallel actually win?
  par_wins_at <- crossover_df$n[crossover_df$par_wins]
  par_loses_at <- crossover_df$n[!crossover_df$par_wins]
}

parallel_crossover_text <- paste0(
  "For this render, the wavefront implementation has a lower median at n = ",
  if (any(crossover_df$par_wins)) paste(crossover_df$n[crossover_df$par_wins], collapse = ", ") else "none of the measured sizes",
  ". It has an equal or higher median at n = ",
  if (any(!crossover_df$par_wins)) paste(crossover_df$n[!crossover_df$par_wins], collapse = ", ") else "none of the measured sizes",
  ". This is a timing comparison; actual threading also requires the reported OpenMP status and diagonal threshold."
)

scale_table |>
  kable(caption = "DP optimization time and CV by chain length and algorithm",
        col.names = c("n", "R DP", "C++ DP", "C++ parallel", "C++ resource-aware",
                      "R DP CV", "C++ DP CV", "C++ parallel CV", "C++ resource-aware CV"),
        align = "ccccccccc")
```

At $n = 5$, all variants are effectively free. But the gap widens quickly:

- By $n = 200$, the table shows that the compiled DP implementations are much faster than the R version. In this render, `R DP` takes `r scale_table[scale_table$n == 200, "R DP"][[1]]` while `C++ DP` takes `r scale_table[scale_table$n == 200, "C++ DP"][[1]]`, which is about a `r fmt_ratio(scale_df$median_us[scale_df$expression == "R DP" & scale_df$n == 200] / scale_df$median_us[scale_df$expression == "C++ DP" & scale_df$n == 200])` speedup.
- The parallel wavefront version now uses the same OpenMP configuration reported above. At small $n$, thread-management overhead makes it slower than the sequential C++ DP: each diagonal is narrow, so the work available per thread is tiny. `r parallel_crossover_text` The scaling table does not extend beyond $n = `r max(scale_df$n)`$, so the exact behaviour for larger chains is not measured here.
- The resource-aware variant evaluates a richer objective at each split; its measured time is shown in the table. That extra work is the price of modeling hardware effects rather than counting FLOPs alone.

### A direct parallel benchmark

The scaling table mixes several effects at once: language, objective function, and thread count. To isolate the OpenMP effect, I benchmark the **same wavefront implementation** twice on a larger optimizer-only problem: once with one thread, once with the thread count selected above. The only thing that changes is the number of threads.

```{r}
#| label: task4-parallel-demo

set.seed(2026)
parallel_demo_n <- 320L
parallel_demo_dims <- integer(parallel_demo_n + 1L)

for (i in seq_along(parallel_demo_dims)) {
  parallel_demo_dims[i] <- if (i %% 2 == 1) sample(150:1200, 1) else sample(2:30, 1)
}

if (isTRUE(openmp_info$enabled) && parallel_threads > 1L) {
  invisible(mcm_dp_parallel(
    parallel_demo_dims,
    n_threads = 1L,
    min_cells_parallel = parallel_min_cells
  ))
  invisible(mcm_dp_parallel(
    parallel_demo_dims,
    n_threads = parallel_threads,
    min_cells_parallel = parallel_min_cells
  ))
  gc()

  bm_parallel_demo <- bench::mark(
    `Wavefront DP (1 thread)` = mcm_dp_parallel(
      parallel_demo_dims,
      n_threads = 1L,
      min_cells_parallel = parallel_min_cells
    ),
    `Wavefront DP (parallel)` = mcm_dp_parallel(
      parallel_demo_dims,
      n_threads = parallel_threads,
      min_cells_parallel = parallel_min_cells
    ),
    check = FALSE,
    min_time = 2,
    min_iterations = 50,
    filter_gc = FALSE,
    max_iterations = 100
  )

  bm_parallel_demo |>
    mutate(
      expression = as.character(expression),
      `CV (sd/mean)` = round(bench_cv(bm_parallel_demo), 3)
    ) |>
    select(expression, min, median, `CV (sd/mean)`, mem_alloc, n_itr) |>
    kable(
      caption = "Direct OpenMP benchmark: same wavefront DP, one thread vs parallel",
      align = "lccccc"
    )
} else {
  tibble(
    Message = "OpenMP is not enabled in this render, so the wavefront benchmark falls back to one thread. Use Docker to reproduce the threaded benchmark."
  ) |>
    kable(caption = "Direct OpenMP benchmark status", align = "l")
}
```

This benchmark isolates the wavefront planner with one thread versus the reported thread count. It establishes the measured timing difference for this input; it does not show a benefit for the five-matrix end-to-end problem.

```{r}
#| label: task4-scaling-plot
#| fig-cap: "DP optimization time vs chain length. Both axes are log-scaled, showing the expected cubic growth."

ggplot(scale_df, aes(x = n, y = median_us, color = expression)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_log10() +
  scale_y_log10(labels = function(x) {
    case_when(
      x < 1000 ~ paste0(round(x), " \u00b5s"),
      x < 1e6  ~ paste0(round(x / 1000, 1), " ms"),
      TRUE     ~ paste0(round(x / 1e6, 2), " s")
    )
  }) +
  scale_color_manual(values = c(
    "R DP"               = palette_main[3],
    "C++ DP"             = palette_main[1],
    "C++ parallel DP"    = palette_main[2],
    "C++ resource-aware" = palette_main[4]
  ), name = NULL) +
  labs(
    title    = "Optimizer Scaling: O(n\u00b3) in Action",
    subtitle = "Log-log plot. Measured times for algorithms with cubic worst-case work.",
    x = "Chain length (n)", y = "Median optimization time"
  )
```

The plot shows measured planning time by chain length. Cubic asymptotic work does not imply a measured slope of exactly three over this finite range, particularly when overhead dominates small inputs.

## Beyond Godbole: a brief note on $O(n \log n)$ methods

For completeness, it is worth mentioning that the $O(n^3)$ dynamic programming approach is not the theoretical limit. Hu and Shing (1982, 1984) showed that the MCOP is isomorphic to the **minimum-weight triangulation of a convex polygon**, and exploited this geometric structure to achieve $O(n \log n)$ time. The key idea is that each parenthesization corresponds to a triangulation: the $n$ matrices map to the sides of an $(n+1)$-sided polygon, and each multiplication corresponds to a triangle with cost equal to the product of its three vertex weights.

Hu and Shing give an asymptotically faster approach. Schwartz and Weiss (2019) revisit its correctness proof. Neither algorithm is implemented here, so this experiment establishes no practical crossover size or performance comparison with them.

## Discussion

Task 4 brings together all the pieces:

1. **The standard DP minimizes scalar multiplication count.** It agrees with exhaustive enumeration on the tested small chains; its cubic work and quadratic storage limit scale.

2. **The full pipeline includes planning overhead.** The grand benchmark reports its measured cost alongside hand-parenthesized R and precomputed RPN execution.

3. **Parallel gains are empirical.** The dedicated benchmark compares the same wavefront implementation with different thread counts, while the scaling table compares separate algorithms.

4. **The resource-aware cost is an illustrative surrogate.** The two-tier penalties and dimension-dependent small-product term are not calibrated hardware measurements. Every full binary plan has exactly $n-1$ multiplications, so a constant penalty per multiplication alone could not change the chosen plan.

The overall conclusion across all four tasks:

```{r}
#| label: task4-overall-summary

grand_speedup <- function(label) {
  as.numeric(bench_value(bm_grand, "R naive", "median")) /
    as.numeric(bench_value(bm_grand, label, "median"))
}

overall_summary <- tibble(
  `What we did` = c(
    "Hand-parenthesized R (Task 1)",
    "C++ with same naive ordering (Task 2)",
    "C++ stack machine + pre-computed optimal (Task 3)",
    "C++ DP optimizer + stack machine (Task 4)"
  ),
  `Relative to R naive` = c(
    fmt_relative_perf(grand_speedup("R optimal (hand)")),
    fmt_relative_perf(grand_speedup("C++ naive (Task 2)")),
    fmt_relative_perf(grand_speedup("C++ RPN optimal (Task 3)")),
    fmt_relative_perf(grand_speedup("C++ DP + execute"))
  )
)

overall_summary |>
  kable(align = "lr", caption = "Overall relative performance derived from the grand benchmark")
```

:::{.callout-note}
## Sanity check on the results

The benchmark patterns are internally consistent with the theory developed in the report:

1. Whenever two strategies are compared **inside the same benchmark table**, the good parenthesization beats the naive one by roughly the factor suggested by the scalar-operation counts.
2. The naive R and naive C++ versions stay in the same broad performance region, which is expected because both ultimately spend most of their time inside the same BLAS kernels once the multiplication order is fixed.
3. In the optimization-only microbenchmark, planning takes microseconds while executing dense matrix products takes hundreds of microseconds to milliseconds, so the optimizer is not the bottleneck for the problem sizes studied here.

That is why the cross-section timing drift is acceptable: the absolute medians move a bit across benchmark sessions, but the qualitative conclusions remain stable.
:::

```{r}
#| label: task4-t3-vs-t4-text
#| include: false

t3 <- as.numeric(bench_value(bm_grand, "C++ RPN optimal (Task 3)", "median"))
t4 <- as.numeric(bench_value(bm_grand, "C++ DP + execute", "median"))
dp <- as.numeric(bench_value(bm_decompose, "C++ DP (optimize only)", "median"))

t3_vs_t4_text <- if (t3 < t4) {
  paste0(
    "The same logic explains why `C++ RPN optimal (Task 3)` (", fmt_time(t3),
    ") edges out `C++ DP + execute` (", fmt_time(t4),
    ") in the overall summary table: Task 3 passes a pre-computed RPN vector with zero optimization cost at benchmark time, while Task 4 folds in the ",
    fmt_time(dp), " DP call on top of the same executor."
  )
} else {
  paste0(
    "In this render `C++ DP + execute` (", fmt_time(t4),
    ") is marginally faster than `C++ RPN optimal (Task 3)` (", fmt_time(t3),
    "). The difference (", fmt_time(abs(t4 - t3)),
    ") is within measurement noise — both pipelines execute the same BLAS calls, and the ",
    fmt_time(dp), " DP overhead is smaller than the run-to-run jitter at these timescales."
  )
}
```

:::{.callout-note}
## Technical Note: the "Hand-Parenthesization Paradox"

The hand-parenthesized baseline takes `r fmt_time(bench_value(bm_grand, "R optimal (hand)", "median"))`, versus `r fmt_time(bench_value(bm_grand, "C++ DP + execute", "median"))` for the automated pipeline. Both use the selected multiplication order; their allocation, input-copying and planning work differ.

The decomposition benchmark separates planning from execution; it does not isolate every allocation or scheduling effect.

The real victory of Task 4 is **scalability**. While hand-tuning is feasible for 5 matrices, it is impractical for 100. The C++ implementation gets extremely close to the hand-written optimum while providing an automated implementation with input validation and explicit computational limits.
:::

The message is consistent: **algorithm first, language second**. Reordering the chain is where the large win comes from; C++ then helps us package that win into an automated and scalable implementation.

---

# Limitations and Future Improvements {.unnumbered}

No solution is complete without honest self-assessment. Here are the main limitations of this implementation and how they could be addressed in a production setting:

**Owned input copies at the R/C++ boundary.** The temporary Armadillo wrappers view R memory, but inserting them into the owning vector and execution stack copies matrix data. The executor is not zero-copy; native allocations require a process-level profiler to measure.

**Thread overhead in the parallel wavefront DP.** The work inside each DP cell is still just a few scalar operations and comparisons, so this is not the kind of kernel that scales like `dgemm`. Even with the lower `min_cells_parallel = 64` threshold used in the report, the threaded version only starts to help on wider diagonals, and the speedup remains modest on laptop-sized benchmarks. A more adaptive strategy would calibrate the threshold at startup and potentially switch between sequential, OpenMP, and task-based runtimes depending on the problem size.

**Iterative RPN reconstruction.** An explicit work stack avoids recursive reconstruction in C++. This does not make arbitrary chain lengths feasible: DP still requires quadratic storage and cubic computation, and allocations can fail.

**Simplified surrogate cost.** Cache-tier and small-product penalties are design assumptions. They are not the processor-allocation algorithm in Nissim et al. (2024), nor a calibrated implementation of Hong and Kung’s I/O analysis. Hardware counters and held-out execution benchmarks would be needed to validate a performance model.

**Numeric range.** Input dimensions must be positive finite integers within the int32 range. Products use doubles; integer costs above $2^{53}$ may lose exactness. Table dimensions are bounded to avoid signed indexing overflow, but large valid requests can still exhaust memory.

**Mixed precision opportunity.** Lower precision changes rounding behavior and may improve throughput on suitable hardware. No fixed speedup is established here; the current analysis uses double precision.

**Benchmark methodology.** The benchmarks measure warm-cache performance (repeated iterations with `bench::mark`). Because each section uses its own benchmark session, absolute medians for the same technique can drift slightly across tables. Cold-start behavior (first execution after data load) can also differ significantly due to TLB misses and cache warming. For a fully rigorous comparison, one would alternate configurations within the same benchmark loop to equalize cache state, or use hardware performance counters (`perf stat` on Linux) to measure L1/L2/L3 miss rates directly.

---

# References {.unnumbered}

1. Cormen, T. H., Leiserson, C. E., Rivest, R. L., & Stein, C. (2009). *Introduction to Algorithms* (3rd ed.). MIT Press. [Publisher page](https://mitpress.mit.edu/9780262033848/introduction-to-algorithms/)

2. Godbole, S. S. (1973). On efficient computation of matrix chain products. *IEEE Transactions on Computers, C-22*(9), 864--866. [https://doi.org/10.1109/TC.1973.5009182](https://doi.org/10.1109/TC.1973.5009182)

3. Hu, T. C., & Shing, M. T. (1982). Computation of matrix chain products. Part I. *SIAM Journal on Computing, 11*(2), 362--373. [https://doi.org/10.1137/0211028](https://doi.org/10.1137/0211028)

4. Hu, T. C., & Shing, M. T. (1984). Computation of matrix chain products. Part II. *SIAM Journal on Computing, 13*(2), 228--251. [https://doi.org/10.1137/0213017](https://doi.org/10.1137/0213017)

5. Schwartz, O., & Weiss, E. (2019). Revisiting "Computation of matrix chain products". *SIAM Journal on Computing, 48*(5), 1481--1486. [https://doi.org/10.1137/18M1195401](https://doi.org/10.1137/18M1195401)

6. Nissim, R., Schwartz, O., & Shabo, R. (2024). *Challenges in parallel matrix chain multiplication*. JSSPP 2024. [Author PDF](https://www.cs.huji.ac.il/~odedsc/papers/JSSPP24-Chain-MM.pdf)

7. Eddelbuettel, D., & Sanderson, C. (2014). RcppArmadillo: Accelerating R with high-performance C++ linear algebra. *Computational Statistics & Data Analysis, 71*, 1054--1063. [https://doi.org/10.1016/j.csda.2013.02.005](https://doi.org/10.1016/j.csda.2013.02.005). Package vignette: [RcppArmadillo introduction](https://cran.r-project.org/package=RcppArmadillo/vignettes/RcppArmadillo-intro.pdf)

8. Hong, J. W., & Kung, H. T. (1981). I/O complexity: The red-blue pebble game. *Proceedings of the 13th Annual ACM Symposium on Theory of Computing*, 326--333. [https://doi.org/10.1145/800076.802486](https://doi.org/10.1145/800076.802486)

9. Matrix chain multiplication. (2026). In *Wikipedia*. Retrieved February 14, 2026, from [https://en.wikipedia.org/wiki/Matrix_chain_multiplication](https://en.wikipedia.org/wiki/Matrix_chain_multiplication)

```{r}
#| label: audit-export
#| include: false
audit_output <- Sys.getenv("MATRIX_AUDIT_OUTPUT", unset = "")
if (nzchar(audit_output)) saveRDS(list(plans = paren_df, grand = as.data.frame(bm_grand)[, c("expression", "median", "mem_alloc")], scale = scale_df, openmp = openmp_info, threads = parallel_threads, surrogate = r3), audit_output)
```
