---
title: "Probabilistic Cancer Classification via RNA-Seq Data"
description: "Multiclass classification of tumor types using RNA-Seq data and probabilistic learning models (LDA, QDA, Naive Bayes)."
author: "Alejandro Treny"
date: last-modified
format:
  html:
    output-file: index.html
    toc: true
    toc-location: left
    code-fold: show
    theme: cosmo
    embed-resources: false
execute:
  warning: true
  message: false
editor: visual
---

# Introduction

This project focuses on applying **Statistical Learning** techniques for the multiclass classification of different cancer types using gene expression data (**RNA-Seq**). The ability to distinguish between tumor subtypes at the molecular level is fundamental for precision medicine, enabling more accurate diagnoses and personalized treatments.

## The Dataset

We utilize an extraction of the **TCGA Pan-Cancer (PANCAN)** dataset, available at the UCI Machine Learning Repository. This dataset consists of gene expression levels measured using the Illumina HiSeq platform.

### Key Characteristics:

-   **Samples (Instances):** 801 individuals, all diagnosed with a type of tumor.
-   **Variables (Genes):** 20,531 dimensions, posing the challenge of the "curse of dimensionality."
-   **Classes (Labels):** 5 distinct tumor types:
    -   **BRCA:** Breast Invasive Carcinoma.
    -   **KIRC:** Kidney Renal Clear Cell Carcinoma.
    -   **COAD:** Colon Adenocarcinoma.
    -   **LUAD:** Lung Adenocarcinoma.
    -   **PRAD:** Prostate Adenocarcinoma.

## Project Objective

The primary goal is not to identify the genetic causes of cancer (comparing healthy vs. sick individuals), but rather to develop a model capable of **correctly classifying the origin of tumor tissue** based solely on its gene expression profile.

## Methodology and Models

To achieve a robust classification, we will adhere to the **Probabilistic Learning** framework covered in the course curriculum. Specifically, we will implement and compare the following models suitable for **multiclass classification**:

1.  **Linear Discriminant Analysis (LDA):** A generative method assuming Gaussian distributions with a common covariance matrix.
2.  **Quadratic Discriminant Analysis (QDA):** A more flexible extension of LDA allowing for class-specific covariance matrices.
3.  **Naive Bayes:** A probabilistic classifier assuming conditional independence of the retained PC features given the class.
4.  **Multinomial Logistic Regression:** An extension of the standard binary logistic regression for multiple classes (used as a baseline).

### Statistical Challenge: High Dimensionality ($p \gg n$)

Our dataset contains **20,531 genes** ($p$) and **801 samples** ($n$). With $p > n$, the unregularized covariance estimates used by classical LDA and QDA are singular. Reducing dimension or regularizing these estimates is necessary; this does not make all high-dimensional classifiers impossible.

This analysis uses **PCA** to reduce dimension. Genes that are constant in the relevant training partition are removed before scaling. Nonzero-variance genes are centered and standardized; the test partition and each validation fold use transformations fitted on their corresponding training observations.

### Evaluation Framework

Following the best practices for **Model Evaluation**, we will rigorously validate our results using:

-   **Stratified Train/Test Split (80/20):** PCA is fitted exclusively on the training set, and the test set is projected independently to prevent data leakage.
-   **10-Fold Cross-Validation:** Applied within the training set to tune the number of principal components and estimate validation performance.
-   **Confusion Matrices:** To verify performance across all 5 cancer types on the held-out test set.
-   **Result Checks:** A shuffled-label diagnostic and train/test integrity assertions.
-   **Metrics:** We will use **Accuracy** and **Kappa** (to account for random chance agreement) as our primary metrics.

------------------------------------------------------------------------

# Exploratory Data Analysis (EDA)

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(DT)
library(caret)
```

## Data Loading

```{r}
#| label: load-data

# Load the local CSV files
# Use read_csv for faster loading of large datasets
data_dir <- Sys.getenv("RNA_SEQ_DATA_DIR", unset = ".")
data_raw <- read_csv(file.path(data_dir, "data.csv"), show_col_types = FALSE)
labels_raw <- read_csv(file.path(data_dir, "labels.csv"), show_col_types = FALSE)
stopifnot(!anyDuplicated(data_raw[[1]]), !anyDuplicated(labels_raw[[1]]),
          !anyNA(data_raw[[1]]), !anyNA(labels_raw[[1]]),
          setequal(data_raw[[1]], labels_raw[[1]]))

# Display dimensions as a proper table
tibble(
  Dataset = c("Gene Expression (data)", "Labels"),
  Rows = c(nrow(data_raw), nrow(labels_raw)),
  Columns = c(ncol(data_raw), ncol(labels_raw))
) |>
  datatable(
    options = list(dom = "t", ordering = FALSE),
    rownames = FALSE,
    caption = "Dimensions of loaded datasets"
  )
```

The expression matrix contains `r nrow(data_raw)` samples and `r ncol(data_raw)` columns, including the sample identifier. The loader checks unique, nonmissing IDs and equality of the two ID sets before a one-to-one join; matching row counts alone would not establish correspondence.

## Preprocessing

```{r}
#| label: merge-data

# Merge the labels (Class) with the gene expression data
# Both files likely have the sample ID as the first column, often named '...1' by default
full_data <- inner_join(labels_raw, data_raw, by = c("...1" = "...1"), relationship = "one-to-one") %>%
    rename(sample_id = ...1) %>%
    mutate(Class = as.factor(Class))
stopifnot(!anyNA(full_data$Class), all(is.finite(as.matrix(select(full_data, starts_with("gene_"))))))

# Fix the split before looking at expression distributions or class associations.
set.seed(42)
train_idx <- createDataPartition(full_data$Class, p = 0.8, list = FALSE)
train_data <- full_data[train_idx, ]
test_data <- full_data[-train_idx, ]

# Quick check of the resulting dataset structure
# Display the first 10 columns of the first 20 rows using an interactive table
datatable(
    full_data %>% select(1:10) %>% head(20),
    options = list(scrollX = TRUE, pageLength = 5),
    caption = "Preview of the first 10 columns (Sample ID, Class + 8 Genes)"
)

```

We join expression profiles and labels by sample ID and fix the stratified split immediately. The preview and class counts describe the cohort; all subsequent feature distributions, correlations and ANOVA summaries use training rows only.

## Class Distribution

```{r}
#| label: plot-class-distribution

# Visualization of class distribution
full_data |>
  count(Class) |>
  mutate(pct = round(n / sum(n) * 100, 1),
         label = paste0(n, " (", pct, "%)")) |>
  ggplot(aes(x = reorder(Class, n), y = n, fill = Class)) +
  geom_col() +
  geom_text(aes(label = label), hjust = 1.1, color = "white", fontface = "bold", size = 4) +
  coord_flip() +
  labs(
    title = "Cancer Class Distribution",
    x = "Cancer Type", y = "Number of samples",
    subtitle = "PANCAN Dataset (UCI)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The dataset exhibits a **moderate class imbalance**. BRCA (Breast Invasive Carcinoma) is the most represented class with roughly 300 samples, while COAD (Colon Adenocarcinoma) is the least represented with approximately 78 samples, nearly a 4:1 ratio. The remaining three classes (KIRC, LUAD, PRAD) sit in a similar mid-range around 130-140 samples each.

This imbalance is worth noting because:

-   Models like **LDA** and **Logistic Regression** may be biased toward predicting BRCA more frequently.
-   Evaluation metrics like raw **Accuracy** can be misleading, a model could achieve decent accuracy by simply favoring the majority class. This motivates our use of **Kappa** and per-class metrics from the **Confusion Matrix**.
-   This study retains the observed class proportions; it does not evaluate resampling techniques such as SMOTE.

## Missing Values Check

```{r}
#| label: eda-missing-values

gene_cols <- train_data |> select(starts_with("gene_"))

tibble(
  `Total Cells` = scales::comma(prod(dim(gene_cols))),
  `Missing Values` = sum(is.na(gene_cols)),
  `% Missing` = round(sum(is.na(gene_cols)) / prod(dim(gene_cols)) * 100, 4)
) |>
  datatable(
    options = list(dom = "t", ordering = FALSE),
    rownames = FALSE,
    caption = "Missing values in training gene expression columns"
  )
```

The table reports missing values in the training expression matrix. The loader also rejects nonfinite gene values in the full input; missing-value checks alone do not establish broader data quality or biological validity.

## Global Gene Expression Distribution

```{r}
#| label: eda-global-distribution

set.seed(42)
sample_genes <- sample(names(gene_cols), 500)

train_data |>
  select(Class, all_of(sample_genes)) |>
  pivot_longer(-Class, names_to = "gene", values_to = "expression") |>
  ggplot(aes(x = expression)) +
  geom_histogram(bins = 80, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Global Gene Expression Distribution (500 random genes)",
    x = "Expression Level", y = "Count"
  ) +
  theme_minimal()
```

The distribution reveals a clear **bimodal pattern**:

-   A large spike at **zero or near-zero** expression in this transformed measurement matrix. This plot alone cannot distinguish biological silencing from measurement or preprocessing effects.
-   A second, broader peak centered around **expression level 8 to 10**, representing the actively expressed genes, roughly following a Gaussian shape (consistent with log-transformed RNA-Seq counts).

A low expression value does not imply that a gene is uninformative. The analysis removes only genes with zero training variance; it does not select features from these exploratory plots.

## Mean Expression by Cancer Type

```{r}
#| label: eda-expression-by-class

train_data |>
  mutate(mean_expr = rowMeans(pick(starts_with("gene_")), na.rm = TRUE)) |>
  ggplot(aes(x = Class, y = mean_expr, fill = Class)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Mean Gene Expression per Sample by Cancer Type",
    x = "Cancer Type", y = "Mean Expression"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The boxplots compare average expression per training sample. Overlap of these summaries does not rule out discrimination by individual genes, and the figure does not establish a biological mechanism.

## Gene Variance Distribution

```{r}
#| label: eda-gene-variance

gene_variances <- gene_cols |>
  summarise(across(everything(), \(x) var(x, na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "gene", values_to = "variance") |>
  arrange(desc(variance))

ggplot(gene_variances, aes(x = variance + 1e-6)) +
  geom_histogram(bins = 100, fill = "darkorange", alpha = 0.7) +
  scale_x_log10() +
  labs(
    title = "Distribution of Gene Variances (log scale)",
    x = "Variance (log10)", y = "Count",
    subtitle = "Raw variance alone does not measure predictive value"
  ) +
  theme_minimal()
```

There are **`r sum(gene_variances$variance == 0)` genes with zero training variance** and **`r sum(gene_variances$variance < 1)` below 1**. Raw variance does not measure predictive value: every retained gene is standardized to unit training variance before PCA, so low raw variance does not imply a negligible loading.

## Correlation Among Top Variable Genes

```{r}
#| label: eda-correlation-top-genes

library(pheatmap)

top_genes <- gene_variances |> slice_head(n = 30) |> pull(gene)

cor_matrix <- train_data |>
  select(all_of(top_genes)) |>
  cor(use = "pairwise.complete.obs")

pheatmap(
  cor_matrix,
  main = "Correlation Among Top 30 Most Variable Genes",
  fontsize = 7,
  color = colorRampPalette(c("blue", "white", "red"))(100)
)
```

The heatmap summarizes pairwise correlations among the 30 genes with largest training variance. Correlation alone does not identify shared pathways or biological causes. PCA gives uncorrelated coordinates on the fitted training sample, not necessarily within each cancer class.

## Top Differentially Expressed Genes

```{r}
#| label: eda-top-genes-anova

# One-way ANOVA F-statistic per gene to identify class-discriminative genes
f_stats <- map_dbl(names(gene_cols), \(g) {
  tryCatch({
    summary(aov(gene_cols[[g]] ~ train_data$Class))[[1]][1, "F value"]
  }, error = \(e) NA_real_)
})

top_diff_genes <- tibble(gene = names(gene_cols), F_stat = f_stats) |>
  arrange(desc(F_stat)) |>
  slice_head(n = 12)

train_data |>
  select(Class, all_of(top_diff_genes$gene)) |>
  pivot_longer(-Class, names_to = "gene", values_to = "expression") |>
  ggplot(aes(x = Class, y = expression, fill = Class)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_wrap(~gene, scales = "free_y", ncol = 4) +
  labs(title = "Top 12 Differentially Expressed Genes (by ANOVA F-stat)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

The plot ranks training genes by their one-way ANOVA F-statistic. This is an exploratory comparison, without multiplicity-adjusted inference or gene annotation. A large F-statistic does not guarantee a large PCA loading, because PCA does not use the labels. These rankings are not used to select classifier inputs.

------------------------------------------------------------------------

# Train/Test Split

The stratified 80/20 split was fixed before feature EDA. Its counts are shown below; preprocessing for cross-validation is fitted separately within each analysis fold.

```{r}
#| label: train-test-split

library(caret)

# The fixed split was created before feature EDA; display its counts here.

tibble(
  Set = c("Training", "Test"),
  Samples = c(nrow(train_data), nrow(test_data))
) |>
  bind_cols(
    bind_rows(
      train_data |> count(Class) |> pivot_wider(names_from = Class, values_from = n),
      test_data  |> count(Class) |> pivot_wider(names_from = Class, values_from = n)
    )
  ) |>
  datatable(
    options = list(dom = "t", ordering = FALSE, scrollX = TRUE),
    rownames = FALSE,
    caption = "Train/Test split with class distribution"
  )
```

The stratified split preserves the relative proportions of each cancer type in both sets. The training set will be used for PCA fitting, model training, and 10-fold cross-validation. The test set is held out entirely until the final evaluation step.

------------------------------------------------------------------------

# Principal Component Analysis (PCA)

The raw gene space has more dimensions than observations. PCA provides the dimension reduction used here for the classical discriminant models; regularized high-dimensional alternatives, including logistic regression, remain possible.

Crucially, PCA is fitted **only on the training set** to avoid data leakage. The test set is then projected onto the same principal components.

```{r}
#| label: pca-compute
#| output: false

library(factoextra)

# Remove zero-variance genes (computed on training set)
train_gene_cols <- train_data |> dplyr::select(starts_with("gene_"))
train_var <- apply(train_gene_cols, 2, var)
nonzero_var_genes <- names(train_var[train_var > 0])

# Fit PCA on training data only
pca_train <- train_data |> dplyr::select(all_of(nonzero_var_genes))
pca_result <- prcomp(pca_train, center = TRUE, scale. = TRUE)

# Project test data using the training PCA
pca_test_scores <- predict(pca_result, newdata = test_data |> dplyr::select(all_of(nonzero_var_genes)))
```

## Scree Plot

```{r}
#| label: pca-scree

pct_var <- (pca_result$sdev^2 / sum(pca_result$sdev^2) * 100)[1:30]

tibble(PC = factor(1:30, levels = 1:30), pct = pct_var) |>
  ggplot(aes(x = PC, y = pct)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_line(aes(x = as.numeric(PC), y = pct), linewidth = 0.5) +
  geom_point(aes(x = as.numeric(PC), y = pct), size = 1.5) +
  geom_text(
    data = tibble(PC = factor(1:10, levels = 1:30),
                  pct = pct_var[1:10]),
    aes(label = paste0(round(pct, 1), "%")),
    vjust = -0.8, hjust = -0.1, size = 3
  ) +
  labs(
    title = "Scree Plot: Variance Explained by Principal Components",
    x = "Principal Component", y = "Percentage of Explained Variance"
  ) +
  theme_minimal()
```

The first three training components explain **`r sprintf("%.1f", pct_var[1])`%**, **`r sprintf("%.1f", pct_var[2])`%** and **`r sprintf("%.1f", pct_var[3])`%** of standardized training variance. Explained variance measures reconstruction, not classification accuracy.

## Cumulative Variance Explained

```{r}
#| label: pca-cumulative-variance

cumvar <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2))
pc_cutoffs <- vapply(c(0.80, 0.90, 0.95), function(target) which(cumvar >= target)[1], integer(1))

tibble(PC = seq_along(cumvar), cumulative_variance = cumvar) |>
  filter(PC <= 500) |>
  ggplot(aes(x = PC, y = cumulative_variance)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = c(0.80, 0.90, 0.95), linetype = "dashed", color = "grey50") +
  annotate("text", x = 490, y = c(0.81, 0.91, 0.96),
           label = c("80%", "90%", "95%"), size = 3.5, hjust = 1) +
  geom_vline(xintercept = pc_cutoffs, linetype = "dotted", color = "red", alpha = 0.5) +
  labs(
    title = "Cumulative Variance Explained by Principal Components",
    x = "Number of Components", y = "Cumulative Proportion"
  ) +
  theme_minimal()
```

The training PCA needs **`r pc_cutoffs[1]`**, **`r pc_cutoffs[2]`** and **`r pc_cutoffs[3]`** components to reach 80%, 90% and 95% of variance, respectively. The classification grid below is fixed explicitly and selected using training cross-validation.

## PCA Scatter Plots

```{r}
#| label: pca-scatter-12

pca_scores <- as_tibble(pca_result$x[, 1:5]) |>
  bind_cols(Class = train_data$Class)

ggplot(pca_scores, aes(x = PC1, y = PC2, color = Class)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "PCA: PC1 vs PC2 by Cancer Type (Training Set)",
    subtitle = "Axes maximize training variance without using class labels"
  ) +
  theme_minimal()
```

This projection displays the first two training PCs, with class labels added only for interpretation. The plot is descriptive; the cross-validation and held-out metrics quantify classification performance.

```{r}
#| label: pca-scatter-34

ggplot(pca_scores, aes(x = PC3, y = PC4, color = Class)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "PCA: PC3 vs PC4 by Cancer Type (Training Set)",
    subtitle = "Additional orthogonal directions of training variance"
  ) +
  theme_minimal()
```

The next two PCs display additional variation. Class overlap or separation in this plane does not determine which dimension is best for classification.

------------------------------------------------------------------------

# Model Training and Evaluation

We now train the four probabilistic classifiers outlined in the introduction:
**LDA**, **QDA**, **Naive Bayes**, and **Multinomial Logistic Regression**.
PCA, centring and scaling are fitted independently inside every resampling fold;
otherwise validation folds would influence their own feature representation.

We evaluate the fixed PC grid 5, 10, 20, 50, 100 and 159. The largest value is a candidate dimension, rather than a guaranteed variance threshold in every fold.

Note that **QDA** requires each class-specific covariance matrix to be invertible, so the number of PCs must be smaller than the size of the smallest class in the training set. We therefore limit QDA to a maximum of 50 PCs.

```{r}
#| label: model-setup
#| output: false

library(MASS)
library(nnet)
library(e1071)
library(naivebayes)

pc_counts <- c(5, 10, 20, 50, 100, 159)
x_train <- train_data |>
  dplyr::select(all_of(nonzero_var_genes))
y_train <- train_data$Class

# The x/y interface avoids expanding a 20k-gene formula.  To avoid fitting the
# same PCA 22 times per fold, each fold fits the full 159-PC transformation once
# on its analysis subset, then every classifier consumes the required prefix.
# This preserves fold isolation: validation samples never affect centring,
# scaling, zero-variance filtering, or PCA.
set.seed(42)
fold_indices <- createFolds(y_train, k = 10, returnTrain = FALSE)
max_pcs <- max(pc_counts)

fold_scores <- lapply(fold_indices, function(validation_idx) {
  analysis_idx <- setdiff(seq_len(nrow(x_train)), validation_idx)
  transformation <- preProcess(
    x_train[analysis_idx, ],
    method = c("zv", "center", "scale", "pca"),
    pcaComp = max_pcs
  )

  list(
    train_x = predict(transformation, x_train[analysis_idx, ])[, seq_len(max_pcs), drop = FALSE],
    valid_x = predict(transformation, x_train[validation_idx, ])[, seq_len(max_pcs), drop = FALSE],
    train_y = y_train[analysis_idx],
    valid_y = y_train[validation_idx]
  )
})

predict_fold <- function(model_name, fold, k) {
  train_x <- fold$train_x[, seq_len(k), drop = FALSE]
  valid_x <- fold$valid_x[, seq_len(k), drop = FALSE]

  if (model_name == "LDA") {
    predict(lda(x = train_x, grouping = fold$train_y), valid_x)$class
  } else if (model_name == "QDA") {
    predict(qda(x = train_x, grouping = fold$train_y), valid_x)$class
  } else if (model_name == "Naive Bayes") {
    fit <- naive_bayes(x = train_x, y = fold$train_y)
    predict(fit, newdata = valid_x)
  } else if (model_name == "Multinomial LogReg") {
    set.seed(42)
    train_md <- as.data.frame(train_x)
    train_md$Class <- fold$train_y
    fit <- multinom(Class ~ ., data = train_md, trace = FALSE, MaxNWts = 5000, maxit = 1000)
    stopifnot(fit$convergence == 0)
    predict(fit, newdata = as.data.frame(valid_x))
  } else {
    stop("Unsupported model: ", model_name)
  }
}

evaluate_configuration <- function(model_name, k) {
  predictions <- unlist(lapply(fold_scores, predict_fold, model_name = model_name, k = k))
  reference <- unlist(lapply(fold_scores, `[[`, "valid_y"))
  comparison <- confusionMatrix(
    factor(predictions, levels = levels(y_train)),
    factor(reference, levels = levels(y_train))
  )

  tibble(
    PCs = k,
    Model = model_name,
    Accuracy = unname(comparison$overall["Accuracy"]),
    Kappa = unname(comparison$overall["Kappa"])
  )
}

models_to_evaluate <- c("LDA", "Naive Bayes", "Multinomial LogReg")
results_list <- list()

for (k in pc_counts) {
  for (model_name in models_to_evaluate) {
    results_list[[paste(model_name, k, sep = "_")]] <- evaluate_configuration(model_name, k)
  }

  # QDA needs more samples per class than retained components.
  if (k <= 50) {
    results_list[[paste("QDA", k, sep = "_")]] <- evaluate_configuration("QDA", k)
  }
}

results_all <- bind_rows(results_list)
```

The training loop compares **22 configurations** with ten fits each (4 classifiers x 6 PC counts, minus QDA for 100 and 159 PCs). Each model is evaluated via 10-fold cross-validation **within the training set**, producing both **Accuracy** and **Kappa** as performance metrics. Kappa is particularly useful here because it adjusts for the class imbalance we observed earlier, providing an additional summary of agreement beyond the class-marginal expectation.

## Accuracy vs Number of Components

```{r}
#| label: model-accuracy-plot

results_all |>
  ggplot(aes(x = PCs, y = Accuracy, color = Model, shape = Model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = pc_counts) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1.0)) +
  labs(
    title = "Classification Accuracy vs Number of Principal Components",
    x = "Number of PCs", y = "Accuracy (10-fold CV, Training Set)"
  ) +
  theme_minimal()
```

```{r}
#| label: model-results-table

results_all |>
  mutate(
    Accuracy = scales::percent(Accuracy, accuracy = 0.1),
    Kappa = round(Kappa, 3)
  ) |>
  pivot_wider(names_from = Model, values_from = c(Accuracy, Kappa)) |>
  datatable(
    options = list(dom = "t", pageLength = 10, scrollX = TRUE),
    rownames = FALSE,
    caption = "10-fold CV results across models and PC counts"
  )
```

The table and curve above are the authoritative record for this execution. They compare model/PC configurations using only the training partition; the hold-out test set remains untouched until the next section. QDA is deliberately limited to 50 PCs because each class covariance estimate requires more observations than retained components. PCA produces orthogonal coordinates, which can improve numerical conditioning, but it does not itself prove that any model assumption is true.

## Per-Model Confusion Matrices (Test Set)

To get an honest estimate of generalization performance, we now evaluate each model on the **held-out test set**. For each classifier, we select the best PC count from cross-validation, retrain on the full training set, and predict on the test set.

```{r}
#| label: per-model-best-setup
#| output: false

# Find best PC count per model (from CV results)
best_per_model <- results_all |>
  group_by(Model) |>
  arrange(desc(Accuracy), desc(Kappa)) |>
  slice_head(n = 1) |>
  ungroup()

# Map model names to caret method names (used in CV loop above)
method_map <- c(
  "LDA" = "lda",
  "QDA" = "qda",
  "Naive Bayes" = "naive_bayes",
  "Multinomial LogReg" = "multinom"
)

# Retrain each best model on full training set, predict on test set
per_model_cms <- list()

for (i in seq_len(nrow(best_per_model))) {
  row <- best_per_model[i, ]
  k <- row$PCs
  model_name <- row$Model
  
  # Training data in PC space
  train_md <- as_tibble(pca_result$x[, 1:k]) |>
    mutate(Class = train_data$Class)
  
  # Test data in PC space (projected using training PCA)
  test_md <- as_tibble(pca_test_scores[, 1:k])
  
  set.seed(42)
  
  # Fit directly to avoid caret tuneGrid issues with method = "none"
  if (model_name == "LDA") {
    fit <- lda(Class ~ ., data = train_md)
    test_preds <- predict(fit, newdata = test_md)$class
  } else if (model_name == "QDA") {
    fit <- qda(Class ~ ., data = train_md)
    test_preds <- predict(fit, newdata = test_md)$class
  } else if (model_name == "Naive Bayes") {
    fit <- naive_bayes(Class ~ ., data = train_md)
    test_preds <- predict(fit, newdata = test_md)
  } else if (model_name == "Multinomial LogReg") {
    fit <- multinom(Class ~ ., data = train_md, trace = FALSE, MaxNWts = 5000, maxit = 1000)
    stopifnot(fit$convergence == 0)
    test_preds <- predict(fit, newdata = test_md)
  }
  
  per_model_cms[[model_name]] <- list(
    cm = confusionMatrix(test_preds, test_data$Class),
    k = k,
    name = model_name
  )
}
```

### LDA

```{r}
#| label: cm-lda

info <- per_model_cms[["LDA"]]

info$cm$table |>
  as.data.frame() |>
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = paste0("LDA Confusion Matrix (Test Set, ", info$k, " PCs)"),
    subtitle = paste0("Test Accuracy: ", scales::percent(info$cm$overall["Accuracy"], accuracy = 0.1)),
    x = "True Class", y = "Predicted Class"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The matrix and subtitle above report LDA's held-out result for the PC count selected by training-only cross-validation. Orthogonal PCA scores can improve numerical conditioning, but they do not by themselves establish LDA's equal-covariance assumption.

### QDA

```{r}
#| label: cm-qda

info <- per_model_cms[["QDA"]]

info$cm$table |>
  as.data.frame() |>
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = paste0("QDA Confusion Matrix (Test Set, ", info$k, " PCs)"),
    subtitle = paste0("Test Accuracy: ", scales::percent(info$cm$overall["Accuracy"], accuracy = 0.1)),
    x = "True Class", y = "Predicted Class"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

QDA permits one covariance matrix per class. Its held-out result is shown above for the configuration selected without using the test set; comparison with LDA should be treated as empirical for this split rather than as evidence that either covariance assumption is universally correct.

### Naive Bayes

```{r}
#| label: cm-nb

info <- per_model_cms[["Naive Bayes"]]

info$cm$table |>
  as.data.frame() |>
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = paste0("Naive Bayes Confusion Matrix (Test Set, ", info$k, " PCs)"),
    subtitle = paste0("Test Accuracy: ", scales::percent(info$cm$overall["Accuracy"], accuracy = 0.1)),
    x = "True Class", y = "Predicted Class"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Naive Bayes' held-out errors and class-specific pattern are shown in the matrix above. Its conditional-independence assumption remains an approximation even when PCA scores are uncorrelated, so this model should be judged from its validation and test metrics rather than from the visualization alone.

### Multinomial Logistic Regression

```{r}
#| label: cm-multinom

info <- per_model_cms[["Multinomial LogReg"]]

info$cm$table |>
  as.data.frame() |>
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = paste0("Multinomial LogReg Confusion Matrix (Test Set, ", info$k, " PCs)"),
    subtitle = paste0("Test Accuracy: ", scales::percent(info$cm$overall["Accuracy"], accuracy = 0.1)),
    x = "True Class", y = "Predicted Class"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Multinomial logistic regression directly estimates decision boundaries in the selected PCA space. The matrix above gives its current held-out result; it should be compared with the other methods using the same fixed test split, while recognising that a single split does not quantify deployment performance.

## Held-Out Comparison

```{r}
#| label: best-model

# Compare test-set accuracy across all four models
test_summary <- tibble(
  Model = names(per_model_cms),
  PCs = map_dbl(per_model_cms, \(x) x$k),
  `Test Accuracy` = map_dbl(per_model_cms, \(x) x$cm$overall["Accuracy"]),
  `Test Kappa` = map_dbl(per_model_cms, \(x) x$cm$overall["Kappa"])
) |>
  arrange(desc(`Test Accuracy`), desc(`Test Kappa`))

test_summary |>
  mutate(
    `Test Accuracy` = scales::percent(`Test Accuracy`, accuracy = 0.1),
    `Test Kappa` = round(`Test Kappa`, 4)
  ) |>
  datatable(
    options = list(dom = "t", ordering = FALSE),
    rownames = FALSE,
    caption = "Test set performance comparison across models (best CV configuration each)"
  )
```

The table above reports one held-out evaluation for each model after selecting its PC count using training-only cross-validation. PCA for the final fits is trained exclusively on the training partition and then applied to the untouched test partition. These results are useful for this dataset and split, but a single hold-out does not by itself establish deployment performance; repeated external evaluation would be needed for that claim.

------------------------------------------------------------------------

# Result Validation

Very high accuracy can be a symptom of data leakage or other methodological errors. The following checks assess selected leakage risks, but they do not replace repeated evaluation on an external cohort.

## Shuffled-Label Diagnostic

Shuffling labels removes their original pairing with expression profiles. The diagnostic uses the same 50-PC LDA configuration in both conditions. Uniform prediction has expected accuracy 20%; because the classes are imbalanced, that is not the expected accuracy of every shuffled-label fit. The training-majority baseline is shown too.

```{r}
#| label: validation-permutation

set.seed(123)
shuffled_labels <- sample(train_data$Class)

# LDA with 50 PCs on shuffled labels
train_md_shuf <- as_tibble(pca_result$x[, 1:50]) |>
  mutate(Class = shuffled_labels)
test_md_shuf <- as_tibble(pca_test_scores[, 1:50])

fit_shuf <- lda(Class ~ ., data = train_md_shuf)
preds_shuf <- predict(fit_shuf, newdata = test_md_shuf)$class
acc_shuf <- mean(preds_shuf == test_data$Class)
train_md_real <- train_md_shuf
train_md_real$Class <- train_data$Class
fit_real_50 <- lda(Class ~ ., data = train_md_real)
acc_real_50 <- mean(predict(fit_real_50, newdata = test_md_shuf)$class == test_data$Class)
majority_class <- names(which.max(table(train_data$Class)))
majority_accuracy <- mean(test_data$Class == majority_class)

tibble(
  Condition = c("Real labels (LDA, 50 PCs)", "Shuffled labels (LDA, 50 PCs)", "Uniform random prediction (5 classes)", "Training-majority prediction"),
  Accuracy = scales::percent(c(
    acc_real_50,
    acc_shuf,
    1/5,
    majority_accuracy
  ), accuracy = 0.1)
) |>
  datatable(
    options = list(dom = "t", ordering = FALSE),
    rownames = FALSE,
    caption = "Single shuffled-label diagnostic with matched LDA dimension"
  )
```

The table contrasts the real-label result with a single shuffled-label sanity check. It is diagnostic rather than a formal permutation test: a formal test would repeat many permutations and report an empirical p-value.

## Degradation with Fewer Components

The two-PC fit checks sensitivity to representation dimension. Reducing dimension can help or harm accuracy; no particular outcome proves that a model is legitimate.

```{r}
#| label: validation-2pc

train_md_2pc <- as_tibble(pca_result$x[, 1:2]) |>
  mutate(Class = train_data$Class)
test_md_2pc <- as_tibble(pca_test_scores[, 1:2])

fit_2pc <- lda(Class ~ ., data = train_md_2pc)
preds_2pc <- predict(fit_2pc, newdata = test_md_2pc)$class
cm_2pc <- confusionMatrix(preds_2pc, test_data$Class)

cm_2pc$table |>
  as.data.frame() |>
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "LDA with Only 2 PCs (Test Set)",
    subtitle = paste0("Test Accuracy: ", scales::percent(cm_2pc$overall["Accuracy"], accuracy = 0.1)),
    x = "True Class", y = "Predicted Class"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The two-PC matrix provides a deliberately information-limited baseline. Any difference from the selected-PC result illustrates sensitivity to representation dimension; it is not, by itself, a proof of absence of leakage.

## Train/Test Integrity

We verify that no sample appears in both sets and that no test sample is an exact duplicate of a training sample.

```{r}
#| label: validation-integrity

# Check sample ID overlap
train_ids <- train_data$sample_id
test_ids  <- test_data$sample_id
id_overlap <- length(intersect(train_ids, test_ids))

# Check for exact duplicate rows across the full analysis feature space.
check_genes <- nonzero_var_genes
train_mat <- train_data |> dplyr::select(all_of(check_genes)) |> as.matrix()
test_mat  <- test_data  |> dplyr::select(all_of(check_genes)) |> as.matrix()

n_duplicates <- 0
for (i in seq_len(nrow(test_mat))) {
  dists <- colSums((t(train_mat) - test_mat[i, ])^2)
  if (min(dists) == 0) n_duplicates <- n_duplicates + 1
}
stopifnot(id_overlap == 0, n_duplicates == 0)

tibble(
  Check = c(
    "Train samples",
    "Test samples",
    "Sample ID overlap",
    "Exact duplicate rows (test in train)"
  ),
  Result = c(
    as.character(length(train_ids)),
    as.character(length(test_ids)),
    as.character(id_overlap),
    as.character(n_duplicates)
  )
) |>
  datatable(
    options = list(dom = "t", ordering = FALSE),
    rownames = FALSE,
    caption = "Train/test set integrity checks"
  )
```

Assertions check zero ID overlap and zero exact expression-row overlap between training and test. These checks address specific leakage routes; they cannot establish the absence of every source of bias or leakage.

## Why Are These Results So High?

High scores can be plausible for this benchmark because the five labels arise from distinct tissues, but that explanation is a hypothesis about this dataset rather than a clinical-performance guarantee. The reported checks reduce common leakage risks; independent cohorts and repeated splits would be required to assess robustness.

------------------------------------------------------------------------

# Conclusion

This study evaluates probabilistic classifiers combined with PCA for five cancer types (BRCA, KIRC, COAD, LUAD and PRAD). The reported values in the generated tables use a train/test split, train-only PCA, and fold-isolated cross-validation for PC selection. They are evidence about this dataset and split, not proof of clinical generalisation.

## Key Findings

1.  **Separate fitting from evaluation.** The split has `r nrow(train_data)` training and `r nrow(test_data)` test samples. Each CV fold fits its own scaling and PCA; final fits use only the training partition.

2.  **PCA is the dimension-reduction method used here.** It makes the selected classical models estimable; regularization and other representations are alternatives. Retained variance does not guarantee retained predictive signal.

3.  **Model ranking should follow the rendered training-CV and held-out tables.** PCA coordinates are orthogonal, but that property does not guarantee the covariance assumptions of LDA or QDA.

4.  **Multinomial Logistic Regression** is a discriminative comparison that learns decision boundaries in the selected PCA space; its current performance is displayed in the generated result tables.

5.  **Naive Bayes is evaluated across the same PC grid.** Its independence assumption is not guaranteed by PCA, so any trend with component count should be read from the current cross-validation curve rather than a fixed narrative value.

6.  **Training-only EDA describes the input.** Correlations and ANOVA rankings are exploratory; they neither identify mechanisms nor select the classifier features.

7.  **The integrity checks are safeguards, not proof of generalisation.** The fold isolation, train/test disjointness and sanity checks reduce common leakage risks; an external cohort would still be required before making clinical-performance claims.

## Limitations and Future Work

-   The dataset is relatively **clean and well-separated**, which partly explains the strong results. Real clinical scenarios may involve noisier data, subtler cancer subtypes, or mixed tissue samples.
-   More advanced methods such as **regularized discriminant analysis**, **support vector machines**, or **ensemble methods** could be explored, although the current models already achieve very high performance on this dataset.
-   A **biological interpretation** of the top PCA loadings and ANOVA genes could provide insights into the molecular pathways distinguishing each cancer type, connecting the statistical results to domain knowledge.

------------------------------------------------------------------------

# Annex: Nonlinear Dimensionality Reduction (UMAP)

## Motivation

Throughout this study, we have relied on **PCA** for dimensionality reduction. PCA is a linear method: it finds the directions of maximum variance and projects the data onto them. These features and loadings are useful for this analysis, but linear projections can fail to capture complex, nonlinear relationships between samples when visualizing them in just two dimensions.

**UMAP** constructs a nonlinear low-dimensional representation from a neighborhood graph. Its axes, apparent cluster sizes and global distances depend on the data and hyperparameters; they should not be interpreted as direct biological distances.

**UMAP is used only for visualization in this study.** It can be used as an input transformation in other properly validated pipelines, but no such classifier is evaluated here.

We fit UMAP on the **first 50 PCA scores of the training set** (a fixed dimension-reduction choice) and then project the test set onto the same embedding.

```{r}
#| label: umap-compute
#| output: false

library(uwot)

# Fit UMAP on training PCA scores (first 50 PCs)
set.seed(42)
umap_train <- umap(
  pca_result$x[, 1:50],
  n_neighbors = 15,
  min_dist = 0.3,
  metric = "euclidean",
  ret_model = TRUE, n_threads = 1, n_sgd_threads = 1
)

# Project test set onto the same UMAP embedding
set.seed(42)
umap_test <- umap_transform(pca_test_scores[, 1:50], umap_train, n_threads = 1, n_sgd_threads = 1)

# Combine into tidy data frames
umap_train_df <- tibble(
  UMAP1 = umap_train$embedding[, 1],
  UMAP2 = umap_train$embedding[, 2],
  Class = train_data$Class,
  Set = "Train"
)

umap_test_df <- tibble(
  UMAP1 = umap_test[, 1],
  UMAP2 = umap_test[, 2],
  Class = test_data$Class,
  Set = "Test"
)

umap_all_df <- bind_rows(umap_train_df, umap_test_df)
```

## UMAP: Training Set

```{r}
#| label: umap-train-scatter

ggplot(umap_train_df, aes(x = UMAP1, y = UMAP2, color = Class)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "UMAP Projection of Training Set by Cancer Type",
    subtitle = "Fitted on the first 50 PCA scores of the training data",
    x = "UMAP 1", y = "UMAP 2"
  ) +
  theme_minimal()
```

This UMAP plot is an exploratory visualization of the training data. It can suggest local neighbourhood structure, but its axes and global distances should not be used as quantitative evidence of class separation or classifier performance.

## UMAP: Test Set Projection

```{r}
#| label: umap-test-overlay

ggplot() +
  geom_point(
    data = umap_train_df,
    aes(x = UMAP1, y = UMAP2, color = Class),
    alpha = 0.2, size = 1.5
  ) +
  geom_point(
    data = umap_test_df,
    aes(x = UMAP1, y = UMAP2, color = Class),
    alpha = 0.9, size = 2.5, shape = 17
  ) +
  labs(
    title = "UMAP: Test Samples Projected onto Training Embedding",
    subtitle = paste0("Triangles = test set (", nrow(test_data), " samples), circles = training set (faded)"),
    x = "UMAP 1", y = "UMAP 2"
  ) +
  theme_minimal()
```

The overlay shows how this split projects into the UMAP fitted on the training partition. It is a visual diagnostic only: proximity in a UMAP embedding is not an independent validation of the classifiers or a measure of generalisation.

## UMAP by Individual Class

```{r}
#| label: umap-faceted-class

umap_train_df |>
  mutate(highlight = Class) |>
  ggplot() +
  geom_point(
    data = umap_train_df |> dplyr::select(-Class) |> rename(UMAP1_ = UMAP1, UMAP2_ = UMAP2),
    aes(x = UMAP1_, y = UMAP2_),
    color = "grey85", size = 1, alpha = 0.5
  ) +
  geom_point(
    aes(x = UMAP1, y = UMAP2, color = Class),
    size = 1.5, alpha = 0.8
  ) +
  facet_wrap(~Class, ncol = 3) +
  labs(
    title = "UMAP: Each Cancer Type Highlighted Against All Others",
    subtitle = "Grey points show all other classes for spatial context",
    x = "UMAP 1", y = "UMAP 2"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The faceted view highlights each class against the full training embedding. Compare the displayed locations within this run; coordinate orientation, apparent spread and inter-class gaps are properties of this fitted visualization, not validated biological distances or estimates of molecular heterogeneity.

## UMAP with Density Contours

```{r}
#| label: umap-density

ggplot(umap_train_df, aes(x = UMAP1, y = UMAP2, color = Class)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_density_2d(linewidth = 0.6, alpha = 0.8) +
  labs(
    title = "UMAP with Density Contours per Cancer Type",
    subtitle = "Contour lines show the concentration regions of each class",
    x = "UMAP 1", y = "UMAP 2"
  ) +
  theme_minimal()
```

The contours summarize density in the two-dimensional visualization. They are descriptive; UMAP can distort global geometry, so this figure should not be treated as a statistical test of class separability.

## UMAP vs PCA: Side-by-Side Comparison

```{r}
#| label: umap-vs-pca

library(patchwork)

p_pca <- pca_scores |>
  ggplot(aes(x = PC1, y = PC2, color = Class)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(title = "PCA (PC1 vs PC2)", x = "PC1", y = "PC2") +
  theme_minimal() +
  theme(legend.position = "none")

p_umap <- umap_train_df |>
  ggplot(aes(x = UMAP1, y = UMAP2, color = Class)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(title = "UMAP", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

p_pca + p_umap +
  plot_annotation(
    title = "Linear (PCA) vs Nonlinear (UMAP) Dimensionality Reduction",
    subtitle = "Both fitted on training data only"
  )
```

The side-by-side comparison highlights the **complementary strengths** of each method:

-   **PCA** displays directions of maximum standardized training variance. A two-dimensional projection may hide structure present in later components.
-   **UMAP** emphasizes local neighborhoods. Apparent separation in this visualization is not a classifier test or proof of independent biological clusters.

These visualizations complement the quantitative results. They do not independently establish generalization, tissue mechanisms, or an absence of leakage.

```{r}
#| label: export-results
#| include: false
audit_output <- Sys.getenv("RNA_SEQ_AUDIT_OUTPUT", unset = "")
if (nzchar(audit_output)) saveRDS(list(cv = results_all, test = test_summary, cutoffs = pc_cutoffs, split = c(train = nrow(train_data), test = nrow(test_data)), diagnostic = c(real = acc_real_50, shuffled = acc_shuf, majority = majority_accuracy), integrity = c(overlap = id_overlap, duplicates = n_duplicates)), audit_output)
```
